### 2020-11-17 Draft/Draft_MEM.Rnw
Initial commit

\documentclass{article}
\usepackage[left=2.5cm,right=2.5cm, top = 2.5cm, bottom = 3cm]{geometry}

\begin{document}

\title{A permutation-based comparison of ILI intensity thresholds from the moving epidemic and WHO methods}
\author{Johannes Bracher}

\maketitle

\abstract{The moving epidemic method (MEM) and the WHO method are widely used approaches to determine intensity levels for seasonal influenza and influenca-like illness (ILI). They are conceptually similar, but differ in two points. Firstly, the MEM involves a log-transformation of incidence data, while the WHO method operates on the original scale of the data. Secondly, the MEM method usually uses more than one observation from each past season, with the exact number depending on the number of available historical seasons. The WHO method, on the other side, uses only the single highest value from each past season. We perform a simulation study to assess the impact of these choices on the resulting intensity thresholds and empirical exceedance proportions. It is based on a permutation approach using historical ILI data from France, Spain, Switzerland and the United States.}

\section{Introduction}

\section{Definition of the moving epidemic and WHO methods}

\section{A brief review of recent applications}

Cite Benedetti who use it as a "gold standard"

\section{Simulaton study}

\subsection{Permutation approach based on historical ILI data}

\subsection{Data}

\subsection{Used software}

MEM version xxx, R software

\subsection{Results}

\section{Discussion}

\section*{Outline of idea}

\begin{itemize}
\item Description of importance: officially embraced by ECDC
\item Describe other approaches, in particular WHO
\item Statistically correct description of what is being done, contract to WHO approach (which only uses peak -- good -- but uses normal rather than log-normal -- not so good. There is a PLOS paper where there is already a kind of combination of the two (WHO with log transformation))
\item Overview of how it is being used in the literature:
\begin{itemize}
\item How much training data was available in each study?
\end{itemize}
\item Simulation study based on permutation of true seasons:
\begin{itemize}
\item Countries: France, Switzerland, some third (European) country? Spain?
\item Important: standardize with time-varying population
\end{itemize}
\end{itemize}

\title{A statistical perspective on the moving epidemic method }
\author{Johannes Bracher}
\maketitle

\bibliography{bibliography_mem}
\end{document}

### 2020-11-17 Simulations/data_preparation.R

@@ -1,0 +1,115 @@
+# Preparing the different data sets for use of mem
+
+setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+
+# France:
+
+# get data:
+dat_fr0 <- read.csv("../Data/raw/ili_fr.csv")
+
+# bring in format for mem:
+dat_fr <- data.frame(value = dat_fr0$inc)
+dat_fr$week <- dat_fr0$week %% 100
+dat_fr$year <- (dat_fr0$week - dat_fr$week)/100
+dat_fr$season[dat_fr$week <= 40] <- paste0(dat_fr$year[dat_fr$week <= 40] - 1, "/", dat_fr$year[dat_fr$week <= 40])
+dat_fr$season[dat_fr$week > 40] <- paste0(dat_fr$year[dat_fr$week > 40], "/", dat_fr$year[dat_fr$week > 40] + 1)
+dat_fr <- dat_fr[order(dat_fr$year, dat_fr$week), ]
+for(i in 1:nrow(dat_fr)) dat_fr$week_in_season[i] <- sum(dat_fr$season[1:i] == dat_fr$season[i])
+
+plot(dat_fr$year + dat_fr$week/52, dat_fr$value, type = "l")
+
+# add population:
+pop_fr <- read.csv("../Data/raw/population_fr.csv")
+pop_fr$season <- paste0(pop_fr$year - 1, "/", pop_fr$year)
+pop_fr$year <- NULL
+
+dat_fr <- merge(dat_fr, pop_fr, by = "season")
+dat_fr$inc <- dat_fr$value*100000/dat_fr$population
+
+plot(dat_fr$year + dat_fr$week/52, dat_fr$inc, type = "l")
+
+# remove half-seasons at the ends and pandemic season:
+dat_fr <- subset(dat_fr, !(season %in% c("1984/1985", "2009/2010", "2020/2021")))
+dat_fr <- subset(dat_fr, week_in_season <= 30)
+dat_fr <- dat_fr[, c("week_in_season", "season", "inc")]
+dat_fr_wide <- reshape(dat_fr, direction = "wide", idvar = "week_in_season", timevar = "season", v.names = "inc")
+
+dat_for_mem_fr <- (dat_fr_wide[, grepl(pattern = "/", colnames(dat_fr_wide), fixed = TRUE)])
+plot(dat_for_mem_fr[, 1])
[…]

### 2020-11-17 Simulations/functions_simulations_mem.R

@@ -1,0 +1,88 @@
+sim_summary <- function(results_0.4, results_0.9, results_0.975, range_i.seasons,
+                        interval = c(0.025, 0.975)){
+  if(length(interval) != 2) stop("interval needs to be of length 2.")
+  interval <- sort(interval)
+
+  ret <- list()
+
+  ret$averages <- matrix(ncol = 3, nrow = length(range_i.seasons))
+
+  ret$averages <- matrix(ncol = 3, nrow = length(range_i.seasons))
+  colnames(ret$averages) <- c("medium_0.4", "high_0.9", "very_high_0.975")
+  rownames(ret$averages) <- paste0("i.seasons_", range_i.seasons)
+
+  ret$averages[, "medium_0.4"] <- colMeans(results_0.4, na.rm = TRUE)
+  ret$averages[, "high_0.9"] <- colMeans(results_0.9, na.rm = TRUE)
+  ret$averages[, "very_high_0.975"] <- colMeans(results_0.975, na.rm = TRUE)
+
+  for(p in interval){
+    ret[[paste0("q_", p)]] <- NA*ret$averages
+
+    ret[[paste0("q_", p)]][, "medium_0.4"] <- apply(results_0.4, MARGIN = 2, FUN = quantile, na.rm = TRUE, p = p)
+    ret[[paste0("q_", p)]][, "high_0.9"] <- apply(results_0.9, MARGIN = 2, FUN = quantile, na.rm = TRUE, p = p)
+    ret[[paste0("q_", p)]][, "very_high_0.975"] <- apply(results_0.975, MARGIN = 2, FUN = quantile, na.rm = TRUE, p = p)
+  }
+  ret$range_i.seasons <- range_i.seasons
+  return(ret)
+}
+
+# modify the alpha value of a given color (to generate transparent versions for prediction bands)
+modify_alpha <- function(col, alpha){
+  x <- col2rgb(col)/255
+  rgb(x[1], x[2], x[3], alpha = alpha)
+}
+
+plot_sim_summary <- function(sim_summary, ylim = NULL, show_bands = TRUE, ...){
+  ind_lower <- names(sim_summary)[grepl("q_", names(sim_summary))][1]
+  ind_upper <- names(sim_summary)[grepl("q_", names(sim_summary))][2]
+
+  if(is.null(ylim)){
+    yl <- c(0, 1.5*max(sim_summary$averages[, "very_high_0.975"], na.rm = TRUE))
+  }else{
+    yl <- ylim
+  }
+
[…]

### 2020-11-17 Simulations/plausibility_mem.R

@@ -1,0 +1,49 @@
+# checking that the mam function does what I think it does:
+
+setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+library(mem)
+library(HIDDA.forecasting)
+
+# get Swiss data:
+data("CHILI")
+# bring in format for mem:
+dat_ch <- data.frame(date = index(CHILI), value = as.vector(CHILI))
+dat_ch$week <- MMWRweek::MMWRweek(dat_ch$date)$MMWRweek
+dat_ch$year <- MMWRweek::MMWRweek(dat_ch$date)$MMWRyear
+dat_ch$season[dat_ch$week <= 40] <- paste0(dat_ch$year[dat_ch$week <= 40] - 1, "/", dat_ch$year[dat_ch$week <= 40])
+dat_ch$season[dat_ch$week > 40] <- paste0(dat_ch$year[dat_ch$week > 40], "/", dat_ch$year[dat_ch$week > 40] + 1)
+for(i in 1:nrow(dat_ch)) dat_ch$week_in_season[i] <- sum(dat_ch$season[1:i] == dat_ch$season[i])
+
+# add population:
+pop_ch <- read.csv("../Data/population_ch.csv")
+pop_ch$season <- paste0(pop_ch$year - 1, "/", pop_ch$year)
+pop_ch$year <- NULL
+dat_ch <- merge(dat_ch, pop_ch, by = "season")
+
+dat_ch$inc <- dat_ch$value*100000/dat_ch$population
+
+dat_ch <- subset(dat_ch, !(season %in% c("1999/2000", "2009/2010", "2016/2017")))
+dat_ch <- subset(dat_ch, week_in_season <= 30)
+dat_ch <- dat_ch[, c("week_in_season", "season", "inc")]
+dat_ch_wide <- reshape(dat_ch, direction = "wide", idvar = "week_in_season", timevar = "season", v.names = "inc")
+
+dat_for_mem <- (dat_ch_wide[, grepl(pattern = "/", colnames(dat_ch_wide), fixed = TRUE)])
+
+
+# run mem models with and without log
+memmod <- memmodel(dat_for_mem, i.seasons = 10,
+                   i.level.intensity = c(0.4, 0.9, 0.975),
+                   i.n.max = 1)
+memmod$intensity.thresholds
+
+memmod_nolog <- memmodel(dat_for_mem, i.seasons = 10,
[…]

### 2020-11-17 Simulations/plot_timeseries.R

@@ -1,0 +1,46 @@
+setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+
+dat_ch <- read.csv("../Data/for_mem/ili_mem_ch.csv")
+dat_es <- read.csv("../Data/for_mem/ili_mem_es.csv")
+dat_fr <- read.csv("../Data/for_mem/ili_mem_fr.csv")
+dat_us <- read.csv("../Data/for_mem/ili_mem_us.csv")
+
+dim(dat_us)
+
+plot(dat_us[, 5])
+
+matr <- dat_ch
+xlim <- c(1990, 2020)
+plot_mem_ts <- function(matr, xlim, ylab = "inc / 100,000", ...){
+  years <- as.numeric(substr(colnames(matr), start = 13, stop = 16)) - 1
+  plot(NULL, xlim = xlim, ylim = c(0, 1.1*max(matr)), xlab = "", ylab = ylab, ...)
+  abline(v = seq(from = floor(xlim[1]), to = ceiling(xlim[2])), col = "grey")
+
+  weeks_per_seas <- nrow(matr)
+  for(i in 1:ncol(matr)){
+    x <- years[i] + (1:weeks_per_seas)/weeks_per_seas
+    lines(x, matr[, i])
+  }
+}
+
+bplot_ranks <- function(matr, ylab = ""){
+  plot(NULL, xlim = c(0.5, 6.5), ylim = c(0, 1.1*max(matr)), xlab = "", ylab = ylab)
+  mtext("rank in season", 1, line = 1.8, cex = 0.6)
+  get_nth <- function(vect, n) sort(vect, decreasing = TRUE)[n]
+  for(i in 1:6) boxplot(apply(matr, 2, get_nth, n = i), at = i, add = TRUE, axes = FALSE)
+}
+
+par(las = 1, mar = c(3, 4, 0.5, 1))
+layout(matrix(1:8, ncol = 2, byrow = TRUE), widths = c(8, 2))
+plot_mem_ts(dat_fr, xlim = c(1986, 2019)); legend("topleft", legend = "France", bty = "n", cex = 1.2)
+bplot_ranks(dat_fr)
+
+plot_mem_ts(dat_es, xlim = c(1986, 2019)); legend("topleft", legend = "Spain", bty = "n", cex = 1.2)
+bplot_ranks(dat_es)
+
[…]

### 2020-11-17 Simulations/simulations_mem_ch.R

@@ -1,0 +1,271 @@
+############################################################
+# Running permutation/simulation study for Switzerland
+
+library(mem)
+source("functions_simulations_mem.R")
+
+dat_for_mem <- read.csv("../Data/for_mem/ili_mem_ch.csv")
+
+plot(dat_for_mem$`inc.2000/2001`)
+
+# simulation settings:
+range_i.seasons <- 5:15 # range to be explored for i.seasons
+n_sim <- 500
+
+#####################################################
+### Run simulation using default settings of mem
+#####################################################
+
+# initialize matrices to store results (thresholds and exceedance proportions)
+thresholds_ch_0.4 <- matrix(NA, nrow = n_sim, ncol = length(range_i.seasons))
+colnames(thresholds_ch_0.4) <- paste0("i.seasons_", range_i.seasons)
+thresholds_ch_0.9 <- thresholds_ch_0.975 <-
+  exceedance_ch_0.4 <- exceedance_ch_0.9 <- exceedance_ch_0.975 <-
+  thresholds_ch_0.4
+
+# run simulation:
+for(k in 1:n_sim){
+  set.seed(k)
+
+  # run through different numbers of past seasons used:
+  for(a in seq_along(range_i.seasons)){
+
+    # split into training and test:
+    inds_train <- sample(1:ncol(dat_for_mem), size = range_i.seasons[a], replace = TRUE)
+    dat_train <- dat_for_mem[, inds_train]
+    dat_test <- dat_for_mem # [, -inds_train]
+    # compute peaks of test seasons:
+    peaks_test <- apply(dat_test, 2, max)
+
+    # run memmodel:
+    memmodel_temp <- memmodel(dat_train, i.seasons = range_i.seasons[a],
+                              i.level.intensity = c(0.4, 0.9, 0.975))
+
+    # store thresholds:
+    thresholds_ch_0.4[k, a] <- memmodel_temp$intensity.thresholds[1]
+    thresholds_ch_0.9[k, a] <- memmodel_temp$intensity.thresholds[2]
+    thresholds_ch_0.975[k, a] <- memmodel_temp$intensity.thresholds[3]
+
+    # compute and store exceedance proportions:
+    exceedance_ch_0.4[k, a] <- mean(peaks_test > thresholds_ch_0.4[k, a])
[…]

### 2020-11-17 Simulations/simulations_mem_es.R

@@ -1,0 +1,275 @@
+############################################################
+# Running permutation/simulation study for Spain
+setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+library(mem)
+
+source("functions_simulations_mem.R")
+
+dat_for_mem <- read.csv("../Data/ili_mem_es.csv")
+
+head(dat_for_mem)
+
+plot(dat_for_mem[, 4])
+
+# simulation settings:
+range_i.seasons <- 5:15 # range to be explored for i.seasons
+n_sim <- 500
+
+#####################################################
+### Run simulation using default settings of mem
+#####################################################
+
+# initialize matrices to store results (thresholds and exceedance proportions)
+thresholds_es_0.4 <- matrix(NA, nrow = n_sim, ncol = length(range_i.seasons))
+colnames(thresholds_es_0.4) <- paste0("i.seasons_", range_i.seasons)
+thresholds_es_0.9 <- thresholds_es_0.975 <-
+  exceedance_es_0.4 <- exceedance_es_0.9 <- exceedance_es_0.975 <-
+  thresholds_es_0.4
+
+# run simulation:
+for(k in 1:n_sim){
+  set.seed(k)
+
+  # run through different numbers of past seasons used:
+  for(a in seq_along(range_i.seasons)){
+
+    # split into training and test:
+    inds_train <- sample(1:ncol(dat_for_mem), size = range_i.seasons[a], replace = TRUE)
+    dat_train <- dat_for_mem[, inds_train]
+    dat_test <- dat_for_mem # [, -inds_train]
+    # compute peaks of test seasons:
+    peaks_test <- apply(dat_test, 2, max)
+
+    # run memmodel:
+    memmodel_temp <- memmodel(dat_train, i.seasons = range_i.seasons[a],
+                              i.level.intensity = c(0.4, 0.9, 0.975))
+
+    # store thresholds:
+    thresholds_es_0.4[k, a] <- memmodel_temp$intensity.thresholds[1]
+    thresholds_es_0.9[k, a] <- memmodel_temp$intensity.thresholds[2]
+    thresholds_es_0.975[k, a] <- memmodel_temp$intensity.thresholds[3]
+
+    # compute and store exceedance proportions:
[…]

### 2020-11-17 Simulations/simulations_mem_fr.R

@@ -1,0 +1,268 @@
+############################################################
+# Running permutation/simulation study for Switzerland
+
+setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+
+library(mem)
+source("functions_simulations_mem.R")
+
+dat_for_mem <- read.csv("../Data/for_mem/ili_mem_fr.csv")
+
+
+# simulation settings:
+range_i.seasons <- 5:15 # range to be explored for i.seasons
+n_sim <- 500
+
+
+#####################################################
+### Run simulation using default settings of mem
+#####################################################
+
+# initialize matrices to store results (thresholds and exceedance proportions)
+thresholds_fr_0.4 <- matrix(NA, nrow = n_sim, ncol = length(range_i.seasons))
+colnames(thresholds_fr_0.4) <- paste0("i.seasons_", range_i.seasons)
+thresholds_fr_0.9 <- thresholds_fr_0.975 <-
+  exceedance_fr_0.4 <- exceedance_fr_0.9 <- exceedance_fr_0.975 <-
+  thresholds_fr_0.4
+
+# run simulation:
+for(k in 1:n_sim){
+  set.seed(k)
+
+  # run through different numbers of past seasons used:
+  for(a in seq_along(range_i.seasons)){
+
+    # split into training and test:
+    inds_train <- sample(1:ncol(dat_for_mem), size = range_i.seasons[a], replace = TRUE)
+    dat_train <- dat_for_mem[, inds_train]
+    dat_test <- dat_for_mem#[, -inds_train]
+    # compute peaks of test seasons:
+    peaks_test <- apply(dat_test, 2, max)
+
+    # run memmodel:
+    memmodel_temp <- memmodel(dat_train, i.seasons = range_i.seasons[a],
+                            i.level.intensity = c(0.4, 0.9, 0.975))
+
+    # store thresholds:
+    thresholds_fr_0.4[k, a] <- memmodel_temp$intensity.thresholds[1]
+    thresholds_fr_0.9[k, a] <- memmodel_temp$intensity.thresholds[2]
+    thresholds_fr_0.975[k, a] <- memmodel_temp$intensity.thresholds[3]
+
+    # compute and store exceedance proportions:
[…]

### 2020-11-17 Simulations/simulations_mem_us.R

@@ -1,0 +1,273 @@
+############################################################
+# Running permutation/simulation study for Spain
+setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+library(mem)
+library(HIDDA.forecasting)
+
+source("functions_simulations_mem.R")
+
+dat_for_mem <- read.csv("../Data/for_mem/ili_mem_us.csv")
+
+plot(dat_for_mem[, 2])
+
+# simulation settings:
+range_i.seasons <- 5:15 # range to be explored for i.seasons
+n_sim <- 500
+
+#####################################################
+### Run simulation using default settings of mem
+#####################################################
+
+# initialize matrices to store results (thresholds and exceedance proportions)
+thresholds_us_0.4 <- matrix(NA, nrow = n_sim, ncol = length(range_i.seasons))
+colnames(thresholds_us_0.4) <- paste0("i.seasons_", range_i.seasons)
+thresholds_us_0.9 <- thresholds_us_0.975 <-
+  exceedance_us_0.4 <- exceedance_us_0.9 <- exceedance_us_0.975 <-
+  thresholds_us_0.4
+
+# run simulation:
+for(k in 1:n_sim){
+  set.seed(k)
+
+  # run through different numbers of past seasons used:
+  for(a in seq_along(range_i.seasons)){
+
+    # split into training and test:
+    inds_train <- sample(1:ncol(dat_for_mem), size = range_i.seasons[a], replace = TRUE)
+    dat_train <- dat_for_mem[, inds_train]
+    dat_test <- dat_for_mem # [, -inds_train]
+    # compute peaks of test seasons:
+    peaks_test <- apply(dat_test, 2, max)
+
+    # run memmodel:
+    memmodel_temp <- memmodel(dat_train, i.seasons = range_i.seasons[a],
+                              i.level.intensity = c(0.4, 0.9, 0.975))
+
+    # store thresholds:
+    thresholds_us_0.4[k, a] <- memmodel_temp$intensity.thresholds[1]
+    thresholds_us_0.9[k, a] <- memmodel_temp$intensity.thresholds[2]
+    thresholds_us_0.975[k, a] <- memmodel_temp$intensity.thresholds[3]
+
+    # compute and store exceedance proportions:
[…]

### 2020-11-17 Simulations/summarize_simulations.R

@@ -1,0 +1,197 @@
+# summarize simulation results
+
+setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+source("functions_simulations_mem.R")
+
+# the range of number of seasons:
+range_i.seasons <- 5:15
+
+for(country in c("ch", "es", "fr", "us")){
+  load(paste0("Results/results_", country, ".rda"))
+  load(paste0("Results/results1_", country, ".rda"))
+  load(paste0("Results/results_nolog_", country, ".rda"))
+  load(paste0("Results/results1_nolog_", country, ".rda"))
+}
+
+# which uncertainty bands to compute?
+interv <- c(0.05, 0.95)
+
+# compute summaries using custom function
+
+# Switzerland:
+summary_thresholds_ch <- sim_summary(thresholds_ch_0.4, thresholds_ch_0.9, thresholds_ch_0.975,
+                                     range_i.seasons, interval = interv)
+summary_exceedance_ch <- sim_summary(exceedance_ch_0.4, exceedance_ch_0.9, exceedance_ch_0.975,
+                                     range_i.seasons, interval = interv)
+summary_thresholds1_ch <- sim_summary(thresholds1_ch_0.4, thresholds1_ch_0.9, thresholds1_ch_0.975,
+                                      range_i.seasons, interval = interv)
+summary_exceedance1_ch <- sim_summary(exceedance1_ch_0.4, exceedance1_ch_0.9, exceedance1_ch_0.975,
+                                      range_i.seasons, interval = interv)
+summary_thresholds_nolog_ch <- sim_summary(thresholds_nolog_ch_0.4, thresholds_nolog_ch_0.9, thresholds_ch_0.975,
+                                           range_i.seasons, interval = interv)
+summary_exceedance_nolog_ch <- sim_summary(exceedance_nolog_ch_0.4, exceedance_nolog_ch_0.9, exceedance_ch_0.975,
+                                           range_i.seasons, interval = interv)
+summary_thresholds1_nolog_ch <- sim_summary(thresholds1_nolog_ch_0.4, thresholds1_nolog_ch_0.9, thresholds1_nolog_ch_0.975,
+                                            range_i.seasons, interval = interv)
+summary_exceedance1_nolog_ch <- sim_summary(exceedance1_nolog_ch_0.4, exceedance1_nolog_ch_0.9, exceedance1_nolog_ch_0.975,
[…]

### 2020-12-27 Draft/Draft_MEM.tex
Changing Rnw to tex.
Update.

--- 2020-11-17 Draft/Draft_MEM.Rnw
@@
 \usepackage[left=2.5cm,right=2.5cm, top = 2.5cm, bottom = 3cm]{geometry}
+\usepackage{graphicx}
+\usepackage{natbib}
+\usepackage{hyperref}
 
 \begin{document}
+% \SweaveOpts{concordance=TRUE}
 
 \title{A permutation-based comparison of ILI intensity thresholds from the moving epidemic and WHO methods}
-\author{Johannes Bracher}
+\author{Johannes Bracher, Jonas Littek}
 
@@
 Firstly, the MEM involves a log-transformation of incidence data, while the WHO method operates on the original scale of the data.
-Secondly, the MEM method usually uses more than one observation from each past season, with the exact number depending on the number of available historical seasons.
+Secondly, the MEM method usually uses more than one observation from each past season, with a fixed total number of observations to include.
 The WHO method, on the other side, uses only the single highest value from each past season.
 We perform a simulation study to assess the impact of these choices on the resulting intensity thresholds and empirical exceedance proportions.
-It is based on a permutation approach using historical ILI data from France, Spain, Switzerland and the United States.}
+It is based on a permutation approach using historical ILI data from France, Spain, Switzerland and the United States.
+When fixing the total number of observations to include, intensity thresholds tend to increase as more data accumulates.
+When data is only available for few seasons, there is a rather large chance of classifying a new season as ``high'' or ``very high'' intenstity.
+We therefore advocate always using one observation per season.}
 
@@
 
+Following the 2009 influenza H1N1 pandemic, the need for a rapid assessment tool for influenza intensity was recognized. In a report, the Review Committee on the Functioning of the
+International Health Regulations (2005) and on Pandemic Influenza (H1N1) 2009 recommended that member states apply and evaluate their methods for severity assessment every year, thus improving their pandemic preparedness \citep[118]{WHO2011}.
+In its subsequent Pandemic Infuenza Severity Assessment (PISA) guide (\citeyear{WHO2017}), two specific statistical methods are recommended to determine influenza intensity thresholds.
+Firstly, the so-called \textit{WHO method}, already described in a previous document \citep{WHO2014} and in more detail in the PISA guide; and secondly, the \textit{moving epidemic method} (MEM) \citep{Vega2013, Vega2015} which is also recommended by the European Centre for Disease Prevention and Control (e.g.\ \citealt{ECDC2017}).
+
+In particular the MEM approach has in the meantime been adopted by numerous public health agencies countries in recent years and seems to deliver practically useful thresholds.
+The statistical properties of the MEM and WHO methods, however, have not yet been studied in detail.
+While it seems challenging to obtain strong mathematical results on their behaviour, we argue that simulation experiments can improve our understanding and intuition of the consequences of different implementation choices.
+We here provide such a simulation study, which in order to ensure a close link to real influenza patterns, is based on permutations of historical surveillance data.
+
+The article is structured as follows.
+Section \ref{sec:definitions} provides concise definitions of the MEM and WHO methods, highlighting two important differences between these conceptually similar approaches.
+Section \ref{sec:recent_applications} consists of an overview of published applications of the MEM and WHO methods.
+In Section \ref{sec:simulation} we describe our simulation approach and the results based on historical data from France, Spain, Switzerland and the United States.
+Section \ref{sec:discussion} concludes with a discussion.
+
 \section{Definition of the moving epidemic and WHO methods}
+\label{sec:definitions}
 
 \section{A brief review of recent applications}
+\label{sec:recent_applications}
 
-Cite Benedetti who use it as a "gold standard"
+Cite Benedetti who use it as a "gold standard". Mention Green GB who compare to percentile method. Vos mention fixed criterion model -- what is that?
+
+\textbf{add table with Jonas' literature review here}
 
 \section{Simulaton study}
+\label{sec:simulation}
 
@@
 \subsection{Data}
+
+\includegraphics[width=0.9\textwidth]{figure/plot_data.pdf}
 
@@
 
+\includegraphics[page=1, width=0.9\textwidth]{figure/plot_results.pdf}
+
+\includegraphics[page=2, width=0.9\textwidth]{figure/plot_results.pdf}
+
 \section{Discussion}
+\label{sec:discussion}
 
@@
 
+\bibliographystyle{apalike}
 \bibliography{bibliography_mem}

### 2020-12-28 Draft/Draft_MEM.tex
Update.

@@
 \usepackage{hyperref}
+\usepackage{todonotes}
+\usepackage{booktabs}
 
@@
 
-In particular the MEM approach has in the meantime been adopted by numerous public health agencies countries in recent years and seems to deliver practically useful thresholds.
+In particular the MEM approach has in the meantime been adopted by numerous national public health agencies and seems to deliver practically useful thresholds.
 The statistical properties of the MEM and WHO methods, however, have not yet been studied in detail.
@@
 
-\section{A brief review of recent applications}
+In the following we focus on the construction of thresholds  for peak intensity, framing the MEM and WHO methods as different special cases of a more general approach.
+Details on the construction of baseline thresholds from pre-season observations are omitted.
+We assume that thresholds are to be constructed from data from $n$ past seasons and subsequently be applied to the current ($n + 1$-th) season.
+\cite{Vega2015} recommend to use $5 \leq n \leq 10$ seasons to ensure a sufficiently large, but recent data basis.
+Thresholds for for a given intensity measure $X$ are then obtained via the following steps:
+
+\begin{enumerate}
+\item For each historical season order all observations, denoting the $i$-th largest observation from season $j$ by $X^{(i)}_j$.
+\item Select the $K$ largest observations from each of the $n$ past seasons to construct a reference set $\mathcal{X} = \{X_j^{(i)}; i = 1, \dots, n; j = 1, \dots, K\}$.
+\item Apply an (inversible) transformation $Y_j^{(i)} = f(X_j^{(i)})$ to all members of the reference set to obtain a reference set $\mathcal{Y}$ of transformed past observations.
+\item Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute its estimated mean $\hat\mu$ and standard deviation $\hat \sigma$.
+\item Define intensity thresholds on the transformed scale as quantiles of this normal distribution.
+A common choice for both methods \citep{WHO2017}, also implemented as the default in the \texttt{R} package \texttt{mem} \citep{Lozano2020} is
+\begin{itemize}
+\item the 40th percentile $q_{Y, 0.4} = \hat\mu - 0.25 \hat\sigma$ as the threshold between low and medium intensity;
+\item the 90th percentile $q_{Y, 0.9} = \hat\mu + 1.28 \hat\sigma$ as the threshold between medium and high intensity;
+\item the 97.5th percentile $q_{Y, 0.975} = \hat\mu + 1.96\hat\sigma$ as the threshold between high and very high intensity.
+\end{itemize}
+\item Obtain thresholds on the original scale by applying the inverse transformation $q_{0.4} = f^{-1}(q_{Y, 0.4})$, $q_{0.9} = f^{-1}(q_{Y, 0.9})$, $q_{0.975} = f^{-1}(q_{Y, 0.975})$.
+\end{enumerate}
+
+\noindent The MEM and WHO methods are special cases of this procedure with the following specifications:
+\begin{itemize}
+\item In the MEM method, $f$ is chosen as the natural logarithm, while the number of observations per season is set to $K = 30/n$.
+(The exact specification used in the \texttt{R} package \texttt{mem} is $K = \max(1, \lfloor 30/n \rceil)$, where $\lfloor \cdot \rceil$ denotes rounding to the nearest integer.) The total nuber of historical observations is thus kept (approximately) fixed.
+\item In the WHO method, the standard procedure is to set $f$ to the identity function (i.e.\ no transformation is applied).
+If peaks vary considerably, a log-tranformation is recommended \citep{WHO2017}.
+Irrespective of the number $n$ of historical seasons, $K = 1$ observation per season is used.
+\end{itemize}
+
+\noindent The rationale behind this approach is that ``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold \citep{WHO2017}.
+In most previous descriptions (e.g., \ \citealt{WHO2014, Vega2015}), the thresholds have been introduced as upper ends of one-sided confidence intervals associated with the the arithmetic (WHO method) or geometric mean (MEM) of the historical reference observations in $\mathcal{X}$.
+This, however, is slightly imprecise terminology as in the computations $\sigma$ rather than $\sigma/\sqrt{n}$ is used.
+If indeed confidence intervals were used, these would get narrower the more historical seasons were available, and the thresholds for all levels would eventually converge to the arithmetic / geometric mean.
+This is obviously not the intended behaviour.
+
+% see function iconfianza.aritmetica
+
+\section{An overview of recent applications}
 \label{sec:recent_applications}
 
-Cite Benedetti who use it as a "gold standard". Mention Green GB who compare to percentile method. Vos mention fixed criterion model -- what is that?
+To get a better understanding of the different settings in which the MEM and WHO methods are applied in practice we performed a literature search among all published articles citing the papers \cite{Vega2015}, \cite{WHO2014} and \cite{WHO2017}.
+The results are summarized in Table \ref{tab:literature}.
+As can be seen from the numerous entries from the years 2019 and 2020, the MEM has quickly become the standard in the determination of intensity thresholds for influenza and other respiratory diseases.
+In most analyses, the suggested threshold levels at the 40th, 90th and 97.5th percentile are used.
+The contributions come from numerous countries and in many cases have been co-authors by representatives of national or regional public health agencies.
+Variability with respect to the number $n$ of historical seasons included is considerable, with a range from 3 to 16 years.
+Consequently, the number of observations included per season ranges from two to ten.
 
-\textbf{add table with Jonas' literature review here}
+\begin{table}
+\caption{Applications of the moving epidemic method and WHO method to determie severity thresholds.
+Does not contain works where only baseline thresholds are computed.
+The number of seasons included to compute thresholds is denoted by $n$, the number of observations used per season by $K$.
+Abbreviations: SARI = severe acute respiratory infection; ILI = influenza-like illness; RSV = respiratory syncytial virus.}
+\label{tab:literature}
+\center
+\footnotesize
+\begin{tabular}{l l l l l l l}
+\multicolumn{7}{c}{(a) Moving epidemic method}\\ \\
+
+\toprule
+region & disease & years covered & $n$ & $K$ & percentiles & authors\\
+\midrule
+Australia & ILI/influenza & 2012--2017 & 5 & 6 & 40, 90, 99 & \cite{Vette2018}\\
+Australia, Chile, & ILI/SARI & 2013--2019 & 6 & 5 & 40, 90, 97.5 & \cite{Sullivan2019}\\
+New Zealand,\\
+South Africa\\
+Catalonia & ILI & 2010--2016 & 5 & 6 & ? & \cite{Basile2018}\\
+Catalonia & ILI & 2005--2018 & 12 & 3 & ? & \cite{Basile2019}\\
+Catalonia & ILI/influenza & 2010--2017 & 7 & 4 & ? & \cite{Torner2019}\\
+Egypt & SARI/ILI & 2010--2017 & 6 & 5 & 40, 90, 97.5 & \cite{AbdElGawad2020}\\
+Egypt & SARI & 2013--2015 & 3 & 10 & ? & \cite{Elhakim2019}\\
+England & ILI & 2010--2016 & 6 & 5 & ? & \cite{Wagner2018}\\
+Finland & influenza & 2011--2016 & 5 & 6 & ? & \cite{Pesaelae2019}\\
+Montenegro & ILI & 2010--2018 & 7 & 4 & 40, 90, 97.5 & \cite{Rakocevic2019}\\
+Morocco & ILI & 2005--2017 & 11 & 3 & 40, 90, 97.5 & \cite{Rguig2020}\\
+Netherlands & RSV & 2005--2017 & 12 & 3 & 40, 90, 97.5 & \cite{Vos2019}\\
+Norway & influenza & 2006--2015 & 9 & 3 & ? & \cite{Benedetti2019}\\
+Pakistan & ILI, SARI & 2008--2017 & 10 & 3 & 40, 90, 97.5 & \cite{Nisar2020}\\
+Portugal & ILI & 2012--2017 & 5 & 6 & 40, 90, 97.5 & \cite{Pascoa2018}\\
+Scotland & influenza & 2010--2018 & 7 & 4 & ? & \cite{Murray2018}\\
+Slovenia & RSV & 2008--2018 & 10 & 3 & 40, 90, 97.5 & \cite{Grilc2020}\\
+Spain (17 regions) & ILI & 2003--2015 & 4--10 & 3--8 & 40, 90, 97.5 & \cite{Bangert2017}\\
+Spain & ILI & 2001--2018 & 16 & 2 & 40, 90, 97.5 & \cite{RedondoBravo2020}\\
+Tunisia & ILI & 2009-2018 & 9 & 3 & 50, 90, 95 & \cite{Bouguerra2020}\\
+United Kingdom & ILI & 2000--2013 & 10 & 3 & 40, 90, 97.5 & \cite{Green2015}\\
+United Kingdom & ILI/RSV & 2011--2018 & 4--6 & 5--8 & 40, 90, 97.5 & \cite{Harcourt2019}\\
+USA & ILI/influenza & 2003--2015 & 11 & 3 & 50, 90, 98 & \cite{Biggerstaff2017}\\
+USA & ILI & 2010--2015 & 5 & 6 & 50, 90, 98 & \cite{Dahlgren2018}\\
+USA & influenza & 2010--2016 & 6 & 5 & 50, 90, 98 & \cite{Dahlgren2019}\\
+\bottomrule\\
+\multicolumn{7}{c}{(b) WHO method}\\ \\
+\toprule
+region & disease & years covered & $n$ & $K$ & percentiles & authors\\
+\midrule
+Cambodia & ILI & 2009--2015 & 7 & 1 & mean, 90, 95 & \cite{Ly2017}\\
+Morocco & ILI & 2005--2017 & 11 & 1 & 40, 90, 97.5 & \cite{Rguig2020}\\
+Philippines & ILI & 2006--2012 & 7 & 1 & 90 & \cite{Lucero2016}\\
+Victoria/Australia & ILI & 2002--2011 & 6--10 & 1 & 90, 95 & \cite{Tay2013}\\
+\end{tabular}
+\end{table}
+
+% Cite Benedetti who use it as a "gold standard". Mention Green GB who compare to percentile method. Vos mention fixed criterion model -- what is that?
 
@@
 
+\subsection{Goal}
+
+As detailed in the previous section, the MEM and WHO methods have been applied extensively in practice and overall seems to produce thresholds which are considered useful by domain experts.
+A more statistical assessment, however has so far been missing.
+We perform a simulation study to compare four versions of the general approach described in Section \ref{sec:definitions}:
+\begin{enumerate}
+\item[(a)] No transformation, $K = 1$. This corresponds to the WHO method.
+\item[(b)] No transformation, $K = 30/n$.
+\item[(c)] Logarithmic transformation, $K = 1$.
+\item[(d)] Logarithmic transformation, $K = 30/n$. This corresponds to the MEM approach.
+\end{enumerate}
+
+Our main focus is on the share of seasons classified as low, medium, high and very high.
+As mentioned before, the rationale is to set thresholds ``in such a way that about 50--60\% of the season peaks should be above the moderate threshold, $\pm$ 10\% above the high threshold and $\pm 2.5\%$ above the extraordinary threshold'' \citep{WHO2017}.
+Concerning the choice between an identity and logarithmic transformation -- or, equivalently, the choice between arithmetic and geometric mean -- one might expect that the logarithm is more appropriate if observations are skewed.
+The choice $K = 30/n$ has the advantage of ensuring a reasonably large number of reference observations.
+However, this comes at the price of using observations which are not actually from peak weeks, and will by definition be lower than the actual peeks.
+If $n$ is low, this may lead to a larger number of seasons classified as high or very high intensity.
+We will assess empirically whether such an effect can be observed, and how pronounced it is depending on the number of vailable historical seasons.
+
 \subsection{Permutation approach based on historical ILI data}
+
+A pure simulation study requires careful specification of a data-generating process.
+Given the rather irregular timing and variability in peak values between influenza seasons, this is not a trivial task.
+We therefore resort to a permutation approach based on historical data in order to closely mimic the actual patterns.
+
+Assume we have $N$ years of historical data on a measure of influenza activity. For each value $n = 5, \dots, 15$ we then perform the following steps a total of 500 iterations:
+
+\begin{itemize}
+\item Sample an ordered sequence of $n$ seasons from the $N$ available seasons.
+As commonly done in bootstrap experiments, this is done with equal probability for each season and \textit{with replacement}, meaning that the same season can appear more than once in the sampled sequence.
+This approach is also referred to as the \textit{seasonal block bootstrap} \citep{Politis2001}.
+\item Apply appproaches (a)--(d) as described in the previous subsection to obtain intensity thresholds for medium (40th percentile), high (90th percentile) and very high intensity (97.5th percentile).
+\item Compute which fraction of the $N$ available seasons would be classified as low, moderate, high and very high.
+\end{itemize}
+
+Then the obtained thresholds and shares of the different intensity levels can be assessed as a function of the number $n$ of seasons used in the computations.
 

### 2021-01-16 Simulations/simulations_mem_ch.R
Re-run simulations with subsetting of paths instead of full re-sampling for different n

@@ -1,12 +1,16 @@
 ############################################################
 # Running permutation/simulation study for Switzerland
 
+library(here)
 library(mem)
 source("functions_simulations_mem.R")
 
 dat_for_mem <- read.csv("../Data/for_mem/ili_mem_ch.csv")
 
-plot(dat_for_mem$`inc.2000/2001`)
+# compute peaks of seasons:
+peaks_test <- apply(dat_for_mem, 2, max)
+
+plot(dat_for_mem$season.2000.2001)
 
 # simulation settings:
 range_i.seasons <- 5:15 # range to be explored for i.seasons
@@ -26,19 +30,19 @@ thresholds_ch_0.9 <- thresholds_ch_0.975 <-
 # run simulation:
 for(k in 1:n_sim){
   set.seed(k)
+  
+  # generate bootstrapped time series:
+  inds_train <- sample(1:ncol(dat_for_mem), size = max(range_i.seasons), replace = TRUE)
+  dat_train <- dat_for_mem[, inds_train]
 
   # run through different numbers of past seasons used:
   for(a in seq_along(range_i.seasons)){
 
-    # split into training and test:
-    inds_train <- sample(1:ncol(dat_for_mem), size = range_i.seasons[a], replace = TRUE)
-    dat_train <- dat_for_mem[, inds_train]
-    dat_test <- dat_for_mem # [, -inds_train]
-    # compute peaks of test seasons:
-    peaks_test <- apply(dat_test, 2, max)
+    # restrict to first a seasons:
+    dat_train_temp <- dat_train[, 1:range_i.seasons[a]]
 
     # run memmodel:
-    memmodel_temp <- memmodel(dat_train, i.seasons = range_i.seasons[a],
+    memmodel_temp <- memmodel(dat_train_temp, i.seasons = range_i.seasons[a],
                               i.level.intensity = c(0.4, 0.9, 0.975))
 
     # store thresholds:
@@ -91,15 +95,16 @@ thresholds1_ch_0.9 <- thresholds1_ch_0.975 <-
 for(k in 1:n_sim){
 
   set.seed(k)
+  
+  # generate bootstrapped time series:
+  inds_train <- sample(1:ncol(dat_for_mem), size = max(range_i.seasons), replace = TRUE)
+  dat_train <- dat_for_mem[, inds_train]
 
   # run through different numbers of past seasons used:
   for(a in seq_along(range_i.seasons)){
[…]

### 2021-01-16 Simulations/simulations_mem_es.R

@@ -1,11 +1,16 @@
 ############################################################
 # Running permutation/simulation study for Spain
-setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+
+library(here)
+# setwd("/home/johannes/Documents/Ideas/mem/Simulations")
 library(mem)
 
 source("functions_simulations_mem.R")
 
-dat_for_mem <- read.csv("../Data/ili_mem_es.csv")
+dat_for_mem <- read.csv("../Data/for_mem/ili_mem_es.csv")
+
+# compute peaks of seasons:
+peaks_test <- apply(dat_for_mem, 2, max)
 
 head(dat_for_mem)
 
@@ -14,6 +19,7 @@ plot(dat_for_mem[, 4])
 # simulation settings:
 range_i.seasons <- 5:15 # range to be explored for i.seasons
 n_sim <- 500
+
 
 #####################################################
 ### Run simulation using default settings of mem
@@ -26,29 +32,30 @@ thresholds_es_0.9 <- thresholds_es_0.975 <-
   exceedance_es_0.4 <- exceedance_es_0.9 <- exceedance_es_0.975 <-
   thresholds_es_0.4
 
+
 # run simulation:
 for(k in 1:n_sim){
   set.seed(k)
-
+  
+  # generate bootstrapped time series:
+  inds_train <- sample(1:ncol(dat_for_mem), size = max(range_i.seasons), replace = TRUE)
+  dat_train <- dat_for_mem[, inds_train]
+  
   # run through different numbers of past seasons used:
   for(a in seq_along(range_i.seasons)){
-
-    # split into training and test:
-    inds_train <- sample(1:ncol(dat_for_mem), size = range_i.seasons[a], replace = TRUE)
-    dat_train <- dat_for_mem[, inds_train]
-    dat_test <- dat_for_mem # [, -inds_train]
-    # compute peaks of test seasons:
-    peaks_test <- apply(dat_test, 2, max)
-
+    
+    # restrict to first a seasons:
+    dat_train_temp <- dat_train[, 1:range_i.seasons[a]]
+    
     # run memmodel:
-    memmodel_temp <- memmodel(dat_train, i.seasons = range_i.seasons[a],
+    memmodel_temp <- memmodel(dat_train_temp, i.seasons = range_i.seasons[a],
                               i.level.intensity = c(0.4, 0.9, 0.975))
-
+    
     # store thresholds:
[…]

### 2021-01-16 Simulations/simulations_mem_fr.R

@@ -1,17 +1,21 @@
 ############################################################
 # Running permutation/simulation study for Switzerland
 
-setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+setwd("/home/johannes/Documents/Projects/mem/Simulations")
 
 library(mem)
 source("functions_simulations_mem.R")
 
 dat_for_mem <- read.csv("../Data/for_mem/ili_mem_fr.csv")
 
+# compute peaks of seasons:
+peaks_test <- apply(dat_for_mem, 2, max)
+
 
 # simulation settings:
 range_i.seasons <- 5:15 # range to be explored for i.seasons
 n_sim <- 500
+
 
 
 #####################################################
@@ -28,26 +32,26 @@ thresholds_fr_0.9 <- thresholds_fr_0.975 <-
 # run simulation:
 for(k in 1:n_sim){
   set.seed(k)
-
+  
+  # generate bootstrapped time series:
+  inds_train <- sample(1:ncol(dat_for_mem), size = max(range_i.seasons), replace = TRUE)
+  dat_train <- dat_for_mem[, inds_train]
+  
   # run through different numbers of past seasons used:
   for(a in seq_along(range_i.seasons)){
-
-    # split into training and test:
-    inds_train <- sample(1:ncol(dat_for_mem), size = range_i.seasons[a], replace = TRUE)
-    dat_train <- dat_for_mem[, inds_train]
-    dat_test <- dat_for_mem#[, -inds_train]
-    # compute peaks of test seasons:
-    peaks_test <- apply(dat_test, 2, max)
-
+    
+    # restrict to first a seasons:
+    dat_train_temp <- dat_train[, 1:range_i.seasons[a]]
+    
     # run memmodel:
-    memmodel_temp <- memmodel(dat_train, i.seasons = range_i.seasons[a],
-                            i.level.intensity = c(0.4, 0.9, 0.975))
-
+    memmodel_temp <- memmodel(dat_train_temp, i.seasons = range_i.seasons[a],
+                              i.level.intensity = c(0.4, 0.9, 0.975))
+    
     # store thresholds:
     thresholds_fr_0.4[k, a] <- memmodel_temp$intensity.thresholds[1]
     thresholds_fr_0.9[k, a] <- memmodel_temp$intensity.thresholds[2]
[…]

### 2021-01-16 Simulations/simulations_mem_us.R

@@ -1,18 +1,21 @@
 ############################################################
 # Running permutation/simulation study for Spain
-setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+setwd("/home/johannes/Documents/Projects/mem/Simulations")
 library(mem)
-library(HIDDA.forecasting)
 
 source("functions_simulations_mem.R")
 
 dat_for_mem <- read.csv("../Data/for_mem/ili_mem_us.csv")
+
+# compute peaks of seasons:
+peaks_test <- apply(dat_for_mem, 2, max)
 
 plot(dat_for_mem[, 2])
 
 # simulation settings:
 range_i.seasons <- 5:15 # range to be explored for i.seasons
 n_sim <- 500
+
 
 #####################################################
 ### Run simulation using default settings of mem
@@ -25,29 +28,30 @@ thresholds_us_0.9 <- thresholds_us_0.975 <-
   exceedance_us_0.4 <- exceedance_us_0.9 <- exceedance_us_0.975 <-
   thresholds_us_0.4
 
+
 # run simulation:
 for(k in 1:n_sim){
   set.seed(k)
-
+  
+  # generate bootstrapped time series:
+  inds_train <- sample(1:ncol(dat_for_mem), size = max(range_i.seasons), replace = TRUE)
+  dat_train <- dat_for_mem[, inds_train]
+  
   # run through different numbers of past seasons used:
   for(a in seq_along(range_i.seasons)){
-
-    # split into training and test:
-    inds_train <- sample(1:ncol(dat_for_mem), size = range_i.seasons[a], replace = TRUE)
-    dat_train <- dat_for_mem[, inds_train]
-    dat_test <- dat_for_mem # [, -inds_train]
-    # compute peaks of test seasons:
-    peaks_test <- apply(dat_test, 2, max)
-
+    
+    # restrict to first a seasons:
+    dat_train_temp <- dat_train[, 1:range_i.seasons[a]]
+    
     # run memmodel:
-    memmodel_temp <- memmodel(dat_train, i.seasons = range_i.seasons[a],
+    memmodel_temp <- memmodel(dat_train_temp, i.seasons = range_i.seasons[a],
                               i.level.intensity = c(0.4, 0.9, 0.975))
-
+    
     # store thresholds:
     thresholds_us_0.4[k, a] <- memmodel_temp$intensity.thresholds[1]
[…]

### 2021-01-16 Simulations/plot_timeseries.R
Updating time series plot.

@@ -13,7 +13,7 @@ matr <- dat_ch
 xlim <- c(1990, 2020)
 plot_mem_ts <- function(matr, xlim, ylab = "inc / 100,000", ...){
   years <- as.numeric(substr(colnames(matr), start = 13, stop = 16)) - 1
-  plot(NULL, xlim = xlim, ylim = c(0, 1.1*max(matr)), xlab = "", ylab = ylab, ...)
+  plot(NULL, xlim = xlim, ylim = c(0, 1.2*max(matr)), xlab = "", ylab = ylab, ...)
   abline(v = seq(from = floor(xlim[1]), to = ceiling(xlim[2])), col = "grey")
 
   weeks_per_seas <- nrow(matr)
@@ -30,17 +30,19 @@ bplot_ranks <- function(matr, ylab = ""){
   for(i in 1:6) boxplot(apply(matr, 2, get_nth, n = i), at = i, add = TRUE, axes = FALSE)
 }
 
+pdf("../Draft/figure/plot_data.pdf", width = 9, height = 5.5)
 par(las = 1, mar = c(3, 4, 0.5, 1))
 layout(matrix(1:8, ncol = 2, byrow = TRUE), widths = c(8, 2))
-plot_mem_ts(dat_fr, xlim = c(1986, 2019)); legend("topleft", legend = "France", bty = "n", cex = 1.2)
+plot_mem_ts(dat_fr, xlim = c(1985.5, 2019.5)); legend("topleft", legend = " France", bty = "n", cex = 1.2)
 bplot_ranks(dat_fr)
 
-plot_mem_ts(dat_es, xlim = c(1986, 2019)); legend("topleft", legend = "Spain", bty = "n", cex = 1.2)
+plot_mem_ts(dat_es, xlim = c(1998.5, 2019.5)); legend("topleft", legend = " Spain", bty = "n", cex = 1.2)
 bplot_ranks(dat_es)
 
-plot_mem_ts(dat_ch, xlim = c(1986, 2019)); legend("topleft", legend = "Switzerland", bty = "n", cex = 1.2)
+plot_mem_ts(dat_ch, xlim = c(2000.5, 2015.5)); legend("topleft", legend = " Switzerland", bty = "n", cex = 1.2)
 bplot_ranks(dat_ch)
 
[…]

### 2021-01-16 Draft/Draft_MEM.tex
Updating manuscript materials.

@@
 
-\title{A permutation-based comparison of ILI intensity thresholds from the moving epidemic and WHO methods}
+\title{A bootstrap-based comparison of ILI intensity thresholds from the moving epidemic and WHO methods}
 \author{Johannes Bracher, Jonas Littek}
+\date{ \small
+Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics}
 
@@
 
-\abstract{The moving epidemic method (MEM) and the WHO method are widely used approaches to determine intensity levels for seasonal influenza and influenca-like illness (ILI).
-They are conceptually similar, but differ in two points.
-Firstly, the MEM involves a log-transformation of incidence data, while the WHO method operates on the original scale of the data.
-Secondly, the MEM method usually uses more than one observation from each past season, with a fixed total number of observations to include.
-The WHO method, on the other side, uses only the single highest value from each past season.
+\abstract{The moving epidemic method (MEM) and the WHO method are widely used approaches to determine intensity levels for seasonal influenza and influenza-like illness (ILI).
+They are conceptually similar, but differ in two aspects.
+Firstly, the MEM involves a log-transformation of incidence data, while the WHO method typically operates on the original scale of the data.
+Secondly, the MEM method by default uses more than one observation from each past season, with a fixed total number of observations to include.
+The WHO method, on the other side, uses only the single highest value from each season to determine peak intensity thresholds.
 We perform a simulation study to assess the impact of these choices on the resulting intensity thresholds and empirical exceedance proportions.
-It is based on a permutation approach using historical ILI data from France, Spain, Switzerland and the United States.
-When fixing the total number of observations to include, intensity thresholds tend to increase as more data accumulates.
+It is based on a bootstrap approach using historical ILI data from France, Spain, Switzerland and the United States.
+Using data on their original scale led to a too large proportion of seasons being classified as ``very high intensity'', which could to some degree be mitigated by using a logarithmic transformation.
+When fixing the total number of observations to include as in the MEM, intensity thresholds tend to increase as more data accumulates.
 When data is only available for few seasons, there is a rather large chance of classifying a new season as ``high'' or ``very high'' intenstity.
-We therefore advocate always using one observation per season.}
+We therefore advocate always using one observation per season with a log transformation, i.e.\ a hybrid between the default settings of the MEM and WHO methods.}
 
@@
 Following the 2009 influenza H1N1 pandemic, the need for a rapid assessment tool for influenza intensity was recognized. In a report, the Review Committee on the Functioning of the
-International Health Regulations (2005) and on Pandemic Influenza (H1N1) 2009 recommended that member states apply and evaluate their methods for severity assessment every year, thus improving their pandemic preparedness \citep[118]{WHO2011}.
-In its subsequent Pandemic Infuenza Severity Assessment (PISA) guide (\citeyear{WHO2017}), two specific statistical methods are recommended to determine influenza intensity thresholds.
+International Health Regulations and on Pandemic Influenza (H1N1) recommended that member states apply and evaluate their methods for severity assessment every year, thus improving their pandemic preparedness \citep[118]{WHO2011}.
+In the subsequent WHO Pandemic Infuenza Severity Assessment (PISA) guideline (\citeyear{WHO2017}), two specific statistical methods are recommended to determine influenza intensity thresholds.
 Firstly, the so-called \textit{WHO method}, already described in a previous document \citep{WHO2014} and in more detail in the PISA guide; and secondly, the \textit{moving epidemic method} (MEM) \citep{Vega2013, Vega2015} which is also recommended by the European Centre for Disease Prevention and Control (e.g.\ \citealt{ECDC2017}).
 
-In particular the MEM approach has in the meantime been adopted by numerous national public health agencies and seems to deliver practically useful thresholds.
+In particular the MEM approach has in the meantime been adopted by numerous national public health agencies (see references in Section \ref{sec:recent_applications}) and seems to deliver practically useful thresholds.
 The statistical properties of the MEM and WHO methods, however, have not yet been studied in detail.
 While it seems challenging to obtain strong mathematical results on their behaviour, we argue that simulation experiments can improve our understanding and intuition of the consequences of different implementation choices.
-We here provide such a simulation study, which in order to ensure a close link to real influenza patterns, is based on permutations of historical surveillance data.
+We here provide such a simulation study, which in order to ensure a close link to real influenza patterns, is based on bootstrapping of historical surveillance data.
 
@@
 
-In the following we focus on the construction of thresholds  for peak intensity, framing the MEM and WHO methods as different special cases of a more general approach.
+In the following we focus on the construction of thresholds  for peak intensity, framing the MEM and WHO methods as different special cases of the same general approach.
 Details on the construction of baseline thresholds from pre-season observations are omitted.
-We assume that thresholds are to be constructed from data from $n$ past seasons and subsequently be applied to the current ($n + 1$-th) season.
-\cite{Vega2015} recommend to use $5 \leq n \leq 10$ seasons to ensure a sufficiently large, but recent data basis.
+We assume that thresholds are based on data from $n$ past seasons and applied to the ($n + 1$)-th season.
+\cite{Vega2015} recommend to use $5 \leq n \leq 10$ seasons for the MEM to ensure a sufficiently large, but recent data basis.
 Thresholds for for a given intensity measure $X$ are then obtained via the following steps:
@@
 \begin{enumerate}
-\item For each historical season order all observations, denoting the $i$-th largest observation from season $j$ by $X^{(i)}_j$.
-\item Select the $K$ largest observations from each of the $n$ past seasons to construct a reference set $\mathcal{X} = \{X_j^{(i)}; i = 1, \dots, n; j = 1, \dots, K\}$.
-\item Apply an (inversible) transformation $Y_j^{(i)} = f(X_j^{(i)})$ to all members of the reference set to obtain a reference set $\mathcal{Y}$ of transformed past observations.
-\item Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute its estimated mean $\hat\mu$ and standard deviation $\hat \sigma$.
-\item Define intensity thresholds on the transformed scale as quantiles of this normal distribution.
+\item Within each historical season order all observations, denoting the $i$-th largest observation from season $j$ by $X^{(i)}_j$.
+\item Select the $K$ largest observations from each of the $n$ past seasons to construct a reference set $\mathcal{X} = \{X_j^{(i)}; j = 1, \dots, n; i = 1, \dots, K\}$.
+\item Apply an (inversible) transformation $Y_j^{(i)} = f(X_j^{(i)})$ to all members of the reference set to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
+\item Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates $\hat\mu, \hat \sigma$ of its mean and standard deviation.
+\item Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\hat\mu, \hat\sigma^2)$.
 A common choice for both methods \citep{WHO2017}, also implemented as the default in the \texttt{R} package \texttt{mem} \citep{Lozano2020} is
@@
 \end{itemize}
-\item Obtain thresholds on the original scale by applying the inverse transformation $q_{0.4} = f^{-1}(q_{Y, 0.4})$, $q_{0.9} = f^{-1}(q_{Y, 0.9})$, $q_{0.975} = f^{-1}(q_{Y, 0.975})$.
+\item Obtain thresholds on the original scale by applying the inverse transformation, i.e. setting $q_{0.4} = f^{-1}(q_{Y, 0.4})$, $q_{0.9} = f^{-1}(q_{Y, 0.9})$, $q_{0.975} = f^{-1}(q_{Y, 0.975})$.
 \end{enumerate}
@@
 \item In the MEM method, $f$ is chosen as the natural logarithm, while the number of observations per season is set to $K = 30/n$.
-(The exact specification used in the \texttt{R} package \texttt{mem} is $K = \max(1, \lfloor 30/n \rceil)$, where $\lfloor \cdot \rceil$ denotes rounding to the nearest integer.) The total nuber of historical observations is thus kept (approximately) fixed.
+(The exact specification used in the \texttt{R} package \texttt{mem} is $K = \max(1, \lfloor 30/n \rceil)$, where $\lfloor \cdot \rceil$ denotes rounding to the nearest integer.) The total number of historical observations is thus kept (approximately) fixed.
 \item In the WHO method, the standard procedure is to set $f$ to the identity function (i.e.\ no transformation is applied).
@@
 
-\noindent The rationale behind this approach is that ``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold \citep{WHO2017}.
-In most previous descriptions (e.g., \ \citealt{WHO2014, Vega2015}), the thresholds have been introduced as upper ends of one-sided confidence intervals associated with the the arithmetic (WHO method) or geometric mean (MEM) of the historical reference observations in $\mathcal{X}$.
-This, however, is slightly imprecise terminology as in the computations $\sigma$ rather than $\sigma/\sqrt{n}$ is used.
-If indeed confidence intervals were used, these would get narrower the more historical seasons were available, and the thresholds for all levels would eventually converge to the arithmetic / geometric mean.
+\noindent The rationale of this approach is that ``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold'' \citep{WHO2017}.
+Note that in most previous descriptions (e.g., \ \citealt{WHO2014, Vega2015}), the thresholds have been referred to as upper ends of one-sided confidence intervals associated with the the arithmetic (WHO method) or geometric mean (MEM) of the historical reference observations in $\mathcal{X}$.
+This, however, is slightly imprecise terminology as in the computations $\hat\sigma$ rather than $\hat\sigma/\sqrt{nK}$ is used.
+If indeed confidence intervals were used, these would tend to get narrower the more historical seasons were available, and the thresholds for all levels would eventually converge to the arithmetic / geometric mean.
 This is obviously not the intended behaviour.
@@
 
-To get a better understanding of the different settings in which the MEM and WHO methods are applied in practice we performed a literature search among all published articles citing the papers \cite{Vega2015}, \cite{WHO2014} and \cite{WHO2017}.
+To get a better understanding of the different settings in which the MEM and WHO methods are applied in practice we performed a literature search among all published articles citing the papers \cite{Vega2015}, \cite{WHO2014} and \cite{WHO2017} until December 2020 (identified via \textit{CrossRef} and \textit{Google Scholar}).
 The results are summarized in Table \ref{tab:literature}.
-As can be seen from the numerous entries from the years 2019 and 2020, the MEM has quickly become the standard in the determination of intensity thresholds for influenza and other respiratory diseases.
+As can be seen from the numerous entries from the years 2019 and 2020, the MEM has quickly become a standard approach in the determination of intensity thresholds for influenza and other respiratory diseases.
+Indeed, the contributions come from numerous countries and in many cases have been co-authored by representatives of national or regional public health agencies.
 In most analyses, the suggested threshold levels at the 40th, 90th and 97.5th percentile are used.
-The contributions come from numerous countries and in many cases have been co-authors by representatives of national or regional public health agencies.
-Variability with respect to the number $n$ of historical seasons included is considerable, with a range from 3 to 16 years.
-Consequently, the number of observations included per season ranges from two to ten.
+Variability with respect to the number $n$ of historical seasons included is considerable, with a range from 3 to 16 seasons.
+Consequently, the number $K$ of observations included per season ranges from two to ten.
+We only found few applications of the WHO methods, one of them providing a comparison to the thresholds from the MEM method \citep{Rguig2020}.
 
@@
 \subsection{Goal}
+\label{subsec:simulation_goal}
 
-As detailed in the previous section, the MEM and WHO methods have been applied extensively in practice and overall seems to produce thresholds which are considered useful by domain experts.
+As detailed in the previous section, the MEM and WHO methods have been applied extensively in practice and overall seem to produce thresholds which are considered useful by domain experts.
 A more statistical assessment, however has so far been missing.
@@
 \item[(c)] Logarithmic transformation, $K = 1$.
-\item[(d)] Logarithmic transformation, $K = 30/n$. This corresponds to the MEM approach.
+\item[(d)] Logarithmic transformation, $K = 30/n$. This corresponds to the default settings of the MEM approach.
 \end{enumerate}
 
-Our main focus is on the share of seasons classified as low, medium, high and very high.
-As mentioned before, the rationale is to set thresholds ``in such a way that about 50--60\% of the season peaks should be above the moderate threshold, $\pm$ 10\% above the high threshold and $\pm 2.5\%$ above the extraordinary threshold'' \citep{WHO2017}.
-Concerning the choice between an identity and logarithmic transformation -- or, equivalently, the choice between arithmetic and geometric mean -- one might expect that the logarithm is more appropriate if observations are skewed.
-The choice $K = 30/n$ has the advantage of ensuring a reasonably large number of reference observations.
+Our main focus is on the average share of seasons classified as low, medium, high and very high.
+As mentioned before, the medium, high and extraordinary threholds should be exceeded by approximately 60\%, 10\% and 2.5\% respectively \citep{WHO2017}.
+Concerning the choice between an identity and logarithmic transformation -- or, equivalently, the choice between arithmetic and geometric mean -- one may expect that the logarithm is more appropriate if observations are skewed.
+Concerning the number of observations used per available season, the choice $K = 30/n$ has the advantage of ensuring a reasonably large number of reference observations.
 However, this comes at the price of using observations which are not actually from peak weeks, and will by definition be lower than the actual peeks.
-If $n$ is low, this may lead to a larger number of seasons classified as high or very high intensity.
-We will assess empirically whether such an effect can be observed, and how pronounced it is depending on the number of vailable historical seasons.
+If $n$ is low, this may lead to too low thresholds and a larger number of seasons classified as high or very high intensity.
+We will assess empirically whether such an effect can be observed, and how pronounced it is depending on the number of available historical seasons.
 
-\subsection{Permutation approach based on historical ILI data}
+\subsection{Bootstrap approach based on historical ILI data}
 
-A pure simulation study requires careful specification of a data-generating process.
-Given the rather irregular timing and variability in peak values between influenza seasons, this is not a trivial task.
-We therefore resort to a permutation approach based on historical data in order to closely mimic the actual patterns.
-
-Assume we have $N$ years of historical data on a measure of influenza activity. For each value $n = 5, \dots, 15$ we then perform the following steps a total of 500 iterations:
+A pure simulation study requires careful specification of a data-generating process mimicking the seasonal patterns of influenza.
+Given the considerable variability in timing and intensity of season peaks, this is not a trivial task.
+We therefore resort to a bootstrap approach based on historical data rather than generating synthetic data.
+Assume we have $N$ years of historical data on a measure of influenza activity.
+We then repeat the following steps a total of 500 times:
 
 \begin{itemize}
-\item Sample an ordered sequence of $n$ seasons from the $N$ available seasons.
+\item Sample an ordered sequence of 15 seasons from the $N$ available seasons.
 As commonly done in bootstrap experiments, this is done with equal probability for each season and \textit{with replacement}, meaning that the same season can appear more than once in the sampled sequence.
 This approach is also referred to as the \textit{seasonal block bootstrap} \citep{Politis2001}.
-\item Apply appproaches (a)--(d) as described in the previous subsection to obtain intensity thresholds for medium (40th percentile), high (90th percentile) and very high intensity (97.5th percentile).
-\item Compute which fraction of the $N$ available seasons would be classified as low, moderate, high and very high.
+\item For each value $n = 5, \dots, 15$:
+\begin{itemize}
+\item Restrict the generated sequence to the first $n$ seasons.
+\item Apply approaches (a)--(d) as described in the previous subsection.
+Compute and store intensity thresholds for medium (40th percentile), high (90th percentile) and very high intensity (97.5th percentile).
+\item Compute which fraction of the $N$ available seasons would be classified as low, moderate, high and very high using the obtained thresholds.
 \end{itemize}
-
-Then the obtained thresholds and shares of the different intensity levels can be assessed as a function of the number $n$ of seasons used in the computations.
+\end{itemize}
+We can then assess how the thresholds and shares of the different intensity levels depend on the employed transformation function, the number of observations $K$ used per season and the number $n$ of seasons included.
 
@@
 
+We use publicly available influenza surveillance data from four countries.
+Data on the weekly incidence of influenza-like-illness in France, 1985--2019, were obtained from the Réseau Sentinelles web platform (INSERM/Sorbonne Université, \url{https://www.sentiweb.fr}, \citealt{Flahault2006}).
+Population data were obtained from Institut national de la statistique et des études économiques.
+Data on weekly confirmed influenza cases in Spain, 1998--2019, were extracted from graphs shown in the \textit{Informe Anual} of the \textit{Sistema de la vigilancia de gripe en Espa\~na} (\url{https://vgripe.isciii.es/}; \citealt{SVGE2019}) using the tool \textit{WebPlotDigitizer} (\url{https://apps.automeris.io/wpd/}).
+Weekly ILI counts from Swiss routine surveillance, collected by the Swiss Federal Office of Public Health, 2000--2016, were obtained from the R package \texttt{HIDDA.forecasting} \citep{Held2019}.
+Population data were obtained from the Swiss Federal Statistical Office.
+Weekly weighted ILI percentage rates (wILI; corresponding to the percentage of general practitioner visits due to influenza-like symptoms) in the US, 1998--2017, from CDC FluView \cite{Charbonneau2019} were obtained from the CDC FluSight influenza forecasting platform (\url{https://github.com/FluSightNetwork/cdc-flusight-ensemble/}).
+
+All four time series are displayed in Figure \ref{fig:data}.
+Note that we removed the pandemic 2009/2010 influenza season from each series as it is not comparable to the usual years with only seasonal influenza.
+We also show boxplots of the first through sixth largest observation per season (right column).
+In the notation introduced in Section \ref{sec:definitions}, the first boxplot thus shows observations $X_j^{(1)}, j = 1, \dots, N$, the second $X_j^{(2)}, j = 1, \dots, N$ and so on.
+Little surprisingly (and by definition), the value get smaller on average for increasing ranks.
+Interestingly, they also get less dispersed, meaning that the variability between e.g.\ the sixth largest observations of the different seasons is always much smaller than that of the largest observations.
+
+% mention exact definition, source, number of seasons
+
+\begin{figure}
+\center
 \includegraphics[width=0.9\textwidth]{figure/plot_data.pdf}
+\caption{Time series of influenza activity measures in four countries: Weekly ILI cases per 100,000 population in France, 1985--2019; weekly number of confirmed influenza cases per 100,000 population in Spain, 1998--2019; weekly ILI cases per 100,000 in Switzerland, 2000--2016; weekly wILI percentages in the United States, 1998--2017.}
+\label{fig:data}
+\end{figure}
 
@@
 
-MEM version xxx, R software
+All of the following analyses were performed using the R language for statistical computing, version 4.0.3 \citep{RCT2020}, particular using the package \texttt{mem}, version 2.16 \cite{Lozano2020}.
+Note that the function \texttt{mem::memmodel} provides sufficient flexibility to apply all four approaches (a)--(d) from Section \ref{subsec:simulation_goal}.
+\todo{check whether app allows to change settings.}
 
@@
 
+Results from our bootstrap simulation study for the four considered countries are shown in Figures \ref{fig:results1} and \ref{fig:results2}
+
+\begin{figure}
 \includegraphics[page=1, width=0.9\textwidth]{figure/plot_results.pdf}
+\label{fig:results1}
+\end{figure}
 
+\begin{figure}
 \includegraphics[page=2, width=0.9\textwidth]{figure/plot_results.pdf}
+\caption{Results of bootstrap simulation study for Spain and the United States.
+First and third column: mean intensity thresholds along with bands delimited by the empirical 5\% and 95\% quantiles.
+SEcond and fourth columns: resulting average shares of seasons classified as low, medium, high and very high intensity.}
+\label{fig:results2}
+\end{figure}
 
@@
 \label{sec:discussion}
+
+Mention that also concerns computation of seasonal thresholds.
 
@@
 \end{itemize}
-\item Simulation study based on permutation of true seasons:
+\item Simulation study based on bootstrapping of true seasons:
 \begin{itemize}

### 2021-01-27 Draft/Draft_MEM.tex
Updating manuscript.

@@
 \usepackage{booktabs}
+\usepackage{amssymb}
+\usepackage{amsmath}
 
@@
 The WHO method, on the other side, uses only the single highest value from each season to determine peak intensity thresholds.
-We perform a simulation study to assess the impact of these choices on the resulting intensity thresholds and empirical exceedance proportions.
-It is based on a bootstrap approach using historical ILI data from France, Spain, Switzerland and the United States.
-Using data on their original scale led to a too large proportion of seasons being classified as ``very high intensity'', which could to some degree be mitigated by using a logarithmic transformation.
-When fixing the total number of observations to include as in the MEM, intensity thresholds tend to increase as more data accumulates.
-When data is only available for few seasons, there is a rather large chance of classifying a new season as ``high'' or ``very high'' intenstity.
-We therefore advocate always using one observation per season with a log transformation, i.e.\ a hybrid between the default settings of the MEM and WHO methods.}
+We provide statistical arguments and a simulation study to assess the impact of these choices on the resulting intensity thresholds and empirical exceedance proportions.
+The simulation study is based on a bootstrap approach using historical ILI data from France, Spain, Switzerland and the United States.
+We find that using data on the original scale leads to a too large proportion of seasons being classified as high or ``very high intensity intensity, which can to some degree be mitigated by using a logarithmic transformation.
+When fixing the total number of observations to include as in the MEM, intensity thresholds tend to increase as more historical data becomes available.
+When data is only available for few seasons, there is a rather large chance of classifying a new season as high or very high intenstity.
+We therefore advocate always using one observation per season with a log transformation, i.e.\ a hybrid between the default settings of the MEM and WHO methods.
+This adaptation is possible within an existing software implementation of the MEM approach and thus easy to apply.}
 
@@
 Following the 2009 influenza H1N1 pandemic, the need for a rapid assessment tool for influenza intensity was recognized. In a report, the Review Committee on the Functioning of the
-International Health Regulations and on Pandemic Influenza (H1N1) recommended that member states apply and evaluate their methods for severity assessment every year, thus improving their pandemic preparedness \citep[118]{WHO2011}.
+International Health Regulations and on Pandemic Influenza (H1N1) recommended that member states apply and evaluate their methods for severity assessment every year, thus improving their pandemic preparedness \citep[p.118]{WHO2011}.
 In the subsequent WHO Pandemic Infuenza Severity Assessment (PISA) guideline (\citeyear{WHO2017}), two specific statistical methods are recommended to determine influenza intensity thresholds.
-Firstly, the so-called \textit{WHO method}, already described in a previous document \citep{WHO2014} and in more detail in the PISA guide; and secondly, the \textit{moving epidemic method} (MEM) \citep{Vega2013, Vega2015} which is also recommended by the European Centre for Disease Prevention and Control (e.g.\ \citealt{ECDC2017}).
+Firstly, the so-called \textit{WHO method}, already described in a previous document \citep{WHO2014} and in more detail in the PISA guide; and secondly, the \textit{moving epidemic method} (MEM; \citealt{Vega2013, Vega2015}) which is also recommended by the European Centre for Disease Prevention and Control (e.g., \citealt{ECDC2017}).
 
 In particular the MEM approach has in the meantime been adopted by numerous national public health agencies (see references in Section \ref{sec:recent_applications}) and seems to deliver practically useful thresholds.
+Its application has been facilitated by the availablility of a user-friendly and well-documented \texttt{R} package (\texttt{mem}; \citealt{Lozano2020}), which can even be used via a graphical user interface \citep{Lozano2018}.
 The statistical properties of the MEM and WHO methods, however, have not yet been studied in detail.
-While it seems challenging to obtain strong mathematical results on their behaviour, we argue that simulation experiments can improve our understanding and intuition of the consequences of different implementation choices.
-We here provide such a simulation study, which in order to ensure a close link to real influenza patterns, is based on bootstrapping of historical surveillance data.
+We obtain some analytical results, complemented by simulation experiments.
+In order to ensure a close link to real influenza patterns, the latter are based on bootstrap re-sampling of historical surveillance data.
+This way, we provide an improved understanding and intuition for the consequences of different implementation choices.
 
@@
 Section \ref{sec:recent_applications} consists of an overview of published applications of the MEM and WHO methods.
+This is followed by a brief examination of statistical properties of the two methods in Section \ref{sec:analytical}.
 In Section \ref{sec:simulation} we describe our simulation approach and the results based on historical data from France, Spain, Switzerland and the United States.
@@
 In the following we focus on the construction of thresholds  for peak intensity, framing the MEM and WHO methods as different special cases of the same general approach.
-Details on the construction of baseline thresholds from pre-season observations are omitted.
-We assume that thresholds are based on data from $n$ past seasons and applied to the ($n + 1$)-th season.
-\cite{Vega2015} recommend to use $5 \leq n \leq 10$ seasons for the MEM to ensure a sufficiently large, but recent data basis.
+Details on the construction of baseline epidemic thresholds from pre-season observations are omitted.
+We assume that thresholds are based on data from $m$ past seasons and applied to the ($m + 1$)-th season.
+\cite{Vega2015} recommend to use $5 \leq m \leq 10$ seasons for the MEM to ensure a recent data basis.
 Thresholds for for a given intensity measure $X$ are then obtained via the following steps:
@@
 \begin{enumerate}
-\item Within each historical season order all observations, denoting the $i$-th largest observation from season $j$ by $X^{(i)}_j$.
-\item Select the $K$ largest observations from each of the $n$ past seasons to construct a reference set $\mathcal{X} = \{X_j^{(i)}; j = 1, \dots, n; i = 1, \dots, K\}$.
-\item Apply an (inversible) transformation $Y_j^{(i)} = f(X_j^{(i)})$ to all members of the reference set to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
+\item Within each historical season $j = 1, \dots, m$ order all observations in decreasing order, denoting the $i$-th largest observation from season $j$ by $X^{(j)}_i$.
+\item Select the $n$ largest observations from each of the $m$ past seasons to construct a reference set $\mathcal{X} = \{X_i^{(j)}; j = 1, \dots, m; i = 1, \dots, n\}$.
+\item Apply an (invertible) transformation $Y_i^{(j)} = f(X_i^{(j)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
 \item Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates $\hat\mu, \hat \sigma$ of its mean and standard deviation.
-\item Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\hat\mu, \hat\sigma^2)$.
-A common choice for both methods \citep{WHO2017}, also implemented as the default in the \texttt{R} package \texttt{mem} \citep{Lozano2020} is
+\item Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\hat\mu, \hat\sigma^2)$, i.e.\ compute
+$$
+q_{Y, \alpha} = \hat{\mu} + z_\alpha \hat{\sigma},
+$$
+where $z_\alpha$ is the $\alpha$ quantile of the standard normal distribution.
+A common choice for both methods \citep{WHO2017}, also implemented as the default in the \texttt{R} package \texttt{mem} \citep{Lozano2020} is to use
 \begin{itemize}
@@
 \begin{itemize}
-\item In the MEM method, $f$ is chosen as the natural logarithm, while the number of observations per season is set to $K = 30/n$.
-(The exact specification used in the \texttt{R} package \texttt{mem} is $K = \max(1, \lfloor 30/n \rceil)$, where $\lfloor \cdot \rceil$ denotes rounding to the nearest integer.) The total number of historical observations is thus kept (approximately) fixed.
 \item In the WHO method, the standard procedure is to set $f$ to the identity function (i.e.\ no transformation is applied).
 If peaks vary considerably, a log-tranformation is recommended \citep{WHO2017}.
-Irrespective of the number $n$ of historical seasons, $K = 1$ observation per season is used.
+Irrespective of the number $m$ of historical seasons, $n = 1$ observation per season is used.
+\item In the MEM method, the default choice for $f$ is the natural logarithm, while the number of observations per season is set to $n = 30/m$.
+(The exact specification used in the \texttt{R} package \texttt{mem} is $n = \max(1, \lfloor 30/m \rceil)$, where $\lfloor \cdot \rceil$ denotes rounding to the nearest integer.) The total number of historical observations is thus kept (approximately) fixed, ensuring a reasonably large number of reference observations
 \end{itemize}
+Note that the software implementation of the moving epidemic method in the \texttt{R} package \texttt{mem} is rather general and permits the user to choose $n$, $m$, and $f$ (between the identify and logarithmic function).
+The term \textit{moving epidemic method} could thus also be used as an umbrella term for the general procedure described above.
+We here use it in a more narrow sense to refer to the specific procedure described in \cite{Vega2015}, \cite{WHO2017} and reflected in the default parameters of the \texttt{mem} package.
 
-\noindent The rationale of this approach is that ``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold'' \citep{WHO2017}.
+The rationale behind the percentile-based approach is that ``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold'' \citep{WHO2017}.
 Note that in most previous descriptions (e.g., \ \citealt{WHO2014, Vega2015}), the thresholds have been referred to as upper ends of one-sided confidence intervals associated with the the arithmetic (WHO method) or geometric mean (MEM) of the historical reference observations in $\mathcal{X}$.
-This, however, is slightly imprecise terminology as in the computations $\hat\sigma$ rather than $\hat\sigma/\sqrt{nK}$ is used.
+This, however, is slightly imprecise terminology as in the computations the standard deviation $\hat\sigma$ rather than the standard error $\hat\sigma/\sqrt{nm}$ of $\hat\mu$ is used.
 If indeed confidence intervals were used, these would tend to get narrower the more historical seasons were available, and the thresholds for all levels would eventually converge to the arithmetic / geometric mean.
@@
 
-To get a better understanding of the different settings in which the MEM and WHO methods are applied in practice we performed a literature search among all published articles citing the papers \cite{Vega2015}, \cite{WHO2014} and \cite{WHO2017} until December 2020 (identified via \textit{CrossRef} and \textit{Google Scholar}).
+To get a better understanding of the different settings in which the MEM and WHO methods are applied in practice we performed a literature search of articles published in English language and citing the papers \cite{Vega2015}, \cite{WHO2014} and \cite{WHO2017} until December 2020 (identified via \textit{CrossRef} and \textit{Google Scholar}).
 The results are summarized in Table \ref{tab:literature}.
@@
 In most analyses, the suggested threshold levels at the 40th, 90th and 97.5th percentile are used.
-Variability with respect to the number $n$ of historical seasons included is considerable, with a range from 3 to 16 seasons.
-Consequently, the number $K$ of observations included per season ranges from two to ten.
-We only found few applications of the WHO methods, one of them providing a comparison to the thresholds from the MEM method \citep{Rguig2020}.
+Variability with respect to the number $m$ of historical seasons included is considerable, with a range from three to 16 seasons.
+Consequently, the number $n$ of observations included per season ranges from two to ten (none of the above papers indicated a modification of the default setting $n = 30/m$ of the moving epidemic method).
+We only found three published applications of the WHO method, one of them providing a comparison to the thresholds from the MEM method \citep{Rguig2020}.
 
-\begin{table}
-\caption{Applications of the moving epidemic method and WHO method to determie severity thresholds.
-Does not contain works where only baseline thresholds are computed.
-The number of seasons included to compute thresholds is denoted by $n$, the number of observations used per season by $K$.
+\begin{table}[h!]
+\caption{Applications of the MEM and WHO method to determine intensity thresholds for respiratory diseases.
+We do not include works where only baseline thresholds are computed.
+The number of seasons included to compute thresholds is denoted by $m$, the number of observations used per season by $n$.
 Abbreviations: SARI = severe acute respiratory infection; ILI = influenza-like illness; RSV = respiratory syncytial virus.}
@@
 \toprule
-region & disease & years covered & $n$ & $K$ & percentiles & authors\\
+region & disease & years covered & $m$ & $n$ & percentiles & authors\\
 \midrule
@@
 Scotland & influenza & 2010--2018 & 7 & 4 & ? & \cite{Murray2018}\\
+Scotland & influenza & 2010--2019 & 7--8 & 4 & 40, 90, 97.5 & \cite{Dickson2020}\\
 Slovenia & RSV & 2008--2018 & 10 & 3 & 40, 90, 97.5 & \cite{Grilc2020}\\
@@
 
+\section{Analytical considerations on the choice of $n$}
+\label{sec:analytical}
+
+As detailed in the previous section, the MEM and WHO methods have been applied extensively in practice and overall seem to produce thresholds which are considered useful by domain experts.
+A more statistical assessment, however has so far been missing.
+Which choice of transformation function $f$ is most reaonable is largely an empirical question.
+Little more can be said than that a logarithmic transformation can be helpful if the distribution of peak values is skewed.
+However, some statements can be made on how the value chosen for $n$ impacts the expected results.
+
+Denote by $\mathbf{Y}^{(j)}$ the vector of the $n$ largest transformed observations from season $j$ in decreasing order, i.e.
+$\mathbf{Y}^{(j)} = (Y^{(j)}_1, \dots, Y^{(j)}_n)^\top$ with $Y^{(j)}_1 \geq Y^{(j)}_2 \geq \dots \geq Y^{(j)}_n$.
+Now assume that $\mathbf{Y}^{(1)}, \dots, \mathbf{Y}^{(m)}$ are independent and identically distributed with mean and covariance matrix
+\begin{align}
+\mathbb{E}\left(\mathbf{Y}^{(j)}\right) = \boldsymbol{\mu} = \left(\begin{array}{c}
+\mu_1\\
+\vdots\\
+\mu_n
+\end{array}\right) \ \ \ \text{and} \ \ \ \text{Cov}\left(\mathbf{Y}^{(j)}\right) = \boldsymbol{\Sigma} =
+\left(\begin{array}{ccc}
+\sigma_{1, 1} & \cdots & \sigma_{1n}\\
+\vdots & \ddots &\vdots\\
+\sigma_{n, 1} & \cdots & \sigma_{nn}
+\end{array}\right),
+\end{align}
+respectively, where to make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$. It can then be shown that the expectations of $\hat\mu$ and $\hat{\sigma}^2$ are given by
+\begin{align}
+\mathbb{E}(\hat{\mu}) & = \frac{1}{n} \sum_{i = 1}^n \mu_i,
+\label{eq:expectation_mu}\\
+\mathbb{E}(\hat{\sigma}^2) & = \frac{m}{mn - 1} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{n(mn - 1)} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ - \ \frac{m}{n(mn - 1)}\left(\sum_{i = 1}^n \mu_i\right)^2,
+\label{eq:expectation_sigma2}
+\end{align}
+%\begin{align}
+%\mathbb{E}(\hat{\mu}) & = \frac{1}{K} \sum_{i = 1}^K \mu_i,
+%\label{eq:expectation_mu}\\
+%\mathbb{E}(\hat{\sigma}^2) & = \frac{n}{nK - 1} \sum_{i = 1}^K (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{K(nK - 1)} \sum_{i = 1}^K \sum_{i' = 1}^K \sigma_{i,i'} \ - \ \frac{n}{K(nK - 1)}\left(\sum_{i = 1}^K \mu_i\right)^2,
+%\label{eq:expectation_sigma2}
+%\end{align}
+respectively. The derivation is provided in Supplement \textbf{XXX}. For reasons detailed there, if the transformation function $f$ is the identity or the natural logarithm,
+\begin{equation}
+\mathbb{E}(q_\alpha) \approx f^{-1}\left\{\mathbb{E}(\hat\mu) + z_\alpha \sqrt{\mathbb{E}(\hat\sigma^2)}\right\}
+\label{eq:expectation_q}
+\end{equation}
+usually holds in good approximation (with $f^{-1}$ the idenity or exponential function). This implies that for $n = 1$, we have
+$$
+\mathbb{E}(q_\alpha) \approx f^{-1}(\mu_1 + z_\alpha \sigma_1).
+$$
+This is a desirable property.
+Loosely speaking, if in addition the $Y_{1}^{(j)}$ come from a normal distribution, the nominal exceedance probabilities for the different thresholds would on average be achieved.
+If a different value $n > 1$ is chosen, this will generally no longer be the case.
+Equations \eqref{eq:expectation_mu}--\eqref{eq:expectation_q} tell us by how much $q_\alpha$ can be expected to differ from the intended value $f^{-1}(\mu_1 + z_\alpha \sigma_1)$.
+By the definition of the $\mu_i$ (with $\mu_i$ the expectation of the $i$-th largest observation in a given season), $\mathbb{E}(\hat\mu)$ decreases in $n$.
+In most (possibly all) settings this also translates to the $q_\alpha$.
+As a consequence, when choosing $n > 1$, one must expect to classify a larger number of season peaks as high or very high intensity.
+When choosing $n = 30/m$ as suggested for the MEM, threholds will then tend to increase the more years of data are used to compute thresholds.
+
 \section{Simulaton study}
@@
 
-As detailed in the previous section, the MEM and WHO methods have been applied extensively in practice and overall seem to produce thresholds which are considered useful by domain experts.
-A more statistical assessment, however has so far been missing.
-We perform a simulation study to compare four versions of the general approach described in Section \ref{sec:definitions}:
+To complement the analytical results from the previous section we perform a simulation study to compare four versions of the general approach described in Section \ref{sec:definitions}:
 \begin{enumerate}
-\item[(a)] No transformation, $K = 1$. This corresponds to the WHO method.
-\item[(b)] No transformation, $K = 30/n$.
-\item[(c)] Logarithmic transformation, $K = 1$.
-\item[(d)] Logarithmic transformation, $K = 30/n$. This corresponds to the default settings of the MEM approach.
+\item[(a)] No transformation, $n = 1$. This corresponds to the WHO method.
+\item[(b)] No transformation, $n = 30/m$.
+\item[(c)] Logarithmic transformation, $n = 1$.
+\item[(d)] Logarithmic transformation, $n = 30/m$. This corresponds to the MEM approach.
 \end{enumerate}
 
-Our main focus is on the average share of seasons classified as low, medium, high and very high.
-As mentioned before, the medium, high and extraordinary threholds should be exceeded by approximately 60\%, 10\% and 2.5\% respectively \citep{WHO2017}.
+Our main focus is on the resulting intensity thresholds and shares of seasons classified as low, medium, high and very high.
+We will put the analytical approximations from the previous section to the test and assess the impact of the chosen $n$ and $f$ on thresholds and exceedance rates empirically.
+%Also, we will examine the effect of applying a logarithmic transformation to the data.
+% As mentioned before, the medium, high and extraordinary threholds should be exceeded by approximately 60\%, 10\% and 2.5\% respectively \citep{WHO2017}.
 Concerning the choice between an identity and logarithmic transformation -- or, equivalently, the choice between arithmetic and geometric mean -- one may expect that the logarithm is more appropriate if observations are skewed.
@@
 We therefore resort to a bootstrap approach based on historical data rather than generating synthetic data.
-Assume we have $N$ years of historical data on a measure of influenza activity.
+Assume we have $M$ years of historical data on a measure of influenza activity.
 We then repeat the following steps a total of 500 times:
@@
 \begin{itemize}
-\item Sample an ordered sequence of 15 seasons from the $N$ available seasons.
+\item Sample an ordered sequence of 15 seasons from the $M$ available seasons.
 As commonly done in bootstrap experiments, this is done with equal probability for each season and \textit{with replacement}, meaning that the same season can appear more than once in the sampled sequence.
 This approach is also referred to as the \textit{seasonal block bootstrap} \citep{Politis2001}.
-\item For each value $n = 5, \dots, 15$:
+\item For each value $m = 5, \dots, 15$:
 \begin{itemize}
-\item Restrict the generated sequence to the first $n$ seasons.
+\item Restrict the generated sequence to the first $m$ seasons.
 \item Apply approaches (a)--(d) as described in the previous subsection.
 Compute and store intensity thresholds for medium (40th percentile), high (90th percentile) and very high intensity (97.5th percentile).
-\item Compute which fraction of the $N$ available seasons would be classified as low, moderate, high and very high using the obtained thresholds.
+\item Compute which fraction of all $M$ available historical seasons would be classified as low, moderate, high and very high using the obtained thresholds.
 \end{itemize}
 \end{itemize}
-We can then assess how the thresholds and shares of the different intensity levels depend on the employed transformation function, the number of observations $K$ used per season and the number $n$ of seasons included.
+We can then assess how the thresholds and shares of the different intensity levels depend on the employed transformation function, the number of observations $n$ used per season and the number $m$ of seasons included.
+Here, the range $m =5, \dots, 15$ is motivated by the range of values found in real-world applications of the moving epidemic method as listed in Section \ref{sec:recent_applications}.
 
@@
 Population data were obtained from the Swiss Federal Statistical Office.
-Weekly weighted ILI percentage rates (wILI; corresponding to the percentage of general practitioner visits due to influenza-like symptoms) in the US, 1998--2017, from CDC FluView \cite{Charbonneau2019} were obtained from the CDC FluSight influenza forecasting platform (\url{https://github.com/FluSightNetwork/cdc-flusight-ensemble/}).
+Weekly weighted ILI percentage rates (wILI) in the US from CDC FluView \citep{Charbonneau2019}, 1998--2017, were obtained from the CDC FluSight influenza forecasting platform (\url{https://github.com/FluSightNetwork/cdc-flusight-ensemble/}).
+These wILI values correspond to the percentage of general practitioner visits due to influenza-like symptoms
 
@@
 We also show boxplots of the first through sixth largest observation per season (right column).
-In the notation introduced in Section \ref{sec:definitions}, the first boxplot thus shows observations $X_j^{(1)}, j = 1, \dots, N$, the second $X_j^{(2)}, j = 1, \dots, N$ and so on.
-Little surprisingly (and by definition), the value get smaller on average for increasing ranks.
-Interestingly, they also get less dispersed, meaning that the variability between e.g.\ the sixth largest observations of the different seasons is always much smaller than that of the largest observations.
+In the notation introduced in Section \ref{sec:definitions}, the first boxplot thus shows observations $X_1^{(j)}, j = 1, \dots, M$, the second $X_2^{(j)}, j = 1, \dots, M$ and so on.
+Little surprisingly (and by definition), the average values get smaller for increasing ranks.
+Interestingly, they also get less dispersed, meaning that the variability between e.g.\ the sixth largest observations of the different seasons is smaller than that of the largest observations.
 
@@
 
-\begin{figure}
+\begin{figure}[h]
 \center
 \includegraphics[width=0.9\textwidth]{figure/plot_data.pdf}
-\caption{Time series of influenza activity measures in four countries: Weekly ILI cases per 100,000 population in France, 1985--2019; weekly number of confirmed influenza cases per 100,000 population in Spain, 1998--2019; weekly ILI cases per 100,000 in Switzerland, 2000--2016; weekly wILI percentages in the United States, 1998--2017.}
+\caption{Time series of influenza activity measures in four countries: Weekly ILI cases per 100,000 population in France, 1985--2019; weekly number of confirmed influenza cases per 100,000 population in Spain, 1998--2019; weekly ILI cases per 100,000 in Switzerland, 2000--2016; weekly wILI percentages in the United States, 1998--2017.
+Off-season weeks are omitted in the plot, with grey lines delimiting the different seasons.}
 \label{fig:data}
@@
 
-All of the following analyses were performed using the R language for statistical computing, version 4.0.3 \citep{RCT2020}, particular using the package \texttt{mem}, version 2.16 \cite{Lozano2020}.
+All analyses were performed using the R language for statistical computing, version 4.0.3 \citep{RCT2020}, in particular using the package \texttt{mem}, version 2.16 \citep{Lozano2020}.
 Note that the function \texttt{mem::memmodel} provides sufficient flexibility to apply all four approaches (a)--(d) from Section \ref{subsec:simulation_goal}.
-\todo{check whether app allows to change settings.}
 
@@
 
-Results from our bootstrap simulation study for the four considered countries are shown in Figures \ref{fig:results1} and \ref{fig:results2}
+Results from our bootstrap simulation study for the four considered countries are shown in Figures \ref{fig:results1} (France, Spain) and \ref{fig:results2} (Switzerland and US).
+In the first and third colum of each figure we also show the analytical approximations from Section \ref{sec:analytical} as small black dots.
+To compute these approximations, we just plugged the empirical mean vectors and covariance matrices into equations \eqref{eq:expectation_mu} -- \eqref{eq:expectation_q}.
+As can be seen, the agreement with the bootstrap results is very good.
+
+In all four countries, the thresholds obtained when using a log transformation as in the MEM are considerably higher than when no transformation is used.
+Especially in France and Spain, thresholds obtained without the log transormation are clearly too low, as can be seen from the large proportion of seasons classified as very high intensity (around 10\% when using $n = 1$).
+Applying the log transformation mitigates this to some degree.
+
+As expected, when letting the number of observations used per season depend on the number of available seasons via $n = 30/m$, average thresholds increase in $m$.
+As a particularly striking example, the average threshold for high intensity for France when using a log transformation increase from %$q_{0.9} = 834, q_{0.975} = 1158$ for $m = 5, n = 6$ to $q_{0.9} = 1011, q_{0.975} = 1363$ for $m = 10, n = 3$ and $q_{0.9} = 1080, q_{0.975} = 1458$ for $m = 15, n = 2$.
+For $n = 1$, in which case the estimators can be interpreted as unbiased (Section \ref{sec:analytical}), we obtain averages of $q_{0.9} = 1147, q_{0.975} = 1559$.
+834 for $m = 5, n = 6$ to 1011 for $m = 10, n = 3$ and 1080 for $m = 15, n = 2$.
+For $n = 1$, in which case the estimators can be interpreted as unbiased (Section \ref{sec:analytical}), we obtain an average of 1147.
+This corresponds to an increase of roughly 35\% relative to the thresholds obtained with $n = 6$.
+Including historical observations which are not from actual peak weeks thus leads to a considerable lowering of alarm thresholds and will necessarily increase the number of alerts for high and very high influenza activity.
+As an example, in Spain, applying the MEM with log transformation and $M = 5, n = 6$ leads to 25\% of all seasons exceeding the high threshold and 11\% exceeding the very high threshold.
+This is obviously rather far from the intended excedance probabilities of 10\% and 2.5\% for the two thresholds.
 
 \begin{figure}
-\includegraphics[page=1, width=0.9\textwidth]{figure/plot_results.pdf}
+\centering
+\includegraphics[page=1, width=0.95\textwidth]{figure/plot_results.pdf}
+\caption{Results of bootstrap simulation study for France and Spain.
+First and third column: mean intensity thresholds along with bands delimited by the empirical 5\% and 95\% quantiles.
+The approximate mean threshold values obtained analytically are marked as little black dots.
+Second and fourth columns: resulting average shares of seasons classified as low, medium, high and very high intensity (shown as in a stacked barplot).}
 \label{fig:results1}
@@
 \begin{figure}
-\includegraphics[page=2, width=0.9\textwidth]{figure/plot_results.pdf}
-\caption{Results of bootstrap simulation study for Spain and the United States.
+\centering
+\includegraphics[page=2, width=0.95\textwidth]{figure/plot_results.pdf}
+\caption{Results of bootstrap simulation study for Switzerland and the United States.
 First and third column: mean intensity thresholds along with bands delimited by the empirical 5\% and 95\% quantiles.
-SEcond and fourth columns: resulting average shares of seasons classified as low, medium, high and very high intensity.}
+The approximate mean threshold values obtained analytically are marked as little black dots.
+Second and fourth columns: resulting average shares of seasons classified as low, medium, high and very high intensity (shown as in a stacked barplot).}
 \label{fig:results2}
 \end{figure}
+
+The increase of thresholds as more historical data accumulates is not only a statistical tendency, but occurs with very high probability.
+Indeed, for all four countries, the probability that the threshold for high intensity increases when using the first ten rather than only the first five years of a sequence was around 90\%.
+To illutrate this aspect further, we examine the following toy example: We repeat a sequence of five seasons of ILI incidence data from France (2014--2019) four times to obtain a time series of twenty seasons.
+We then apply the moving epidemic method with log transformation and $n = 30/m$ using the first $m = 5, 6, \dots, 20$ years of data, thus mimicking the development of thresholds as more historical data becomes available.
+As can be seen from Figure \ref{fig:example_france}, this results in a quite steady increase of thresholds, despite the fact that the data in each five-year block behave exactly the same.
+For example, season 6 is classified as very high intensity based on seasons 1 through 5, but seasons 11 and 16 are not, even though they are identical to season 6 and the respective data used to construct thresholds follow exactly the same patterns.
+A similar discrepancy can be observed for season 7, which is classified as medium intensity, while seasons 12 and 17 are marked as low intensity.
+
+\begin{figure}
+\center
+\includegraphics[scale=0.6]{figure/example_france.pdf}
+\caption{Toy example to illustrate increasing thresholds as historical data accumulates: Data from France, 2014/2015--2018/2019 are appended three times and thresholds are computed using $m = 5, \dots, 15$ years of training data.
+The thresholds based on years 1 through $m$ are overlaid with the incidence time series from the $(m + 1)$-th year.}
+\label{fig:example_france}
+\end{figure}
+
+When just using $n = 1$ irrespective of the number of historical seasons available, the average thresholds and shares of the different categories are more well-behaved also for small $m$.
+Of course, there is considerable variability in the estimated thresholds.
+This difficulty, however, is inherent in the problem of estimating a 90\% or 97.5\% quantile based on only 5--10 observations.
 
@@
 
-Mention that also concerns computation of seasonal thresholds.
+We provided a statistical assessment of different implementation choices in a simple and widely used framework for the computation of influenza intensity thresholds.
+We found that using a log transformation to intensity measures prior to the computation of thresholds empirically led to better behaved thresholds with cloesr to nominal exceedance rates.
+Concerning the question of how many observations per historical season should be included in the computation of thresholds, we provided analytical arguments and empirical evidence that the common choice $n = 30/m$ leads to too low thresholds when few historical seasons are available.
+We therefore recommend to adopt $n = 1$ irrespective of the number $M$ of available historical seasons.
+This is possible using the available implementation in the \texttt{R} package \texttt{mem} and only requires setting \texttt{i.n.max = 1} in the call of the \texttt{memmodel} to override the default value 30/m.
 
-\section*{Outline of idea}
+We focused on the computation of influenza intensity thresholds, but the same general arguments apply to the computation of thresholds to determine the onset of a season \citep{Vega2012}.
+As variability among observations around the season onset may be lower than around the peak the problem may be less severe; we did not assess this empirically as not all our data sets fully covered the off-season.
 
-\begin{itemize}
-\item Description of importance: officially embraced by ECDC
-\item Describe other approaches, in particular WHO
-\item Statistically correct description of what is being done, contract to WHO approach (which only uses peak -- good -- but uses normal rather than log-normal -- not so good.
-There is a PLOS paper where there is already a kind of combination of the two (WHO with log transformation))
-\item Overview of how it is being used in the literature:
-\begin{itemize}
-\item How much training data was available in each study?
-\end{itemize}
-\item Simulation study based on bootstrapping of true seasons:
-\begin{itemize}
-\item Countries: France, Switzerland, some third (European) country? Spain?
-\item Important: standardize with time-varying population
-\end{itemize}
-\end{itemize}
-
-\title{A statistical perspective on the moving epidemic method }
-\author{Johannes Bracher}
-\maketitle
+More sophisticated methods to determine influenza intensity thresholds could be imagined, e.g.\ based on the broad literature on outbreak and anomaly detection \citep{Unkel2012}.
+However, we think that in practice a simple and interpretable tool with a robust software implementation like the MEM is a good choice.
+The emergence of a standard approach at the European level and beyond will contribute to improved comparability of results and facilitate the evaluation and refinement of the applied tool.
+With this work we hope to contribute a statistical perspective on this topic, which can complement public health practitioners' experience from applied analyses.
 

### 2021-01-27 Simulations/data_preparation.R
Updating codes + results.

@@ -1,6 +1,7 @@
 # Preparing the different data sets for use of mem
 
-setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+# setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+library(here)
 
 # France:
 

### 2021-01-27 Simulations/functions_simulations_mem.R

@@ -1,3 +1,4 @@
+# function to summarize simulation results
 sim_summary <- function(results_0.4, results_0.9, results_0.975, range_i.seasons,
                         interval = c(0.025, 0.975)){
   if(length(interval) != 2) stop("interval needs to be of length 2.")
@@ -32,6 +33,7 @@ modify_alpha <- function(col, alpha){
   rgb(x[1], x[2], x[3], alpha = alpha)
 }
 
+# plotting function for thresholds
 plot_sim_summary <- function(sim_summary, ylim = NULL, show_bands = TRUE, ...){
   ind_lower <- names(sim_summary)[grepl("q_", names(sim_summary))][1]
   ind_upper <- names(sim_summary)[grepl("q_", names(sim_summary))][2]
@@ -68,6 +70,7 @@ plot_sim_summary <- function(sim_summary, ylim = NULL, show_bands = TRUE, ...){
   }
 }
 
+# plotting function for exceedance proportions
 plot_exceedane_summary <- function(summary_exceedance, ...){
   summary_exceedance$averages
   range_i.seasons <- as.numeric(gsub("i.seasons_", "", rownames(summary_exceedance$averages), 1))
@@ -85,4 +88,21 @@ plot_exceedane_summary <- function(summary_exceedance, ...){
           col = "chartreuse4", border = NA)
 
   abline(h = c(0.4, 0.9, 0.975), lty = 3)
+}
+
+legend_summary <- function(){
+  plot(NULL, xlim = 0:1, ylim = 0:1, axes = FALSE, xlab = "", ylab = "")
+  legend("top", legend = c("Intensity levels:", "very high", "high", "medium", "low", 
+                           "First and third column:", "mean threshold", "empirical 5% and 95% quantiles", "analytical approximation of", "mean threshold",
+                           "Second and fourth column:",  "intended shares of intensity levels", "empirical shares", ""), 
+         col = c(NA, "red", "darkorange", "darkgoldenrod1", "chartreuse4",
+                 NA, "black", "grey", "black", NA,
+                 NA, "black", "red"),
+         pch = c(NA, 15, 15, 15, 15, 
+                 NA, 15, 15, 1, NA,
+                 NA, NA, 15, NA), 
+         lty = c(NA, NA, NA, NA, NA,
+                 NA, 2, NA, NA, NA,
+                 NA, 3, NA, NA),
+         ncol = 3, bty = "n")
 }

### 2021-01-27 Simulations/plot_timeseries.R

@@ -11,9 +11,16 @@ plot(dat_us[, 5])
 
 matr <- dat_ch
 xlim <- c(1990, 2020)
+
 plot_mem_ts <- function(matr, xlim, ylab = "inc / 100,000", ...){
   years <- as.numeric(substr(colnames(matr), start = 13, stop = 16)) - 1
-  plot(NULL, xlim = xlim, ylim = c(0, 1.2*max(matr)), xlab = "", ylab = ylab, ...)
+  labels_years <- gsub(".", "/", substr(colnames(matr), start = 8, stop = 16), fixed = TRUE)
+  
+  plot(NULL, xlim = xlim, ylim = c(0, 1.2*max(matr)), xlab = "", ylab = ylab, axes = FALSE,...)
+  axis(1, at = years + 0.5, labels = labels_years)
+  axis(2)
+  box()
+  
   abline(v = seq(from = floor(xlim[1]), to = ceiling(xlim[2])), col = "grey")
 
   weeks_per_seas <- nrow(matr)

### 2021-01-27 Simulations/summarize_simulations.R

@@ -1,6 +1,6 @@
 # summarize simulation results
 
-setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+library(here) # set working directory to current
 source("functions_simulations_mem.R")
 
 # the range of number of seasons:
@@ -27,9 +27,9 @@ summary_thresholds1_ch <- sim_summary(thresholds1_ch_0.4, thresholds1_ch_0.9, th
                                       range_i.seasons, interval = interv)
 summary_exceedance1_ch <- sim_summary(exceedance1_ch_0.4, exceedance1_ch_0.9, exceedance1_ch_0.975,
                                       range_i.seasons, interval = interv)
-summary_thresholds_nolog_ch <- sim_summary(thresholds_nolog_ch_0.4, thresholds_nolog_ch_0.9, thresholds_ch_0.975,
+summary_thresholds_nolog_ch <- sim_summary(thresholds_nolog_ch_0.4, thresholds_nolog_ch_0.9, thresholds_nolog_ch_0.975,
                                            range_i.seasons, interval = interv)
-summary_exceedance_nolog_ch <- sim_summary(exceedance_nolog_ch_0.4, exceedance_nolog_ch_0.9, exceedance_ch_0.975,
+summary_exceedance_nolog_ch <- sim_summary(exceedance_nolog_ch_0.4, exceedance_nolog_ch_0.9, exceedance_nolog_ch_0.975,
                                            range_i.seasons, interval = interv)
 summary_thresholds1_nolog_ch <- sim_summary(thresholds1_nolog_ch_0.4, thresholds1_nolog_ch_0.9, thresholds1_nolog_ch_0.975,
                                             range_i.seasons, interval = interv)
@@ -45,9 +45,9 @@ summary_thresholds1_fr <- sim_summary(thresholds1_fr_0.4, thresholds1_fr_0.9, th
                                       range_i.seasons, interval = interv)
 summary_exceedance1_fr <- sim_summary(exceedance1_fr_0.4, exceedance1_fr_0.9, exceedance1_fr_0.975,
                                       range_i.seasons, interval = interv)
-summary_thresholds_nolog_fr <- sim_summary(thresholds_nolog_fr_0.4, thresholds_nolog_fr_0.9, thresholds_fr_0.975,
[…]

### 2021-01-27 Draft/Draft_MEM.tex
Integrating Jonas' corrections.

@@
 The simulation study is based on a bootstrap approach using historical ILI data from France, Spain, Switzerland and the United States.
-We find that using data on the original scale leads to a too large proportion of seasons being classified as high or ``very high intensity intensity, which can to some degree be mitigated by using a logarithmic transformation.
+We find that using data on the original scale leads to a too large proportion of seasons being classified as high or very high intensity intensity, which can to some degree be mitigated by using a logarithmic transformation.
 When fixing the total number of observations to include as in the MEM, intensity thresholds tend to increase as more historical data becomes available.
@@
 \item Within each historical season $j = 1, \dots, m$ order all observations in decreasing order, denoting the $i$-th largest observation from season $j$ by $X^{(j)}_i$.
-\item Select the $n$ largest observations from each of the $m$ past seasons to construct a reference set $\mathcal{X} = \{X_i^{(j)}; j = 1, \dots, m; i = 1, \dots, n\}$.
+\item Select the $n$ largest observations from each of the $m$ past seasons to construct a reference set $\mathcal{X} = \{X_i^{(j)}: j = 1, \dots, m; i = 1, \dots, n\}$.
 \item Apply an (invertible) transformation $Y_i^{(j)} = f(X_i^{(j)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
@@
 \item In the WHO method, the standard procedure is to set $f$ to the identity function (i.e.\ no transformation is applied).
-If peaks vary considerably, a log-tranformation is recommended \citep{WHO2017}.
+If peaks incidence vary strongly across seasons, a log-tranformation is recommended \citep{WHO2017}.
 Irrespective of the number $m$ of historical seasons, $n = 1$ observation per season is used.
@@
 Scotland & influenza & 2010--2019 & 7--8 & 4 & 40, 90, 97.5 & \cite{Dickson2020}\\
-Slovenia & RSV & 2008--2018 & 10 & 3 & 40, 90, 97.5 & \cite{Grilc2020}\\
+Slovenia & RSV & 2008--2018 & 10 & 3 & 40, 90, 97.5 & \cite{Grilc2021}\\
 Spain (17 regions) & ILI & 2003--2015 & 4--10 & 3--8 & 40, 90, 97.5 & \cite{Bangert2017}\\
@@
 \left(\begin{array}{ccc}
-\sigma_{1, 1} & \cdots & \sigma_{1n}\\
+\sigma_{1, 1} & \cdots & \sigma_{1, n}\\
 \vdots & \ddots &\vdots\\
-\sigma_{n, 1} & \cdots & \sigma_{nn}
+\sigma_{n, 1} & \cdots & \sigma_{n, n}
 \end{array}\right),
 \end{align}
-respectively, where to make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$. It can then be shown that the expectations of $\hat\mu$ and $\hat{\sigma}^2$ are given by
+respectively, where to make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$.
+It can then be shown that the expectations of $\hat\mu$ and $\hat{\sigma}^2$ when using $m$ seasons and $n$ observations from each season are given by
 \begin{align}
@@
 We provided a statistical assessment of different implementation choices in a simple and widely used framework for the computation of influenza intensity thresholds.
-We found that using a log transformation to intensity measures prior to the computation of thresholds empirically led to better behaved thresholds with cloesr to nominal exceedance rates.
+We found that using a log transformation to intensity measures prior to the computation of thresholds empirically led to better behaved thresholds with closer to nominal exceedance rates.
 Concerning the question of how many observations per historical season should be included in the computation of thresholds, we provided analytical arguments and empirical evidence that the common choice $n = 30/m$ leads to too low thresholds when few historical seasons are available.
@@
 
+\section*{Data and code}
+
+Materials to reproduce the presented results are available at \url{https://github.com/jbracher/mem}.
+
 \bibliographystyle{apalike}

### 2021-01-28 Draft/Derivations_MEM.tex
Corrections after discussion with Jonas.

\documentclass{article}
\usepackage[left=2.5cm,right=2.5cm, top = 2.5cm, bottom = 3cm]{geometry}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{amsmath}

\newcommand{\sd}{s}
\newcommand{\mean}{\bar{Y}}

\begin{document}
% \SweaveOpts{concordance=TRUE}

\title{A bootstrap-based comparison of ILI intensity thresholds from the moving epidemic and WHO methods: Derivations}
\author{Johannes Bracher, Jonas Littek}
\date{ \small
Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics}

\maketitle

Denote by $\mathbf{Y}^{(j)}$ the vector of the $n$ largest transformed observations (after transformation with $f$) from season $j$ in decreasing order, i.e.
$$
\mathbf{Y}^{(j)} = \left(
\begin{array}{c}
Y^{(j)}_1\\
\vdots\\
Y^{(j)}_n
\end{array}
\right)
$$
Now assume that $\mathbf{Y}^{(1)}, \dots, \mathbf{Y}^{(m)}$ are independent and identically distributed with mean and covariance matrix
\begin{align}
\mathbb{E}(\mathbf{Y}^{(j)}) = \boldsymbol{\mu} = \left(\begin{array}{c}
\mu_1\\
\vdots\\
\mu_n
\end{array}\right) \ \ \ \text{and} \ \ \ \text{Cov}(\mathbf{Y}^{(j)}) = \boldsymbol{\Sigma} =
\left(\begin{array}{ccc}
\sigma_{1, 1} & \cdots & \sigma_{1, n}\\
\vdots & \ddots &\vdots\\
\sigma_{n, 1} & \cdots & \sigma_{n, n}
\end{array}\right),
\end{align}
respectively, where to make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$. We are interested in the expectations of the random variables $\mean$ and $\sd$, where
\begin{align*}
\mean & = \frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n Y_i^{(j)}\\
\sd & = \sqrt{\frac{1}{mn - 1} \sum_{j = 1}^m \sum_{i = 1}^n \left(Y_i^{(j)} - \mean\right)^2}.
\end{align*}
It is straightforward to see that 
\begin{equation}
\mathbb{E}(\mean) = \frac{1}{n} \sum_{i = 1}^n \mu_i.
\label{eq:expectation_mu}
\end{equation}
Instead of the empirical standard deviation $\sd$ we first consider the variance $\sd^2$, which can also be written as
\begin{align}
\sd^2 & = \frac{mn}{mn - 1} \left\{\frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n \left(Y_i^{(j)} - \mean\right)^2 \right\}\\
& = \frac{mn}{mn - 1} \left\{ \underbrace{\frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n Y_i^{(j)2}}_{= a} \ \ \ - \ \ \ \underbrace{\left(\frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n Y_i^{(j)} \right)^2}_{= b} \right\} \label{eq:sigma2hat}
\end{align}
%$$
%\sd^2 = \underbrace{\left(\frac{1}{nK - 1} \sum_{j = 1}^n \sum_{i = 1}^K Y_i^{(j)2}\right)}_{ = a} \ \ \ - \ \ \ \underbrace{\left(\frac{1}{nK - 1} \sum_{j = 1}^n \sum_{i = 1}^K Y_i^{(j)} \right)^2}_{= b}.
%$$
We consider the two terms $a$ and $b$ separately, starting by
\begin{align*}
\mathbb{E}(a) & = \frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n \mathbb{E}\left(Y_i^{(j)2}\right)\\
& = \frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n \left\{ \text{Var}\left(Y_i^{(j)}\right) + \mathbb{E}\left(Y_i^{(j)}\right)^2 \right\}\\
& = \frac{m}{mn} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2).
\end{align*}
Then we note that
\begin{align*}
\mathbb{E}(b) & = \text{Var}\left( \frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n Y_i^{(j)} \right) \ \ \ + \ \ \ \mathbb{E}\left(\frac{1}{mn}  \sum_{j = 1}^m \sum_{i = 1}^n Y_i^{(j)} \right)^2\\
& = \frac{1}{(mn)^2}\sum_{j = 1}^m \text{Var}\left(\sum_{i = 1}^n Y_i^{(j)} \right) \ \ \ + \ \ \ \left(\frac{m}{mn} \sum_{i = 1}^n \mu_i\right)^2\\
& = \frac{m}{(mn)^2} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ \ \ + \ \ \ \frac{m^2}{(mn)^2}\left(\sum_{i = 1}^n \mu_i\right)^2.
\end{align*}
Plugging this back into equation \eqref{eq:sigma2hat} we obtain
\begin{equation}
\mathbb{E}(\sd^2) = \frac{m}{mn - 1} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2) \ \ \ - \ \ \ \frac{1}{n(mn - 1)} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ \ \ - \ \ \ \frac{m}{n(mn - 1)}\left(\sum_{i = 1}^K \mu_i\right)^2.
\label{eq:expectation_sigma2}
\end{equation}
It is straightforward to see that for $n = 1$ the expressions \eqref{eq:expectation_mu} and \eqref{eq:expectation_sigma2} simplify to
$$
\mathbb{E}(\mean^2) = \mu_1 \ \ \text{ and } \ \ \mathbb{E}(\sd^2) = \sigma^2_1.
$$

There is no general way of computing the expectation $\mathbb{E}(\sd)$, but unless the true distribution of  $\sd$ has strong excess curtosis,
\begin{equation}
\mathbb{E}(\sd) \approx \sqrt{\mathbb{E}(\sd^2)}
\label{eq:expectation_sigma}
\end{equation}
is a reasonable approximation. We can then plug equations \eqref{eq:expectation_mu} and \eqref{eq:expectation_sigma} into the formulae for the thresholds $q_{Y, \alpha}$ (with $\alpha \in \{0.4, 0.9, 0.975\}$) on the transformed scale and obtain
$$
\mathbb{E}(q_{Y, \alpha}) \approx \mathbb{E}(\mean) + z_\alpha \sqrt{\mathbb{E}(\sd^2)},
$$
where $z_\alpha$ is the $\alpha$ quantile of the standard normal distribution.

If $f$ was set to the natural logarithm, the question remains how to obtain statements concerning thresholds $q_\alpha$ on the original scale. Approximation via a second-order Taylor expansion yields
\begin{equation}
\mathbb{E}(q_\alpha) = \mathbb{E}\left\{\exp(q_{Y, \alpha})\right\} \approx \exp\left\{\mathbb{E}(q_{Y, \alpha})\right\} \times \left\{1 + \frac{\text{Var}(q_{Y, \alpha})}{2} \right\}.
\label{eq:taylor}
\end{equation}

Empirically, after transformation to the log scale, the variance of season peak values is low (between 0.02 and 0.4 for the adta sets analysed in Section \textbf{XXX} of the main manuscript). The resulting variances of $q_{Y, \alpha}$ are likewise quite small and do not play an importnat role in equation \eqref{eq:taylor}. We can thus use the even simpler approximation
\begin{equation}
\mathbb{E}(q_\alpha) \approx \exp\left\{\mathbb{E}(q_{Y, \alpha})\right\}.
\end{equation}

\end{document}

### 2021-01-28 Draft/Draft_MEM.tex
Corrections after discussion with Jonas.

@@
 \usepackage{amsmath}
+
+\newcommand{\sd}{s}
+\newcommand{\mean}{\bar{Y}}
 
@@
 \item Apply an (invertible) transformation $Y_i^{(j)} = f(X_i^{(j)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
-\item Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates $\hat\mu, \hat \sigma$ of its mean and standard deviation.
-\item Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\hat\mu, \hat\sigma^2)$, i.e.\ compute
-$$
-q_{Y, \alpha} = \hat{\mu} + z_\alpha \hat{\sigma},
-$$
+\item Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates $\mean, \sd$ of its mean and standard deviation.
+\item Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\mean, \sd^2)$, i.e.\ compute
+\begin{equation}
+q_{Y, \alpha} = \mean + z_\alpha \sd, \label{eq:def_q}
+\end{equation}
 where $z_\alpha$ is the $\alpha$ quantile of the standard normal distribution.
@@
 \begin{itemize}
-\item the 40th percentile $q_{Y, 0.4} = \hat\mu - 0.25 \hat\sigma$ as the threshold between low and medium intensity;
-\item the 90th percentile $q_{Y, 0.9} = \hat\mu + 1.28 \hat\sigma$ as the threshold between medium and high intensity;
-\item the 97.5th percentile $q_{Y, 0.975} = \hat\mu + 1.96\hat\sigma$ as the threshold between high and very high intensity.
+\item the 40th percentile $q_{Y, 0.4} = \mean - 0.25 \sd$ as the threshold between low and medium intensity;
+\item the 90th percentile $q_{Y, 0.9} = \mean + 1.28 \sd$ as the threshold between medium and high intensity;
+\item the 97.5th percentile $q_{Y, 0.975} = \mean + 1.96\sd$ as the threshold between high and very high intensity.
 \end{itemize}
@@
 Note that in most previous descriptions (e.g., \ \citealt{WHO2014, Vega2015}), the thresholds have been referred to as upper ends of one-sided confidence intervals associated with the the arithmetic (WHO method) or geometric mean (MEM) of the historical reference observations in $\mathcal{X}$.
-This, however, is slightly imprecise terminology as in the computations the standard deviation $\hat\sigma$ rather than the standard error $\hat\sigma/\sqrt{nm}$ of $\hat\mu$ is used.
+This, however, is slightly imprecise terminology as in the computations the standard deviation $\sd$ rather than the standard error $\sd/\sqrt{nm}$ of $\mean$ is used.
 If indeed confidence intervals were used, these would tend to get narrower the more historical seasons were available, and the thresholds for all levels would eventually converge to the arithmetic / geometric mean.
@@
 Little more can be said than that a logarithmic transformation can be helpful if the distribution of peak values is skewed.
-However, some statements can be made on how the value chosen for $n$ impacts the expected results.
+However, some statements can be made on how the value chosen for $n$ impacts the expected results for the thresholds $q_\alpha$.
 
+We start by considering how the expectations of the empirical mean $\mean$ and standard deviation $\sd$ entering into equation \eqref{eq:def_q} depend on $n$.
 Denote by $\mathbf{Y}^{(j)}$ the vector of the $n$ largest transformed observations from season $j$ in decreasing order, i.e.
@@
 respectively, where to make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$.
-It can then be shown that the expectations of $\hat\mu$ and $\hat{\sigma}^2$ when using $m$ seasons and $n$ observations from each season are given by
+It can then be shown that the expectations of $\mean$ and $\sd^2$ when using $m$ seasons and $n$ observations from each season are given by
 \begin{align}
-\mathbb{E}(\hat{\mu}) & = \frac{1}{n} \sum_{i = 1}^n \mu_i,
+\mathbb{E}(\mean) & = \frac{1}{n} \sum_{i = 1}^n \mu_i,
 \label{eq:expectation_mu}\\
-\mathbb{E}(\hat{\sigma}^2) & = \frac{m}{mn - 1} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{n(mn - 1)} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ - \ \frac{m}{n(mn - 1)}\left(\sum_{i = 1}^n \mu_i\right)^2,
+\mathbb{E}(\sd^2) & = \frac{m}{mn - 1} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{n(mn - 1)} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ - \ \frac{m}{n(mn - 1)}\left(\sum_{i = 1}^n \mu_i\right)^2,
 \label{eq:expectation_sigma2}
@@
 %\begin{align}
-%\mathbb{E}(\hat{\mu}) & = \frac{1}{K} \sum_{i = 1}^K \mu_i,
+%\mathbb{E}(\mean) & = \frac{1}{K} \sum_{i = 1}^K \mu_i,
 %\label{eq:expectation_mu}\\
-%\mathbb{E}(\hat{\sigma}^2) & = \frac{n}{nK - 1} \sum_{i = 1}^K (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{K(nK - 1)} \sum_{i = 1}^K \sum_{i' = 1}^K \sigma_{i,i'} \ - \ \frac{n}{K(nK - 1)}\left(\sum_{i = 1}^K \mu_i\right)^2,
+%\mathbb{E}(\sd^2) & = \frac{n}{nK - 1} \sum_{i = 1}^K (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{K(nK - 1)} \sum_{i = 1}^K \sum_{i' = 1}^K \sigma_{i,i'} \ - \ \frac{n}{K(nK - 1)}\left(\sum_{i = 1}^K \mu_i\right)^2,
 %\label{eq:expectation_sigma2}
@@
 \begin{equation}
-\mathbb{E}(q_\alpha) \approx f^{-1}\left\{\mathbb{E}(\hat\mu) + z_\alpha \sqrt{\mathbb{E}(\hat\sigma^2)}\right\}
+\mathbb{E}(q_\alpha) \approx f^{-1}\left\{\mathbb{E}(\mean) + z_\alpha \sqrt{\mathbb{E}(\sd^2)}\right\}
 \label{eq:expectation_q}
 \end{equation}
-usually holds in good approximation (with $f^{-1}$ the idenity or exponential function). This implies that for $n = 1$, we have
+usually holds in good approximation in our applied setting (with $f^{-1}$ the idenity or exponential function). This implies that for $n = 1$, we have
 $$
@@
 Equations \eqref{eq:expectation_mu}--\eqref{eq:expectation_q} tell us by how much $q_\alpha$ can be expected to differ from the intended value $f^{-1}(\mu_1 + z_\alpha \sigma_1)$.
-By the definition of the $\mu_i$ (with $\mu_i$ the expectation of the $i$-th largest observation in a given season), $\mathbb{E}(\hat\mu)$ decreases in $n$.
+By the definition of the $\mu_i$ (with $\mu_i$ the expectation of the $i$-th largest observation in a given season), $\mathbb{E}(\mean)$ decreases in $n$.
 In most (possibly all) settings this also translates to the $q_\alpha$.
@@
 In all four countries, the thresholds obtained when using a log transformation as in the MEM are considerably higher than when no transformation is used.
-Especially in France and Spain, thresholds obtained without the log transormation are clearly too low, as can be seen from the large proportion of seasons classified as very high intensity (around 10\% when using $n = 1$).
+Especially in France and Spain, thresholds obtained without the log transformation are clearly too low, as can be seen from the large proportion of seasons classified as very high intensity (around 10\% when using $n = 1$).
 Applying the log transformation mitigates this to some degree.

### 2021-01-29 Draft/Draft_MEM.tex
Using correct terminology for wILI.

@@
 Population data were obtained from the Swiss Federal Statistical Office.
-Weekly weighted ILI percentage rates (wILI) in the US from CDC FluView \citep{Charbonneau2019}, 1998--2017, were obtained from the CDC FluSight influenza forecasting platform (\url{https://github.com/FluSightNetwork/cdc-flusight-ensemble/}).
-These wILI values correspond to the percentage of general practitioner visits due to influenza-like symptoms
+Weekly weighted ILI (wILI) data at the US national level from CDC FluView \citep{Charbonneau2019}, 1998--2017, were obtained via the CDC FluSight influenza forecasting platform (\url{https://github.com/FluSightNetwork/cdc-flusight-ensemble/}).
+These wILI values correspond to the fraction of general practitioner visits due to influenza-like symptoms
 
@@
 \includegraphics[width=0.9\textwidth]{figure/plot_data.pdf}
-\caption{Time series of influenza activity measures in four countries: Weekly ILI cases per 100,000 population in France, 1985--2019; weekly number of confirmed influenza cases per 100,000 population in Spain, 1998--2019; weekly ILI cases per 100,000 in Switzerland, 2000--2016; weekly wILI percentages in the United States, 1998--2017.
+\caption{Time series of influenza activity measures in four countries: Weekly ILI cases per 100,000 population in France, 1985--2019; weekly number of confirmed influenza cases per 100,000 population in Spain, 1998--2019; weekly ILI cases per 100,000 in Switzerland, 2000--2016; weekly wILI values in the United States, 1998--2017.
 Off-season weeks are omitted in the plot, with grey lines delimiting the different seasons.}

### 2021-01-29 Simulations/plot_timeseries.R

@@ -1,4 +1,4 @@
-setwd("/home/johannes/Documents/Ideas/mem/Simulations")
+library(here)
 
 dat_ch <- read.csv("../Data/for_mem/ili_mem_ch.csv")
 dat_es <- read.csv("../Data/for_mem/ili_mem_es.csv")
@@ -49,7 +49,7 @@ bplot_ranks(dat_es)
 plot_mem_ts(dat_ch, xlim = c(2000.5, 2015.5)); legend("topleft", legend = " Switzerland", bty = "n", cex = 1.2)
 bplot_ranks(dat_ch)
 
-plot_mem_ts(dat_us, xlim = c(1998.5, 2017.5), ylab = "wILI %"); legend("topleft", legend = " United States", bty = "n", cex = 1.2)
+plot_mem_ts(dat_us, xlim = c(1998.5, 2017.5), ylab = "wILI"); legend("topleft", legend = " United States", bty = "n", cex = 1.2)
 bplot_ranks(dat_us)
 dev.off()
 

### 2021-02-10 Simulations/simulations_mem_ci_fr.R
Updating simulation codes.

@@ -1,0 +1,280 @@
+############################################################
+# Running permutation/simulation study for Switzerland
+
+setwd("/home/johannes/Documents/Projects/mem/Simulations")
+
+library(mem)
+source("functions_simulations_mem.R")
+
+dat_for_mem <- read.csv("../Data/for_mem/ili_mem_fr.csv")
+
+# compute peaks of seasons:
+peaks_test <- apply(dat_for_mem, 2, max)
+
+
+# simulation settings:
+range_i.seasons <- 5:15 # range to be explored for i.seasons
+n_sim <- 500
+
+
+
+#####################################################
+### Run simulation using default settings of mem
+#####################################################
+
+# initialize matrices to store results (thresholds and exceedance proportions)
+thresholds_ci_fr_0.4 <- matrix(NA, nrow = n_sim, ncol = length(range_i.seasons))
+colnames(thresholds_ci_fr_0.4) <- paste0("i.seasons_", range_i.seasons)
+thresholds_ci_fr_0.9 <- thresholds_ci_fr_0.975 <-
+  exceedance_ci_fr_0.4 <- exceedance_ci_fr_0.9 <- exceedance_ci_fr_0.975 <-
+  thresholds_ci_fr_0.4
+
+# run simulation:
+for(k in 1:n_sim){
+  set.seed(k)
+  
+  # generate bootstrapped time series:
+  inds_train <- sample(1:ncol(dat_for_mem), size = max(range_i.seasons), replace = TRUE)
+  dat_train <- dat_for_mem[, inds_train]
+  
+  # run through different numbers of past seasons used:
+  for(a in seq_along(range_i.seasons)){
+    
+    # restrict to first a seasons:
+    dat_train_temp <- dat_train[, 1:range_i.seasons[a]]
+    
+    # run memmodel:
+    memmodel_temp <- memmodel(dat_train_temp, i.seasons = range_i.seasons[a],
+                              i.level.intensity = c(0.4, 0.9, 0.975),
+                              i.type.intensity = 2)
+    
+    # store thresholds:
+    thresholds_ci_fr_0.4[k, a] <- memmodel_temp$intensity.thresholds[1]
[…]

### 2021-02-10 Simulations/summarize_simulations.R

@@ -240,3 +240,51 @@ legend_summary() # legend defined in separate function
 
 dev.off()
 
+# Additional plot for France using CIs:
+
+# load additional French results using confidence intervals:
+load(paste0("Results/results_ci_fr.rda"))
+load(paste0("Results/results1_ci_fr.rda"))
+load(paste0("Results/results_ci_nolog_fr.rda"))
+load(paste0("Results/results1_ci_nolog_fr.rda"))
+
+
+# France:
+summary_thresholds_ci_fr <- sim_summary(thresholds_ci_fr_0.4, thresholds_ci_fr_0.9, thresholds_ci_fr_0.975,
+                                     range_i.seasons, interval = interv)
+summary_exceedance_ci_fr <- sim_summary(exceedance_ci_fr_0.4, exceedance_ci_fr_0.9, exceedance_ci_fr_0.975,
+                                     range_i.seasons, interval = interv)
+summary_thresholds1_ci_fr <- sim_summary(thresholds1_ci_fr_0.4, thresholds1_ci_fr_0.9, thresholds1_ci_fr_0.975,
+                                      range_i.seasons, interval = interv)
+summary_exceedance1_ci_fr <- sim_summary(exceedance1_ci_fr_0.4, exceedance1_ci_fr_0.9, exceedance1_ci_fr_0.975,
+                                      range_i.seasons, interval = interv)
+summary_thresholds_ci_nolog_fr <- sim_summary(thresholds_ci_nolog_fr_0.4, thresholds_ci_nolog_fr_0.9, thresholds_ci_nolog_fr_0.975,
+                                           range_i.seasons, interval = interv)
+summary_exceedance_ci_nolog_fr <- sim_summary(exceedance_ci_nolog_fr_0.4, exceedance_ci_nolog_fr_0.9, exceedance_ci_nolog_fr_0.975,
+                                           range_i.seasons, interval = interv)
+summary_thresholds1_ci_nolog_fr <- sim_summary(thresholds1_ci_nolog_fr_0.4, thresholds1_ci_nolog_fr_0.9, thresholds1_ci_nolog_fr_0.975,
+                                            range_i.seasons, interval = interv)
+summary_exceedance1_ci_nolog_fr <- sim_summary(exceedance1_ci_nolog_fr_0.4, exceedance1_ci_nolog_fr_0.9, exceedance1_ci_nolog_fr_0.975,
[…]

### 2021-02-10 Draft/Derivations_MEM.tex
Updating drafts.

@@
 
-Empirically, after transformation to the log scale, the variance of season peak values is low (between 0.02 and 0.4 for the adta sets analysed in Section \textbf{XXX} of the main manuscript).
-The resulting variances of $q_{Y, \alpha}$ are likewise quite small and do not play an importnat role in equation \eqref{eq:taylor}.
+Empirically, after transformation to the log scale, the variance of season peak values is low (between 0.02 and 0.4 for the data sets analysed in Section 5 of the main manuscript).
+The resulting variances of $q_{Y, \alpha}$ are then quite small and do not play an important role in equation \eqref{eq:taylor}.
 We can thus use the even simpler approximation

### 2021-02-10 Draft/Draft_MEM.tex
Updating drafts.

@@
 \newcommand{\sd}{s}
-\newcommand{\mean}{\bar{Y}}
+\newcommand{\mean}{\bar{y}}
 
@@
 
-\title{A bootstrap-based comparison of ILI intensity thresholds from the moving epidemic and WHO methods}
-\author{Johannes Bracher, Jonas Littek}
+% \title{A bootstrap-based comparison of ILI intensity thresholds from the moving epidemic and WHO methods}
+\title{A statistical comparison of influenza intensity thresholds obtained from the moving epidemic and WHO methods}
+\author{Johannes Bracher$^{1, 2}$, Jonas Littek$^1$}
 \date{ \small
-Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics}
+$^1$Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics\\
+$^2$Heidelberg Institute for Theoretical Studies, Computational Statistics Group}
 
@@
 Secondly, the MEM method by default uses more than one observation from each past season, with a fixed total number of observations to include.
-The WHO method, on the other side, uses only the single highest value from each season to determine peak intensity thresholds.
+The WHO method, on the other hand, uses only the single highest value from each season to determine peak intensity thresholds.
 We provide statistical arguments and a simulation study to assess the impact of these choices on the resulting intensity thresholds and empirical exceedance proportions.
-The simulation study is based on a bootstrap approach using historical ILI data from France, Spain, Switzerland and the United States.
-We find that using data on the original scale leads to a too large proportion of seasons being classified as high or very high intensity intensity, which can to some degree be mitigated by using a logarithmic transformation.
-When fixing the total number of observations to include as in the MEM, intensity thresholds tend to increase as more historical data becomes available.
-When data is only available for few seasons, there is a rather large chance of classifying a new season as high or very high intenstity.
+The simulation study is based on a re-sampling approach using historical ILI data from France, Spain, Switzerland and the United States.
+We find that using data on the original scale leads to a rather large proportion of season peaks being classified as high or very high intensity, which can be mitigated to some degree by using a logarithmic transformation.
+When fixing the total number of observations to be included as in the MEM, intensity thresholds tend to increase as more historical data becomes available.
+When data are only available for few seasons, there is a high chance of classifying a new season as high or very high intenstity.
 We therefore advocate always using one observation per season with a log transformation, i.e.\ a hybrid between the default settings of the MEM and WHO methods.
@@
 
-Following the 2009 influenza H1N1 pandemic, the need for a rapid assessment tool for influenza intensity was recognized. In a report, the Review Committee on the Functioning of the
-International Health Regulations and on Pandemic Influenza (H1N1) recommended that member states apply and evaluate their methods for severity assessment every year, thus improving their pandemic preparedness \citep[p.118]{WHO2011}.
-In the subsequent WHO Pandemic Infuenza Severity Assessment (PISA) guideline (\citeyear{WHO2017}), two specific statistical methods are recommended to determine influenza intensity thresholds.
+Following the 2009 influenza H1N1 pandemic, the need for a rapid assessment tool for influenza intensity was recognized. In a report, the \textit{Review Committee on the Functioning of the
+International Health Regulations and on Pandemic Influenza (H1N1)} recommended that member states apply and evaluate their methods for severity assessment every year, thus improving their pandemic preparedness \citep[p.118]{WHO2011}.
+In the subsequent \textit{WHO Pandemic Infuenza Severity Assessment (PISA)} guideline \citep{WHO2017}, two specific statistical methods are recommended to determine influenza intensity thresholds.
 Firstly, the so-called \textit{WHO method}, already described in a previous document \citep{WHO2014} and in more detail in the PISA guide; and secondly, the \textit{moving epidemic method} (MEM; \citealt{Vega2013, Vega2015}) which is also recommended by the European Centre for Disease Prevention and Control (e.g., \citealt{ECDC2017}).
 
-In particular the MEM approach has in the meantime been adopted by numerous national public health agencies (see references in Section \ref{sec:recent_applications}) and seems to deliver practically useful thresholds.
+In particular the MEM approach has in the meantime been adopted by numerous national public health agencies (see references in Section \ref{sec:recent_applications}) and is considered to deliver practically useful thresholds.
 Its application has been facilitated by the availablility of a user-friendly and well-documented \texttt{R} package (\texttt{mem}; \citealt{Lozano2020}), which can even be used via a graphical user interface \citep{Lozano2018}.
@@
 We obtain some analytical results, complemented by simulation experiments.
-In order to ensure a close link to real influenza patterns, the latter are based on bootstrap re-sampling of historical surveillance data.
-This way, we provide an improved understanding and intuition for the consequences of different implementation choices.
+In order to closely mimic real influenza patterns, the latter are based on bootstrap re-sampling of historical surveillance data.
+In this way, we provide an improved understanding and intuition for the consequences of different implementation choices.
 
 The article is structured as follows.
-Section \ref{sec:definitions} provides concise definitions of the MEM and WHO methods, highlighting two important differences between these conceptually similar approaches.
+Section \ref{sec:definitions} provides concise definitions of the MEM and WHO methods, highlighting two differences between these conceptually similar approaches.
 Section \ref{sec:recent_applications} consists of an overview of published applications of the MEM and WHO methods.
@@
 Details on the construction of baseline epidemic thresholds from pre-season observations are omitted.
-We assume that thresholds are based on data from $m$ past seasons and applied to the ($m + 1$)-th season.
+We assume that thresholds are based on influenza intensity data (typically weekly incidences) from $m$ past seasons and applied to the ($m + 1$)-th season.
 \cite{Vega2015} recommend to use $5 \leq m \leq 10$ seasons for the MEM to ensure a recent data basis.
-Thresholds for for a given intensity measure $X$ are then obtained via the following steps:
+Thresholds for a given intensity measure $X$ are then obtained via the following steps.
+In the following we will call $X$ the incidence for simplicity, even if it can also be a different measure.
 
 \begin{enumerate}
-\item Within each historical season $j = 1, \dots, m$ order all observations in decreasing order, denoting the $i$-th largest observation from season $j$ by $X^{(j)}_i$.
-\item Select the $n$ largest observations from each of the $m$ past seasons to construct a reference set $\mathcal{X} = \{X_i^{(j)}: j = 1, \dots, m; i = 1, \dots, n\}$.
-\item Apply an (invertible) transformation $Y_i^{(j)} = f(X_i^{(j)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
+\item Within each historical season $j = 1, \dots, m$ order all observations in decreasing order, denoting the $i$-th largest observation from season $j$ by $x^{(j)}_i$.
+\item Select the $n$ largest observations from each of the $m$ past seasons to construct a reference set $\mathcal{X} = \{x_i^{(j)}: j = 1, \dots, m; i = 1, \dots, n\}$.
+\item Apply an (invertible) transformation $y_i^{(j)} = f(x_i^{(j)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
 \item Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates $\mean, \sd$ of its mean and standard deviation.
@@
 \begin{itemize}
-\item In the WHO method, the standard procedure is to set $f$ to the identity function (i.e.\ no transformation is applied).
-If peaks incidence vary strongly across seasons, a log-tranformation is recommended \citep{WHO2017}.
-Irrespective of the number $m$ of historical seasons, $n = 1$ observation per season is used.
+\item In the WHO method, $n = 1$ observation per season is used irrespective of the number $m$ of historical seasons.
+The standard procedure is to apply no transformation (i.e.\ $f$ is set to the identity function).
+If peak incidences vary strongly across seasons, a log-tranformation is recommended \citep{WHO2017}.
 \item In the MEM method, the default choice for $f$ is the natural logarithm, while the number of observations per season is set to $n = 30/m$.
-(The exact specification used in the \texttt{R} package \texttt{mem} is $n = \max(1, \lfloor 30/m \rceil)$, where $\lfloor \cdot \rceil$ denotes rounding to the nearest integer.) The total number of historical observations is thus kept (approximately) fixed, ensuring a reasonably large number of reference observations
+(The exact specification used in the \texttt{R} package \texttt{mem} is $n = \max(1, \lfloor 30/m \rceil)$, where $\lfloor \cdot \rceil$ denotes rounding to the nearest integer.) The total number of historical observations is thus kept approximately fixed, ensuring a reasonably large number of reference observations.
 \end{itemize}
-Note that the software implementation of the moving epidemic method in the \texttt{R} package \texttt{mem} is rather general and permits the user to choose $n$, $m$, and $f$ (between the identify and logarithmic function).
+The software implementation of the moving epidemic method in the \texttt{R} package \texttt{mem} is rather general and permits the user to choose $n$, $m$, and $f$ (the identify function coresponding to \texttt{i.type.intensity	= 5} and logarithmic function to \texttt{i.type.intensity = 6}).
 The term \textit{moving epidemic method} could thus also be used as an umbrella term for the general procedure described above.
-We here use it in a more narrow sense to refer to the specific procedure described in \cite{Vega2015}, \cite{WHO2017} and reflected in the default parameters of the \texttt{mem} package.
+We use it here in a more narrow sense to refer to the specific procedure described in \cite{Vega2015}, \cite{WHO2017} and reflected in the default parameters of the \texttt{mem} package.
 
@@
 Note that in most previous descriptions (e.g., \ \citealt{WHO2014, Vega2015}), the thresholds have been referred to as upper ends of one-sided confidence intervals associated with the the arithmetic (WHO method) or geometric mean (MEM) of the historical reference observations in $\mathcal{X}$.
-This, however, is slightly imprecise terminology as in the computations the standard deviation $\sd$ rather than the standard error $\sd/\sqrt{nm}$ of $\mean$ is used.
-If indeed confidence intervals were used, these would tend to get narrower the more historical seasons were available, and the thresholds for all levels would eventually converge to the arithmetic / geometric mean.
+This, however, is imprecise terminology as in the computations the standard deviation $\sd$ is used rather than the estimated standard error $\sd/\sqrt{nm}$ (see the default settings in the R package \texttt{mem} and the description of the WHO method in \citealt{WHO2014}, p.
+69).
+We note that the use of actual confidence intervals is in fact also possible using the \texttt{mem} package (\texttt{i.type.intensity = 1} for the geometric mean, \texttt{i.type.intensity = 2} for the arithmetic mean).
+However, under this choice, the thresholds for all levels eventually converge to the arithmetic or geometric mean if sufficient historical data are available (as $\sd/\sqrt{nm}$ gets smaller and smaller).
 This is obviously not the intended behaviour.
+To illustrate this aspect we also provide simulation studies on these two choices, but these have been deferred to Appendix \textbf{XXX}.
 
@@
 
-To get a better understanding of the different settings in which the MEM and WHO methods are applied in practice we performed a literature search of articles published in English language and citing the papers \cite{Vega2015}, \cite{WHO2014} and \cite{WHO2017} until December 2020 (identified via \textit{CrossRef} and \textit{Google Scholar}).
+To get a better understanding of the different settings in which the MEM and WHO methods are applied in practice we performed a literature search of articles published in English and citing the papers \cite{Vega2015}, \cite{WHO2014} and \cite{WHO2017} until December 2020 (identified via \textit{CrossRef} and \textit{Google Scholar}).
 The results are summarized in Table \ref{tab:literature}.
@@
 Indeed, the contributions come from numerous countries and in many cases have been co-authored by representatives of national or regional public health agencies.
-In most analyses, the suggested threshold levels at the 40th, 90th and 97.5th percentile are used.
+In most analyses, the suggested threshold levels at the 40th, 90th and 97.5th percentile as described in Section \ref{sec:definitions} are used.
 Variability with respect to the number $m$ of historical seasons included is considerable, with a range from three to 16 seasons.
@@
 
-As detailed in the previous section, the MEM and WHO methods have been applied extensively in practice and overall seem to produce thresholds which are considered useful by domain experts.
+As detailed in the previous section, the MEM and WHO methods have been applied extensively in practice and overall seem to produce thresholds that are considered useful by domain experts.
 A more statistical assessment, however has so far been missing.
-Which choice of transformation function $f$ is most reaonable is largely an empirical question.
+Determining which choice of transformation function $f$ is most reaonable is largely an empirical question.
 Little more can be said than that a logarithmic transformation can be helpful if the distribution of peak values is skewed.
@@
 
-We start by considering how the expectations of the empirical mean $\mean$ and standard deviation $\sd$ entering into equation \eqref{eq:def_q} depend on $n$.
-Denote by $\mathbf{Y}^{(j)}$ the vector of the $n$ largest transformed observations from season $j$ in decreasing order, i.e.
+We start by considering how the expectations of the empirical mean $\bar{Y}$ and standard deviation $S$ entering into equation \eqref{eq:def_q} depend on $n$ (note that we now capitalize these as we consider the random variables rather than concrete observed values).
+Denote by $\mathbf{Y}^{(j)}$ the vector of the $n$ largest transformed incidences from season $j$ in decreasing order, i.e.
 $\mathbf{Y}^{(j)} = (Y^{(j)}_1, \dots, Y^{(j)}_n)^\top$ with $Y^{(j)}_1 \geq Y^{(j)}_2 \geq \dots \geq Y^{(j)}_n$.
@@
 This is a desirable property.
-Loosely speaking, if in addition the $Y_{1}^{(j)}$ come from a normal distribution, the nominal exceedance probabilities for the different thresholds would on average be achieved.
+Loosely speaking, if in addition the $Y_{1}^{(j)}$ come from a normal distribution, the nominal exceedance probabilities for the different thresholds would be achieved on average.
 If a different value $n > 1$ is chosen, this will generally no longer be the case.
@@
 In the notation introduced in Section \ref{sec:definitions}, the first boxplot thus shows observations $X_1^{(j)}, j = 1, \dots, M$, the second $X_2^{(j)}, j = 1, \dots, M$ and so on.
-Little surprisingly (and by definition), the average values get smaller for increasing ranks.
-Interestingly, they also get less dispersed, meaning that the variability between e.g.\ the sixth largest observations of the different seasons is smaller than that of the largest observations.
+Not surprisingly (and in fact by definition), the average values get smaller for increasing ranks.
+Interestingly, they also get less dispersed, meaning that the variability among e.g.\ the sixth largest observations of the different seasons is smaller than that among the largest observations.
 
@@
 As an example, in Spain, applying the MEM with log transformation and $M = 5, n = 6$ leads to 25\% of all seasons exceeding the high threshold and 11\% exceeding the very high threshold.
-This is obviously rather far from the intended excedance probabilities of 10\% and 2.5\% for the two thresholds.
+This is obviously rather far from the intended excedance probabilities of 10\% and 2.5\% for the two thresholds.\todo{Add France low/medium}
 
@@
 
-The increase of thresholds as more historical data accumulates is not only a statistical tendency, but occurs with very high probability.
+The increase of thresholds as more historical data accumulate is not only a statistical tendency, but occurs with very high probability.
 Indeed, for all four countries, the probability that the threshold for high intensity increases when using the first ten rather than only the first five years of a sequence was around 90\%.
 To illutrate this aspect further, we examine the following toy example: We repeat a sequence of five seasons of ILI incidence data from France (2014--2019) four times to obtain a time series of twenty seasons.
-We then apply the moving epidemic method with log transformation and $n = 30/m$ using the first $m = 5, 6, \dots, 20$ years of data, thus mimicking the development of thresholds as more historical data becomes available.
+We then apply the moving epidemic method with log transformation and $n = 30/m$ using the first $m = 5, 6, \dots, 20$ years of data, thus mimicking the development of thresholds as more historical data become available.
 As can be seen from Figure \ref{fig:example_france}, this results in a quite steady increase of thresholds, despite the fact that the data in each five-year block behave exactly the same.
@@
 We provided a statistical assessment of different implementation choices in a simple and widely used framework for the computation of influenza intensity thresholds.
-We found that using a log transformation to intensity measures prior to the computation of thresholds empirically led to better behaved thresholds with closer to nominal exceedance rates.
+We found that log-transforming intensity measures prior to the computation of thresholds empirically led to better behaved thresholds with closer to nominal exceedance rates.
 Concerning the question of how many observations per historical season should be included in the computation of thresholds, we provided analytical arguments and empirical evidence that the common choice $n = 30/m$ leads to too low thresholds when few historical seasons are available.
-We therefore recommend to adopt $n = 1$ irrespective of the number $M$ of available historical seasons.
+We therefore recommend adopting $n = 1$ irrespective of the number $m$ of available historical seasons.
 This is possible using the available implementation in the \texttt{R} package \texttt{mem} and only requires setting \texttt{i.n.max = 1} in the call of the \texttt{memmodel} to override the default value 30/m.
@@
 The emergence of a standard approach at the European level and beyond will contribute to improved comparability of results and facilitate the evaluation and refinement of the applied tool.
-With this work we hope to contribute a statistical perspective on this topic, which can complement public health practitioners' experience from applied analyses.
+With this work we hope to contribute a statistical perspective on this topic that can complement public health practitioners' experience from applied analyses.
 

### 2021-02-10 Draft/Draft_MEM_short.tex
Adding short version.

\documentclass{article}
\usepackage[left=2.5cm,right=2.5cm, top = 2.5cm, bottom = 3cm]{geometry}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{amsmath}

\newcommand{\sd}{s}
\newcommand{\mean}{\bar{y}}

\begin{document}
% \SweaveOpts{concordance=TRUE}

% \title{A bootstrap-based comparison of ILI intensity thresholds from the moving epidemic and WHO methods}
\title{An empirical assessment of influenza intensity thresholds obtained from the moving epidemic and WHO methods}
\author{Johannes Bracher$^{1, 2}$, Jonas Littek$^1$}
\date{ \small
$^1$Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics\\
$^2$Heidelberg Institute for Theoretical Studies, Computational Statistics Group}

\maketitle

\abstract{The moving epidemic method (MEM) and the WHO method are widely used to determine intensity levels for seasonal influenza. The two approaches are conceptually similar, but differ in two aspects. Firstly, the MEM involves a log transformation of incidence data, while the WHO method operates on the original scale. Secondly, the MEM uses more than one observation from each past season to compute thresholds, fixing the total number to include. The WHO method uses only the highest value from each season. We perform simulation studies on how these choices impact thresholds and exceedance proportions. These are based on re-sampling of ILI data from France, Spain, Switzerland and the US. When no transformation is applied, a large proportion of season peaks are classified as high or very high intensity. This can be mitigated by a logarithmic transformation. When fixing the total number of included past observations, thresholds increase the more seasons are available. When only few are available, there is a high chance of classifying a new season peak as high or very high intensity. We therefore suggest using one observation per season and a log transformation, i.e.\ a hybrid of the default settings of the MEM and WHO methods.}

\medskip

\noindent \textbf{Keywords:} influenza-like illness, intensity thresholds, moving epidmeic method, seasonal influenza, WHO method

\section{Introduction}

Following the 2009 influenza H1N1 pandemic, the need for a rapid assessment tool for influenza intensity was recognized as a part of improved pandemic preparedness. The \textit{Review Committee on the Functioning of the
International Health Regulations and on Pandemic Influenza (H1N1)} recommended that member states perform yearly updates and evaluations of intensity thresholds \citep[p.118]{WHO2011}. In the subsequent \textit{WHO Pandemic Infuenza Severity Assessment (PISA)} guideline \citep{WHO2017} the so-called WHO method \citep{WHO2014} and the moving epidemic method (MEM; \citealt{Vega2013, Vega2015}) were recommended to this end. The latter, which has also been recommended by the European Centre for Disease Prevention and Control (e.g., \citealt{ECDC2017}), has been adopted by numerous national public health agencies; see Supplement \textbf{XXX} for an overview of recent applications. The statistical properties of the MEM and WHO methods, however, have not yet been studied in detail. We here address this aspect via simulation experiments based on re-sampling of historical influenza data. Our results indicate that it may be beneficial to use a hybrid of the default settings of these different methods.% These are complement with some analytical statistical results.

% The article is structured as follows. Section \ref{sec:definitions} provides concise definitions of the MEM and WHO methods, highlighting two differences between these conceptually similar approaches. Section \ref{sec:recent_applications} consists of an overview of published applications of the MEM and WHO methods. This is followed by a brief examination of statistical properties of the two methods in Section \ref{sec:analytical}. In Section \ref{sec:simulation} we describe our simulation approach and the results based on historical data from France, Spain, Switzerland and the United States. Section \ref{sec:discussion} concludes with a discussion.

\section{Definition of the moving epidemic and WHO methods}
\label{sec:definitions}

We describe the computation of influenza/influenza-like illness (ILI) intensity thresholds, framing the MEM and WHO methods as two special cases of the same general approach. We assume that thresholds are based on weekly data (typically incidences) from $m$ past seasons and applied to the ($m + 1$)-th season. \cite{Vega2015} recommend to use $5 \leq m \leq 10$ seasons for the MEM to ensure a recent data basis. Computation of thresholds then proceeds as follows:

\begin{enumerate}
\item Select the $n$ largest observations from each of the $m$ past seasons to construct a set of reference observations.
\item Apply an (invertible) transformation $f$ to all selected observations.
\item Assume that the $m \times n$ transformed reference observarions come from a normal distribution and compute estimates $\mean, \sd$ of its mean and standard deviation.
\item Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\mean, \sd^2)$. A common choice for both methods is
\begin{itemize}
\item the 40th percentile $\mean - 0.25 \sd$ as the threshold between low and medium intensity;
\item the 90th percentile $\mean + 1.28 \sd$ as the threshold between medium and high intensity;
\item the 97.5th percentile $\mean + 1.96\sd$ as the threshold between high and very high intensity.
\end{itemize}
\item Transform thresholds back to the original scale by applying the inverse transformation $f^{-1}$ (e.g., the exponential function if $f$ is the natural logarithm).
\end{enumerate}

\noindent The MEM and WHO methods are special cases of this procedure:
\begin{itemize}
\item In the WHO method, $n = 1$ observation per season is used and by default no transformation is applied. If peak incidences vary strongly across seasons, a log transformation is recommended \citep{WHO2017}. 
\item In the MEM method, the default for $f$ is the natural logarithm, while the number of included observations per season is set to $n = 30/m$, rounded to the nearest integer. The total number of historical observations is thus kept approximately fixed.
\end{itemize}
The rationale behind the percentile-based approach is that ``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold'' \citep{WHO2017}.

The implementation of the moving epidemic method in the R package \texttt{mem} \citep{Lozano2020} permits the user to choose $n$, $m$, and $f$ (\texttt{i.type.intensity	= 5} for no transformation, \texttt{i.type.intensity = 6} for the log transformation). The term \textit{moving epidemic method} could thus also be used as an umbrella term for the general procedure described above. We here use it in a more narrow sense for the specification from \cite{Vega2015}, reflected in the default settings of the \texttt{mem} package.

In previous works (\citealt{WHO2014, Vega2015}), the above thresholds have been described as upper ends of one-sided confidence intervals for the the arithmetic (WHO method) or geometric mean (MEM) of the reference observations. This, however, is imprecise terminology as in the computations the standard deviation $\sd$ rather than the standard error $\sd/\sqrt{nm}$ is used (see documentation of the \texttt{mem} package and \citealt{WHO2014}, p. 69). We note that the use of actual confidence intervals is also possible in the \texttt{mem} package (\texttt{i.type.intensity = 1} for the geometric, \texttt{i.type.intensity = 2} for the arithmetic mean). However, thresholds for all levels will then converge to the arithmetic or geometric mean if sufficient historical data are available. As illustrated in Supplement \textbf{XXX}, this is not a desirable behaviour.

\section{Simulaton study}
\label{sec:simulation}

\subsection{Goal}

We aim to assess how thresholds depend on the employed transformation, the number of observations $n$ used per season and the number $m$ of seasons included. In particular, the following aspects are of interest:

\begin{itemize}
\item The application of a logarithmic transformation is expected to lead to higher thresholds for the extreme categories and thus a lower proportion of seasons peaks classified as high or very high.
\item In the MEM, the reference set also contains past obervations which are not actually peak observations. This may introduce a downward bias in thresholds, especially if due to a small number $m$ of available seasons $n$ is large.
\end{itemize}
The latter aspect is intuitive, but also supported by formal statistical considerations detailed in Supplement \textbf{XXX}. These imply that thresholds are unbiased for $n = 1$ in the sense that in the long run they will be exceeded by around 40\%, 90\% and 97.5\% of the season peaks, depending on the level. When $n > 1$, thresholds will tend to be lower and exceeded more frequently.

\subsection{Simulation setup}
\label{subsec:simulation_setup}

We compare four versions of the general approach described in Section \ref{sec:definitions}:
\begin{enumerate}
\item[(a)] No transformation, $n = 1$. This corresponds to the WHO method.
\item[(b)] No transformation, $n = 30/m$.
\item[(c)] Logarithmic transformation, $n = 1$.
\item[(d)] Logarithmic transformation, $n = 30/m$. This corresponds to the MEM approach.
\end{enumerate}
In order to closely mimic the seasonal patterns of influenza, we re-sample historical surveillance data rather than generating fully synthetic data. Assume we have $M$ years of historical data on a measure of influenza activity. We then repeat the following steps 500 times:

\begin{itemize}
\item Sample a sequence of 15 seasons from the $M$ available seasons. This is done with equal probability for each season and \textit{with replacement}, meaning that the same season can appear more than once. This approach is called the \textit{seasonal block bootstrap} \citep{Politis2001}.
\item For each value $m = 5, \dots, 15$:
\begin{itemize}
\item Restrict the generated sequence to the first $m$ seasons.
\item Apply approaches (a)--(d) to compute thresholds for medium (40th percentile), high (90th percentile) and very high intensity (97.5th percentile).
\item Compute which fraction of all $M$ available historical season peaks would be classified as low, moderate, high and very high.
\end{itemize}
\end{itemize}
The range $m =5, \dots, 15$ is motivated by the range of values found in real-world applications, see our overview in Supplement \textbf{XXX}. All analyses were performed using the R language for statistical computing \citep{RCT2020} and the package \texttt{mem} \citep{Lozano2020}.

\subsection{Data}

We use publicly available influenza surveillance data from four countries. Data on the weekly incidence of influenza-like illness in France, 1985--2019, were obtained from Réseau Sentinelles (INSERM/Sorbonne Université, \url{https://www.sentiweb.fr}, \citealt{Flahault2006}). Data on weekly confirmed influenza cases in Spain, 1998--2019, were extracted from graphs shown in the \textit{Informe Anual} of the \textit{Sistema de la vigilancia de gripe en Espa\~na} (\url{https://vgripe.isciii.es/}; \citealt{SVGE2019}). Weekly ILI counts from Switzerland, 2000--2016, collected by the Swiss Federal Office of Public Health are available in the R package \texttt{HIDDA.forecasting} \citep{Held2019}. Weekly weighted ILI (wILI) data at the US national level from CDC FluView \citep{Charbonneau2019}, 1998--2017, were obtained via the CDC FluSight influenza forecasting platform (\url{https://github.com/FluSightNetwork/cdc-flusight-ensemble/}). These wILI values correspond to the fraction of general practitioner visits due to influenza-like symptoms. All four time series are displayed in Figure \ref{fig:data}, with the pandemic 2009/2010 season removed. We also show boxplots of the first through sixth largest observation per season. Not surprisingly, values on average get smaller for increasing ranks. Interestingly, they also get less dispersed, meaning that variability among e.g.\ the sixth largest observations per season is smaller than among the peak values.

% mention exact definition, source, number of seasons

\begin{figure}[h]
\center
\includegraphics[width=0.9\textwidth]{figure/plot_data.pdf}
\caption{Time series of influenza activity measures in four countries: Weekly ILI cases per 100,000 population in France, 1985--2019; weekly number of confirmed influenza cases per 100,000 population in Spain, 1998--2019; weekly ILI cases per 100,000 in Switzerland, 2000--2016; weekly wILI values in the United States, 1998--2017. Off-season weeks are omitted in the plot, with grey lines delimiting the different seasons. The right column shows boxplots of the largest value per season, second largest etc.}
\label{fig:data}
\end{figure}

\subsection{Results}

Results from our simulation study are shown in Figures \ref{fig:results1} (France, Spain) and \ref{fig:results2} (Switzerland and US). In the first and third colum of each figure we also show an analytical approximation of the expected thresholds, see Supplement \textbf{XXX} for details. In all four countries, using a log transformation leads to increased thresholds, in particular for the very high level. Especially in France and Spain, thresholds obtained without the log transformation are too low, as can be seen from the large proportion of season peaks classified as very high (around 10\% when using $n = 1$). As expected, when letting the number of observations used per season depend on the number of available seasons via $n = 30/m$, average thresholds increase in $m$. As a particularly striking example, the  average threshold for high intensity in France (when using a log transformation) increases from 834 for $m = 5, n = 6$ to 1011 for $m = 10, n = 3$ and 1080 for $m = 15, n = 2$. For $n = 1$, in which case thresholds can be interpreted as unbiased (Supplement \textbf{XXX}), the average is 1147. Including historical observations which are not from actual peak weeks thus leads to a considerable lowering of alarm thresholds and will increase the number of alerts for high and very high influenza activity. This is not surprising given the pronounced differences between the distribution of season peak values and e.g. fifth largest observation per season, see right column of Figure \ref{fig:data}. In the US and France, on average one in three season peaks is classified as high and one in six as very high intensity when applying the MEM with log transformation and $m = 5, n = 6$. This is rather far from the intended excedance probabilities of 10\% and 2.5\%.

\begin{figure}
\centering
\includegraphics[page=1, width=0.95\textwidth]{figure/plot_results.pdf}
\caption{Results of re-sampling simulation study, France and Spain. First and third column: mean intensity thresholds along with bands delimited by the empirical 5\% and 95\% quantiles. Analytical approximations of the mean threshold values are marked as small black dots. Second and fourth columns: resulting average shares of seasons classified as low, medium, high and very high intensity.}
\label{fig:results1}
\end{figure}

\begin{figure}
\centering
\includegraphics[page=2, width=0.95\textwidth]{figure/plot_results.pdf}
\caption{Results of re-sampling simulation study, Switzerland and United States. First and third column: mean intensity thresholds along with bands delimited by the empirical 5\% and 95\% quantiles. Analytical approximations of the mean threshold values are marked as small black dots. Second and fourth columns: resulting average shares of seasons classified as low, medium, high and very high intensity.}
\label{fig:results2}
\end{figure}

When always using $n = 1$, the average thresholds and shares of the different categories are more well-behaved also for small $m$. This holds especially when applying a log transformation, even though for the French data a certain mismatch with the nominal exceedance probabilities remains. Also, there is considerable variability in the estimated thresholds. These difficulties, however, are inherent in the problem of estimating a 90\% or 97.5\% quantile based on 5--10 observations.

% The increase of thresholds obtained with $n = 30/m$ as more historical data accumulate is not only a statistical tendency. For all four countries, the probability that the threshold for high intensity increases when using the first ten rather than only the first five years of a sampled sequence was around 90\%. For illustration of this aspect, consider the following example: We repeat a sequence of five ILI seasons from France (2014--2019) four times to obtain a time series of twenty seasons. We then apply the moving epidemic method with log transformation and $n = 30/m$ using the first $m = 5, 6, \dots, 20$ years of data, mimicking the development of thresholds as more historical data become available. As can be seen from the top panel of Figure \ref{fig:example_france}, this results in gradually increasing thresholds, despite the fact that the data in each five-year block behave the same. For example, season 6 is classified as very high intensity, but seasons 11 and 16 are not, even though they are identical to season 6 and the respective data used to construct thresholds follow exactly the same patterns. The bottom panel shows results obtained for the same time series with $n = 1$. While thresholds fluctuate quite a bit (with fluctuations dampening as the time series gets longer), there is no systematic trend.

%\begin{figure}
%\center
%\includegraphics[scale=0.6]{figure/example_france.pdf}
%\caption{Toy example to illustrate increasing thresholds as historical data accumulates: Data from France, 2014/2015--2018/2019 are appended three times and thresholds are computed using $m = 5, \dots, 15$ years of training data. The thresholds based on years 1 through $m$ are overlaid with the incidence time series from the $(m + 1)$-th year.}
%\label{fig:example_france}
%\end{figure}

\section{Discussion}
\label{sec:discussion}

We provided a statistical assessment of implementation choices in a widely used framework for the computation of influenza intensity thresholds. We found applying a log transformation led to better behaved thresholds and closer to nominal exceedance rates. Concerning the question of how many observations per historical season should be included in the computation of thresholds, we found that the common choice $n = 30/m$ results in too low thresholds when few historical seasons are available. We therefore recommend adopting $n = 1$ irrespective of the number of available historical seasons. This is possible in the available \texttt{R} package \texttt{mem} by setting \texttt{i.n.max = 1} when calling the function \texttt{memmodel}.

% We focused on the computation of influenza intensity thresholds, but the same general arguments apply to the computation of thresholds to determine the onset of a season \citep{Vega2012}. As variability among observations around the season onset may be lower than around the peak the problem may be less severe; we did not assess this empirically as not all our data sets fully covered the off-season.

We think that a simple and interpretable tool with a robust software implementation like the MEM is a valuable tool in practice. The use of a standard approach at the European level will improve comparability of results, facilitating the evaluation and refinement of the tool. With this work we hope to contribute a statistical perspective on this topic which complements public health practitioners' experience from applied analyses.

\section*{Data and code}

Materials to reproduce the presented results are available at \url{https://github.com/jbracher/mem}.

\section*{Author contributions}

Study concept: JB; Data processing: JB and JL; Literature review: JB and JL; Statistical analysis / implementation: JB; Writing and editing: JB and JL.

\section*{Acknowledgements}

We would like to thank Laura Werlen for helpful feedback. Johannes Bracher was supported by the Helmholtz Foundation via the SIMCARD Information and Data Science Pilot Project.

\bibliographystyle{apalike}
\bibliography{bibliography_mem}
\end{document}

### 2021-02-10 Draft/Supplement_MEM_short.tex
Adding short version.

--- 2021-02-10 Draft/Draft_MEM.tex
@@
 \newcommand{\sd}{s}
-\newcommand{\mean}{\bar{y}}
+\newcommand{\mean}{\bar{Y}}
 
@@
 
-% \title{A bootstrap-based comparison of ILI intensity thresholds from the moving epidemic and WHO methods}
-\title{A statistical comparison of influenza intensity thresholds obtained from the moving epidemic and WHO methods}
-\author{Johannes Bracher$^{1, 2}$, Jonas Littek$^1$}
+\title{Supplementary Materials for ``An empirical assessment of influenza intensity thresholds obtained from the moving epidemic and WHO methods''}
+\author{Johannes Bracher, Jonas Littek}
 \date{ \small
-$^1$Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics\\
-$^2$Heidelberg Institute for Theoretical Studies, Computational Statistics Group}
+Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics}
 
@@
 
-\abstract{The moving epidemic method (MEM) and the WHO method are widely used approaches to determine intensity levels for seasonal influenza and influenza-like illness (ILI).
-They are conceptually similar, but differ in two aspects.
-Firstly, the MEM involves a log-transformation of incidence data, while the WHO method typically operates on the original scale of the data.
-Secondly, the MEM method by default uses more than one observation from each past season, with a fixed total number of observations to include.
-The WHO method, on the other hand, uses only the single highest value from each season to determine peak intensity thresholds.
-We provide statistical arguments and a simulation study to assess the impact of these choices on the resulting intensity thresholds and empirical exceedance proportions.
-The simulation study is based on a re-sampling approach using historical ILI data from France, Spain, Switzerland and the United States.
-We find that using data on the original scale leads to a rather large proportion of season peaks being classified as high or very high intensity, which can be mitigated to some degree by using a logarithmic transformation.
-When fixing the total number of observations to be included as in the MEM, intensity thresholds tend to increase as more historical data becomes available.
-When data are only available for few seasons, there is a high chance of classifying a new season as high or very high intenstity.
-We therefore advocate always using one observation per season with a log transformation, i.e.\ a hybrid between the default settings of the MEM and WHO methods.
-This adaptation is possible within an existing software implementation of the MEM approach and thus easy to apply.}
+\section{Formal statistical analysis}
 
-\section{Introduction}
-
-Following the 2009 influenza H1N1 pandemic, the need for a rapid assessment tool for influenza intensity was recognized. In a report, the \textit{Review Committee on the Functioning of the
-International Health Regulations and on Pandemic Influenza (H1N1)} recommended that member states apply and evaluate their methods for severity assessment every year, thus improving their pandemic preparedness \citep[p.118]{WHO2011}.
-In the subsequent \textit{WHO Pandemic Infuenza Severity Assessment (PISA)} guideline \citep{WHO2017}, two specific statistical methods are recommended to determine influenza intensity thresholds.
-Firstly, the so-called \textit{WHO method}, already described in a previous document \citep{WHO2014} and in more detail in the PISA guide; and secondly, the \textit{moving epidemic method} (MEM; \citealt{Vega2013, Vega2015}) which is also recommended by the European Centre for Disease Prevention and Control (e.g., \citealt{ECDC2017}).
-
-In particular the MEM approach has in the meantime been adopted by numerous national public health agencies (see references in Section \ref{sec:recent_applications}) and is considered to deliver practically useful thresholds.
-Its application has been facilitated by the availablility of a user-friendly and well-documented \texttt{R} package (\texttt{mem}; \citealt{Lozano2020}), which can even be used via a graphical user interface \citep{Lozano2018}.
-The statistical properties of the MEM and WHO methods, however, have not yet been studied in detail.
-We obtain some analytical results, complemented by simulation experiments.
-In order to closely mimic real influenza patterns, the latter are based on bootstrap re-sampling of historical surveillance data.
-In this way, we provide an improved understanding and intuition for the consequences of different implementation choices.
-
-The article is structured as follows.
-Section \ref{sec:definitions} provides concise definitions of the MEM and WHO methods, highlighting two differences between these conceptually similar approaches.
-Section \ref{sec:recent_applications} consists of an overview of published applications of the MEM and WHO methods.
-This is followed by a brief examination of statistical properties of the two methods in Section \ref{sec:analytical}.
-In Section \ref{sec:simulation} we describe our simulation approach and the results based on historical data from France, Spain, Switzerland and the United States.
-Section \ref{sec:discussion} concludes with a discussion.
-
-\section{Definition of the moving epidemic and WHO methods}
+\subsection{Re-stating the definitions}
 \label{sec:definitions}
 
-In the following we focus on the construction of thresholds  for peak intensity, framing the MEM and WHO methods as different special cases of the same general approach.
-Details on the construction of baseline epidemic thresholds from pre-season observations are omitted.
-We assume that thresholds are based on influenza intensity data (typically weekly incidences) from $m$ past seasons and applied to the ($m + 1$)-th season.
-\cite{Vega2015} recommend to use $5 \leq m \leq 10$ seasons for the MEM to ensure a recent data basis.
+We start by re-stating the definition of the MEM and WHO method form Section \textbf{XXX} of the main manuscript in a slighty more formal way and introducing relevant notation.
+We assume again that thresholds are based on influenza intensity data from $m$ past seasons and applied to the ($m + 1$)-th season.
 Thresholds for a given intensity measure $X$ are then obtained via the following steps.
-In the following we will call $X$ the incidence for simplicity, even if it can also be a different measure.
 
@@
 \item Apply an (invertible) transformation $y_i^{(j)} = f(x_i^{(j)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
-\item Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates $\mean, \sd$ of its mean and standard deviation.
-\item Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\mean, \sd^2)$, i.e.\ compute
+\item Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates $\bar{y}, \sd$ of its mean and standard deviation.
+\item Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\bar{y}, \sd^2)$, i.e.\ compute
 \begin{equation}
-q_{Y, \alpha} = \mean + z_\alpha \sd, \label{eq:def_q}
+q_{Y, \alpha} = \bar{y} + z_\alpha \sd, \label{eq:def_q}
 \end{equation}
-where $z_\alpha$ is the $\alpha$ quantile of the standard normal distribution.
-A common choice for both methods \citep{WHO2017}, also implemented as the default in the \texttt{R} package \texttt{mem} \citep{Lozano2020} is to use
+where $z_\alpha$ is the $\alpha$ quantile of the standard normal distribution. A common choice for both methods is
 \begin{itemize}
@@
 The standard procedure is to apply no transformation (i.e.\ $f$ is set to the identity function).
-If peak incidences vary strongly across seasons, a log-tranformation is recommended \citep{WHO2017}.
 \item In the MEM method, the default choice for $f$ is the natural logarithm, while the number of observations per season is set to $n = 30/m$.
@@
 \end{itemize}
-The software implementation of the moving epidemic method in the \texttt{R} package \texttt{mem} is rather general and permits the user to choose $n$, $m$, and $f$ (the identify function coresponding to \texttt{i.type.intensity	= 5} and logarithmic function to \texttt{i.type.intensity = 6}).
-The term \textit{moving epidemic method} could thus also be used as an umbrella term for the general procedure described above.
-We use it here in a more narrow sense to refer to the specific procedure described in \cite{Vega2015}, \cite{WHO2017} and reflected in the default parameters of the \texttt{mem} package.
 
-The rationale behind the percentile-based approach is that ``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold'' \citep{WHO2017}.
-Note that in most previous descriptions (e.g., \ \citealt{WHO2014, Vega2015}), the thresholds have been referred to as upper ends of one-sided confidence intervals associated with the the arithmetic (WHO method) or geometric mean (MEM) of the historical reference observations in $\mathcal{X}$.
-This, however, is imprecise terminology as in the computations the standard deviation $\sd$ is used rather than the estimated standard error $\sd/\sqrt{nm}$ (see the default settings in the R package \texttt{mem} and the description of the WHO method in \citealt{WHO2014}, p.
-69).
-We note that the use of actual confidence intervals is in fact also possible using the \texttt{mem} package (\texttt{i.type.intensity = 1} for the geometric mean, \texttt{i.type.intensity = 2} for the arithmetic mean).
-However, under this choice, the thresholds for all levels eventually converge to the arithmetic or geometric mean if sufficient historical data are available (as $\sd/\sqrt{nm}$ gets smaller and smaller).
-This is obviously not the intended behaviour.
-To illustrate this aspect we also provide simulation studies on these two choices, but these have been deferred to Appendix \textbf{XXX}.
+\subsection{Results}
 
-% see function iconfianza.aritmetica
+We start by considering how the expectations of the empirical mean $\bar{Y}$ and standard deviation $S$ entering into equation \eqref{eq:def_q} depend on $n$ (note that we now capitalize these as we consider the random variables rather than concrete observed values).
+Denote by $\mathbf{Y}^{(j)}$ the vector of the $n$ largest transformed incidences from season $j$ in decreasing order, i.e.
+$\mathbf{Y}^{(j)} = (Y^{(j)}_1, \dots, Y^{(j)}_n)^\top$ with $Y^{(j)}_1 \geq Y^{(j)}_2 \geq \dots \geq Y^{(j)}_n$.
+Now assume that $\mathbf{Y}^{(1)}, \dots, \mathbf{Y}^{(m)}$ are independent and identically distributed with mean and covariance matrix
+\begin{align}
+\mathbb{E}\left(\mathbf{Y}^{(j)}\right) = \boldsymbol{\mu} = \left(\begin{array}{c}
+\mu_1\\
+\vdots\\
+\mu_n
+\end{array}\right) \ \ \ \text{and} \ \ \ \text{Cov}\left(\mathbf{Y}^{(j)}\right) = \boldsymbol{\Sigma} =
+\left(\begin{array}{ccc}
+\sigma_{1, 1} & \cdots & \sigma_{1, n}\\
+\vdots & \ddots &\vdots\\
+\sigma_{n, 1} & \cdots & \sigma_{n, n}
+\end{array}\right),
+\end{align}
+respectively, where to make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$.
+It can then be shown that the expectations of $\mean$ and $\sd^2$ when using $m$ seasons and $n$ observations from each season are given by
+\begin{align}
+\mathbb{E}(\mean) & = \frac{1}{n} \sum_{i = 1}^n \mu_i,
+\label{eq:expectation_mu}\\
+\mathbb{E}(\sd^2) & = \frac{m}{mn - 1} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{n(mn - 1)} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ - \ \frac{m}{n(mn - 1)}\left(\sum_{i = 1}^n \mu_i\right)^2,
+\label{eq:expectation_sigma2}
+\end{align}
+%\begin{align}
+%\mathbb{E}(\mean) & = \frac{1}{K} \sum_{i = 1}^K \mu_i,
+%\label{eq:expectation_mu}\\
+%\mathbb{E}(\sd^2) & = \frac{n}{nK - 1} \sum_{i = 1}^K (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{K(nK - 1)} \sum_{i = 1}^K \sum_{i' = 1}^K \sigma_{i,i'} \ - \ \frac{n}{K(nK - 1)}\left(\sum_{i = 1}^K \mu_i\right)^2,
+%\label{eq:expectation_sigma2}
+%\end{align}
+respectively. The derivation is provided in the next subsection. For reasons detailed there, if the transformation function $f$ is the identity or the natural logarithm,
+\begin{equation}
+\mathbb{E}(q_\alpha) \approx f^{-1}\left\{\mathbb{E}(\mean) + z_\alpha \sqrt{\mathbb{E}(\sd^2)}\right\}
+\label{eq:expectation_q}
+\end{equation}
+usually holds in good approximation in our applied setting (with $f^{-1}$ the idenity or exponential function). This implies that for $n = 1$, we have
+$$
+\mathbb{E}(q_\alpha) \approx f^{-1}(\mu_1 + z_\alpha \sigma_1).
+$$
+This is a desirable property.
+Loosely speaking, if in addition the $Y_{1}^{(j)}$ come from a normal distribution, the nominal exceedance probabilities for the different thresholds would be achieved on average.
+If a different value $n > 1$ is chosen, this will generally no longer be the case.
+Equations \eqref{eq:expectation_mu}--\eqref{eq:expectation_q} tell us by how much $q_\alpha$ can be expected to differ from the intended value $f^{-1}(\mu_1 + z_\alpha \sigma_1)$.
+By the definition of the $\mu_i$ (with $\mu_i$ the expectation of the $i$-th largest observation in a given season), $\mathbb{E}(\mean)$ decreases in $n$.
+In most (possibly all) settings this also translates to the $q_\alpha$.
+As a consequence, when choosing $n > 1$, one must expect to classify a larger number of season peaks as high or very high intensity.
+When choosing $n = 30/m$ as suggested for the MEM, threholds will then tend to increase the more years of data are used to compute thresholds.
+
+\subsection{Derivation}
+
+We assume that the different historical seasons are independent realizations of the same random process, which is certainly not strictly true, but should hold in good approximation.
+Then $\mathbf{Y}^{(1)}, \dots, \mathbf{Y}^{(m)}$ are independent and identically distributed with mean and covariance matrix
+\begin{align}
+\mathbb{E}(\mathbf{Y}^{(j)}) = \boldsymbol{\mu} = \left(\begin{array}{c}
+\mu_1\\
+\vdots\\
+\mu_n
+\end{array}\right) \ \ \ \text{and} \ \ \ \text{Cov}(\mathbf{Y}^{(j)}) = \boldsymbol{\Sigma} =
+\left(\begin{array}{ccc}
+\sigma_{1, 1} & \cdots & \sigma_{1, n}\\
+\vdots & \ddots &\vdots\\
+\sigma_{n, 1} & \cdots & \sigma_{n, n}
+\end{array}\right),
+\end{align}
+respectively, where to make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$. We are interested in the expectations of the random variables $\mean$ and $\sd$, where
+\begin{align*}
+\mean & = \frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n Y_i^{(j)}\\
+\sd & = \sqrt{\frac{1}{mn - 1} \sum_{j = 1}^m \sum_{i = 1}^n \left(Y_i^{(j)} - \mean\right)^2}.
+\end{align*}
+It is straightforward to see that 
+\begin{equation}
+\mathbb{E}(\mean) = \frac{1}{n} \sum_{i = 1}^n \mu_i.
+\label{eq:expectation_mu}
+\end{equation}
+Instead of the empirical standard deviation $\sd$ we first consider the variance $\sd^2$, which can also be written as
+\begin{align}
+\sd^2 & = \frac{mn}{mn - 1} \left\{\frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n \left(Y_i^{(j)} - \mean\right)^2 \right\}\\
+& = \frac{mn}{mn - 1} \left\{ \underbrace{\frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n Y_i^{(j)2}}_{= a} \ \ \ - \ \ \ \underbrace{\left(\frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n Y_i^{(j)} \right)^2}_{= b} \right\} \label{eq:sigma2hat}
+\end{align}
+%$$
+%\sd^2 = \underbrace{\left(\frac{1}{nK - 1} \sum_{j = 1}^n \sum_{i = 1}^K Y_i^{(j)2}\right)}_{ = a} \ \ \ - \ \ \ \underbrace{\left(\frac{1}{nK - 1} \sum_{j = 1}^n \sum_{i = 1}^K Y_i^{(j)} \right)^2}_{= b}.
+%$$
+We consider the two terms $a$ and $b$ separately, starting by
+\begin{align*}
+\mathbb{E}(a) & = \frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n \mathbb{E}\left(Y_i^{(j)2}\right)\\
+& = \frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n \left\{ \text{Var}\left(Y_i^{(j)}\right) + \mathbb{E}\left(Y_i^{(j)}\right)^2 \right\}\\
+& = \frac{m}{mn} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2).
+\end{align*}
+Then we note that
+\begin{align*}
+\mathbb{E}(b) & = \text{Var}\left( \frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n Y_i^{(j)} \right) \ \ \ + \ \ \ \mathbb{E}\left(\frac{1}{mn}  \sum_{j = 1}^m \sum_{i = 1}^n Y_i^{(j)} \right)^2\\
+& = \frac{1}{(mn)^2}\sum_{j = 1}^m \text{Var}\left(\sum_{i = 1}^n Y_i^{(j)} \right) \ \ \ + \ \ \ \left(\frac{m}{mn} \sum_{i = 1}^n \mu_i\right)^2\\
+& = \frac{m}{(mn)^2} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ \ \ + \ \ \ \frac{m^2}{(mn)^2}\left(\sum_{i = 1}^n \mu_i\right)^2.
+\end{align*}
+Plugging this back into equation \eqref{eq:sigma2hat} we obtain
+\begin{equation}
+\mathbb{E}(\sd^2) = \frac{m}{mn - 1} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2) \ \ \ - \ \ \ \frac{1}{n(mn - 1)} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ \ \ - \ \ \ \frac{m}{n(mn - 1)}\left(\sum_{i = 1}^K \mu_i\right)^2.
+\label{eq:expectation_sigma2}
+\end{equation}
+It is straightforward to see that for $n = 1$ the expressions \eqref{eq:expectation_mu} and \eqref{eq:expectation_sigma2} simplify to
+$$
+\mathbb{E}(\mean^2) = \mu_1 \ \ \text{ and } \ \ \mathbb{E}(\sd^2) = \sigma^2_1.
+$$
+
+There is no general way of computing the expectation $\mathbb{E}(\sd)$, but unless the true distribution of  $\sd$ has strong excess curtosis,
+\begin{equation}
+\mathbb{E}(\sd) \approx \sqrt{\mathbb{E}(\sd^2)}
+\label{eq:expectation_sigma}
+\end{equation}
+is a reasonable approximation.
+We can then plug equations \eqref{eq:expectation_mu} and \eqref{eq:expectation_sigma} into the formulae for the thresholds $q_{Y, \alpha}$ (with $\alpha \in \{0.4, 0.9, 0.975\}$) on the transformed scale and obtain
+$$
+\mathbb{E}(q_{Y, \alpha}) \approx \mathbb{E}(\mean) + z_\alpha \sqrt{\mathbb{E}(\sd^2)},
+$$
+where $z_\alpha$ is the $\alpha$ quantile of the standard normal distribution.
+
+If $f$ was set to the natural logarithm, the question remains how to obtain statements concerning thresholds $q_\alpha$ on the original scale. Approximation via a second-order Taylor expansion yields
+\begin{equation}
+\mathbb{E}(q_\alpha) = \mathbb{E}\left\{\exp(q_{Y, \alpha})\right\} \approx \exp\left\{\mathbb{E}(q_{Y, \alpha})\right\} \times \left\{1 + \frac{\text{Var}(q_{Y, \alpha})}{2} \right\}.
+\label{eq:taylor}
+\end{equation}
+
+Empirically, after transformation to the log scale, the variance of season peak values is low (between 0.02 and 0.4 for the data sets analysed in Section 5 of the main manuscript).
+The resulting variances of $q_{Y, \alpha}$ are then quite small and do not play an important role in equation \eqref{eq:taylor}.
+We can thus use the even simpler approximation
+\begin{equation}
+\mathbb{E}(q_\alpha) \approx \exp\left\{\mathbb{E}(q_{Y, \alpha})\right\}.
+\end{equation}
+
+\newpage
+
+\section{Intensity thresholds based on confidence intervals}
+
+Previous works have referred to the thresholds discussed in our manuscript as upper ends of one-sided confidence intervals associated with the the arithmetic (WHO) or geometric mean (MEM) of the reference observations (\citealt{WHO2014, Vega2015}).
+This is an imprecise use of terminology as the thresholds are computed using the standard deviation $\sd$.
+Confidence intervals, on the other hand, would be computed using the standard error $\sd/\sqrt{nm}$.
+A more precise term would thus be ``prediction interval'', which implies that the interval refers to one future realization rather than a theoretical mean.
+
+These terminology questions aside, we note that the \texttt{mem} R package also offers threshold computation based on actual confidence intervals (\texttt{i.type.intensity = 1} for the geometric, \texttt{i.type.intensity = 2} for the arithmetic mean).
+On the transformed scale these are computed as
+\begin{equation}
+q_{Y, \alpha} = \bar{y} + z_\alpha \frac{\sd}{\sqrt{mn}}. \label{eq:q_ci}
+\end{equation}
+These thresholds show somewhat peculiar behaviour.
+We illustrate this by repeating the simulation study from Section \textbf{XXX} of the main manuscript for France with the respective settings.
+The result is shown in Figure \ref{fig:france_ci}.
+
+\begin{figure}[h!]
+\includegraphics[width=0.9\textwidth]{figure/plot_results_ci.pdf}
+\caption{Simulation results for France as in Figure \textbf{XX} of the main manuscript, but using confidence intervals based on standard errors rather than prediction intervals based on standard deviations.}
+\label{fig:france_ci}
+\end{figure}
+
+Here, two different mechanisms are at work.
+If one chooses $n = 1$ per season irrespective of the number $m$ of available seasons, the number of included observations $mn$ increases in $m$.
+With growing $m$ this leads to confidence intervals which get narrower and a funnel-like pattern in the different thresholds (bottom row of Figure \ref{fig:france_ci}.
+If one chooses $n = 30/m$, the number $mn$ will remain constant (or approximately, as some rounding is necessary).
+The funnel-like shape is thus not observed and an increasing pattern like in our main analyses emerges instead.
+In both cases, there is an excessive number of seasons classified as very high intensity.
+This reflects the fact that the extreme thresholds necessarily get smaller when using the standard error rather than standard deviation in equation \eqref{eq:q_ci}.
+We thus conclude that these variants of the moving epidemic method (corresponding to options \texttt{i.type.intensity = 1} and  \texttt{i.type.intensity = 2} in the R package \texttt{mem}) should not be applied in practice.
 
@@
 
-% Cite Benedetti who use it as a "gold standard". Mention Green GB who compare to percentile method. Vos mention fixed criterion model -- what is that?
-
-\section{Analytical considerations on the choice of $n$}
-\label{sec:analytical}
-
-As detailed in the previous section, the MEM and WHO methods have been applied extensively in practice and overall seem to produce thresholds that are considered useful by domain experts.
-A more statistical assessment, however has so far been missing.
-Determining which choice of transformation function $f$ is most reaonable is largely an empirical question.
-Little more can be said than that a logarithmic transformation can be helpful if the distribution of peak values is skewed.
-However, some statements can be made on how the value chosen for $n$ impacts the expected results for the thresholds $q_\alpha$.
-
-We start by considering how the expectations of the empirical mean $\bar{Y}$ and standard deviation $S$ entering into equation \eqref{eq:def_q} depend on $n$ (note that we now capitalize these as we consider the random variables rather than concrete observed values).
-Denote by $\mathbf{Y}^{(j)}$ the vector of the $n$ largest transformed incidences from season $j$ in decreasing order, i.e.
-$\mathbf{Y}^{(j)} = (Y^{(j)}_1, \dots, Y^{(j)}_n)^\top$ with $Y^{(j)}_1 \geq Y^{(j)}_2 \geq \dots \geq Y^{(j)}_n$.
-Now assume that $\mathbf{Y}^{(1)}, \dots, \mathbf{Y}^{(m)}$ are independent and identically distributed with mean and covariance matrix
-\begin{align}
-\mathbb{E}\left(\mathbf{Y}^{(j)}\right) = \boldsymbol{\mu} = \left(\begin{array}{c}
-\mu_1\\
-\vdots\\
-\mu_n
-\end{array}\right) \ \ \ \text{and} \ \ \ \text{Cov}\left(\mathbf{Y}^{(j)}\right) = \boldsymbol{\Sigma} =
-\left(\begin{array}{ccc}
-\sigma_{1, 1} & \cdots & \sigma_{1, n}\\
-\vdots & \ddots &\vdots\\
-\sigma_{n, 1} & \cdots & \sigma_{n, n}
-\end{array}\right),
-\end{align}
-respectively, where to make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$.
-It can then be shown that the expectations of $\mean$ and $\sd^2$ when using $m$ seasons and $n$ observations from each season are given by
-\begin{align}
-\mathbb{E}(\mean) & = \frac{1}{n} \sum_{i = 1}^n \mu_i,
-\label{eq:expectation_mu}\\
-\mathbb{E}(\sd^2) & = \frac{m}{mn - 1} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{n(mn - 1)} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ - \ \frac{m}{n(mn - 1)}\left(\sum_{i = 1}^n \mu_i\right)^2,
-\label{eq:expectation_sigma2}
-\end{align}
-%\begin{align}
-%\mathbb{E}(\mean) & = \frac{1}{K} \sum_{i = 1}^K \mu_i,
-%\label{eq:expectation_mu}\\
-%\mathbb{E}(\sd^2) & = \frac{n}{nK - 1} \sum_{i = 1}^K (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{K(nK - 1)} \sum_{i = 1}^K \sum_{i' = 1}^K \sigma_{i,i'} \ - \ \frac{n}{K(nK - 1)}\left(\sum_{i = 1}^K \mu_i\right)^2,
-%\label{eq:expectation_sigma2}
-%\end{align}
-respectively. The derivation is provided in Supplement \textbf{XXX}. For reasons detailed there, if the transformation function $f$ is the identity or the natural logarithm,
-\begin{equation}
-\mathbb{E}(q_\alpha) \approx f^{-1}\left\{\mathbb{E}(\mean) + z_\alpha \sqrt{\mathbb{E}(\sd^2)}\right\}
-\label{eq:expectation_q}
-\end{equation}
-usually holds in good approximation in our applied setting (with $f^{-1}$ the idenity or exponential function). This implies that for $n = 1$, we have
-$$
-\mathbb{E}(q_\alpha) \approx f^{-1}(\mu_1 + z_\alpha \sigma_1).
-$$
-This is a desirable property.
-Loosely speaking, if in addition the $Y_{1}^{(j)}$ come from a normal distribution, the nominal exceedance probabilities for the different thresholds would be achieved on average.
-If a different value $n > 1$ is chosen, this will generally no longer be the case.
-Equations \eqref{eq:expectation_mu}--\eqref{eq:expectation_q} tell us by how much $q_\alpha$ can be expected to differ from the intended value $f^{-1}(\mu_1 + z_\alpha \sigma_1)$.
-By the definition of the $\mu_i$ (with $\mu_i$ the expectation of the $i$-th largest observation in a given season), $\mathbb{E}(\mean)$ decreases in $n$.
-In most (possibly all) settings this also translates to the $q_\alpha$.
-As a consequence, when choosing $n > 1$, one must expect to classify a larger number of season peaks as high or very high intensity.
-When choosing $n = 30/m$ as suggested for the MEM, threholds will then tend to increase the more years of data are used to compute thresholds.
-
-\section{Simulaton study}
-\label{sec:simulation}
-
-\subsection{Goal}
-\label{subsec:simulation_goal}
-
-To complement the analytical results from the previous section we perform a simulation study to compare four versions of the general approach described in Section \ref{sec:definitions}:
-\begin{enumerate}
-\item[(a)] No transformation, $n = 1$. This corresponds to the WHO method.
-\item[(b)] No transformation, $n = 30/m$.
-\item[(c)] Logarithmic transformation, $n = 1$.
-\item[(d)] Logarithmic transformation, $n = 30/m$. This corresponds to the MEM approach.
-\end{enumerate}
-
-Our main focus is on the resulting intensity thresholds and shares of seasons classified as low, medium, high and very high.
-We will put the analytical approximations from the previous section to the test and assess the impact of the chosen $n$ and $f$ on thresholds and exceedance rates empirically.
-%Also, we will examine the effect of applying a logarithmic transformation to the data.
-% As mentioned before, the medium, high and extraordinary threholds should be exceeded by approximately 60\%, 10\% and 2.5\% respectively \citep{WHO2017}.
-Concerning the choice between an identity and logarithmic transformation -- or, equivalently, the choice between arithmetic and geometric mean -- one may expect that the logarithm is more appropriate if observations are skewed.
-Concerning the number of observations used per available season, the choice $K = 30/n$ has the advantage of ensuring a reasonably large number of reference observations.
-However, this comes at the price of using observations which are not actually from peak weeks, and will by definition be lower than the actual peeks.
-If $n$ is low, this may lead to too low thresholds and a larger number of seasons classified as high or very high intensity.
-We will assess empirically whether such an effect can be observed, and how pronounced it is depending on the number of available historical seasons.
-
-\subsection{Bootstrap approach based on historical ILI data}
-
-A pure simulation study requires careful specification of a data-generating process mimicking the seasonal patterns of influenza.
-Given the considerable variability in timing and intensity of season peaks, this is not a trivial task.
-We therefore resort to a bootstrap approach based on historical data rather than generating synthetic data.
-Assume we have $M$ years of historical data on a measure of influenza activity.
-We then repeat the following steps a total of 500 times:
-
-\begin{itemize}
-\item Sample an ordered sequence of 15 seasons from the $M$ available seasons.
-As commonly done in bootstrap experiments, this is done with equal probability for each season and \textit{with replacement}, meaning that the same season can appear more than once in the sampled sequence.
-This approach is also referred to as the \textit{seasonal block bootstrap} \citep{Politis2001}.
-\item For each value $m = 5, \dots, 15$:
-\begin{itemize}
-\item Restrict the generated sequence to the first $m$ seasons.
-\item Apply approaches (a)--(d) as described in the previous subsection.
-Compute and store intensity thresholds for medium (40th percentile), high (90th percentile) and very high intensity (97.5th percentile).
-\item Compute which fraction of all $M$ available historical seasons would be classified as low, moderate, high and very high using the obtained thresholds.
-\end{itemize}
-\end{itemize}
-We can then assess how the thresholds and shares of the different intensity levels depend on the employed transformation function, the number of observations $n$ used per season and the number $m$ of seasons included.
-Here, the range $m =5, \dots, 15$ is motivated by the range of values found in real-world applications of the moving epidemic method as listed in Section \ref{sec:recent_applications}.
-
-\subsection{Data}
-
-We use publicly available influenza surveillance data from four countries.
-Data on the weekly incidence of influenza-like-illness in France, 1985--2019, were obtained from the Réseau Sentinelles web platform (INSERM/Sorbonne Université, \url{https://www.sentiweb.fr}, \citealt{Flahault2006}).
-Population data were obtained from Institut national de la statistique et des études économiques.
-Data on weekly confirmed influenza cases in Spain, 1998--2019, were extracted from graphs shown in the \textit{Informe Anual} of the \textit{Sistema de la vigilancia de gripe en Espa\~na} (\url{https://vgripe.isciii.es/}; \citealt{SVGE2019}) using the tool \textit{WebPlotDigitizer} (\url{https://apps.automeris.io/wpd/}).
-Weekly ILI counts from Swiss routine surveillance, collected by the Swiss Federal Office of Public Health, 2000--2016, were obtained from the R package \texttt{HIDDA.forecasting} \citep{Held2019}.
-Population data were obtained from the Swiss Federal Statistical Office.
-Weekly weighted ILI (wILI) data at the US national level from CDC FluView \citep{Charbonneau2019}, 1998--2017, were obtained via the CDC FluSight influenza forecasting platform (\url{https://github.com/FluSightNetwork/cdc-flusight-ensemble/}).
-These wILI values correspond to the fraction of general practitioner visits due to influenza-like symptoms
-
-All four time series are displayed in Figure \ref{fig:data}.
-Note that we removed the pandemic 2009/2010 influenza season from each series as it is not comparable to the usual years with only seasonal influenza.
-We also show boxplots of the first through sixth largest observation per season (right column).
-In the notation introduced in Section \ref{sec:definitions}, the first boxplot thus shows observations $X_1^{(j)}, j = 1, \dots, M$, the second $X_2^{(j)}, j = 1, \dots, M$ and so on.
-Not surprisingly (and in fact by definition), the average values get smaller for increasing ranks.
-Interestingly, they also get less dispersed, meaning that the variability among e.g.\ the sixth largest observations of the different seasons is smaller than that among the largest observations.
-
-% mention exact definition, source, number of seasons
-
-\begin{figure}[h]
-\center
-\includegraphics[width=0.9\textwidth]{figure/plot_data.pdf}
-\caption{Time series of influenza activity measures in four countries: Weekly ILI cases per 100,000 population in France, 1985--2019; weekly number of confirmed influenza cases per 100,000 population in Spain, 1998--2019; weekly ILI cases per 100,000 in Switzerland, 2000--2016; weekly wILI values in the United States, 1998--2017.
-Off-season weeks are omitted in the plot, with grey lines delimiting the different seasons.}
-\label{fig:data}
-\end{figure}
-
-\subsection{Used software}
-
-All analyses were performed using the R language for statistical computing, version 4.0.3 \citep{RCT2020}, in particular using the package \texttt{mem}, version 2.16 \citep{Lozano2020}.
-Note that the function \texttt{mem::memmodel} provides sufficient flexibility to apply all four approaches (a)--(d) from Section \ref{subsec:simulation_goal}.
-
-\subsection{Results}
-
-Results from our bootstrap simulation study for the four considered countries are shown in Figures \ref{fig:results1} (France, Spain) and \ref{fig:results2} (Switzerland and US).
-In the first and third colum of each figure we also show the analytical approximations from Section \ref{sec:analytical} as small black dots.
-To compute these approximations, we just plugged the empirical mean vectors and covariance matrices into equations \eqref{eq:expectation_mu} -- \eqref{eq:expectation_q}.
-As can be seen, the agreement with the bootstrap results is very good.
-
-In all four countries, the thresholds obtained when using a log transformation as in the MEM are considerably higher than when no transformation is used.
-Especially in France and Spain, thresholds obtained without the log transformation are clearly too low, as can be seen from the large proportion of seasons classified as very high intensity (around 10\% when using $n = 1$).
-Applying the log transformation mitigates this to some degree.
-
-As expected, when letting the number of observations used per season depend on the number of available seasons via $n = 30/m$, average thresholds increase in $m$.
-As a particularly striking example, the average threshold for high intensity for France when using a log transformation increase from %$q_{0.9} = 834, q_{0.975} = 1158$ for $m = 5, n = 6$ to $q_{0.9} = 1011, q_{0.975} = 1363$ for $m = 10, n = 3$ and $q_{0.9} = 1080, q_{0.975} = 1458$ for $m = 15, n = 2$.
-For $n = 1$, in which case the estimators can be interpreted as unbiased (Section \ref{sec:analytical}), we obtain averages of $q_{0.9} = 1147, q_{0.975} = 1559$.
-834 for $m = 5, n = 6$ to 1011 for $m = 10, n = 3$ and 1080 for $m = 15, n = 2$.
-For $n = 1$, in which case the estimators can be interpreted as unbiased (Section \ref{sec:analytical}), we obtain an average of 1147.
-This corresponds to an increase of roughly 35\% relative to the thresholds obtained with $n = 6$.
-Including historical observations which are not from actual peak weeks thus leads to a considerable lowering of alarm thresholds and will necessarily increase the number of alerts for high and very high influenza activity.
-As an example, in Spain, applying the MEM with log transformation and $M = 5, n = 6$ leads to 25\% of all seasons exceeding the high threshold and 11\% exceeding the very high threshold.
-This is obviously rather far from the intended excedance probabilities of 10\% and 2.5\% for the two thresholds.\todo{Add France low/medium}
-
-\begin{figure}
-\centering
-\includegraphics[page=1, width=0.95\textwidth]{figure/plot_results.pdf}
-\caption{Results of bootstrap simulation study for France and Spain.
-First and third column: mean intensity thresholds along with bands delimited by the empirical 5\% and 95\% quantiles.
-The approximate mean threshold values obtained analytically are marked as little black dots.
-Second and fourth columns: resulting average shares of seasons classified as low, medium, high and very high intensity (shown as in a stacked barplot).}
-\label{fig:results1}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[page=2, width=0.95\textwidth]{figure/plot_results.pdf}
-\caption{Results of bootstrap simulation study for Switzerland and the United States.
-First and third column: mean intensity thresholds along with bands delimited by the empirical 5\% and 95\% quantiles.
-The approximate mean threshold values obtained analytically are marked as little black dots.
-Second and fourth columns: resulting average shares of seasons classified as low, medium, high and very high intensity (shown as in a stacked barplot).}
-\label{fig:results2}
-\end{figure}
-
-The increase of thresholds as more historical data accumulate is not only a statistical tendency, but occurs with very high probability.
-Indeed, for all four countries, the probability that the threshold for high intensity increases when using the first ten rather than only the first five years of a sequence was around 90\%.
-To illutrate this aspect further, we examine the following toy example: We repeat a sequence of five seasons of ILI incidence data from France (2014--2019) four times to obtain a time series of twenty seasons.
-We then apply the moving epidemic method with log transformation and $n = 30/m$ using the first $m = 5, 6, \dots, 20$ years of data, thus mimicking the development of thresholds as more historical data become available.
-As can be seen from Figure \ref{fig:example_france}, this results in a quite steady increase of thresholds, despite the fact that the data in each five-year block behave exactly the same.
-For example, season 6 is classified as very high intensity based on seasons 1 through 5, but seasons 11 and 16 are not, even though they are identical to season 6 and the respective data used to construct thresholds follow exactly the same patterns.
-A similar discrepancy can be observed for season 7, which is classified as medium intensity, while seasons 12 and 17 are marked as low intensity.
-
-\begin{figure}
-\center
-\includegraphics[scale=0.6]{figure/example_france.pdf}
-\caption{Toy example to illustrate increasing thresholds as historical data accumulates: Data from France, 2014/2015--2018/2019 are appended three times and thresholds are computed using $m = 5, \dots, 15$ years of training data.
-The thresholds based on years 1 through $m$ are overlaid with the incidence time series from the $(m + 1)$-th year.}
-\label{fig:example_france}
-\end{figure}
-
-When just using $n = 1$ irrespective of the number of historical seasons available, the average thresholds and shares of the different categories are more well-behaved also for small $m$.
-Of course, there is considerable variability in the estimated thresholds.
-This difficulty, however, is inherent in the problem of estimating a 90\% or 97.5\% quantile based on only 5--10 observations.
-
-\section{Discussion}
-\label{sec:discussion}
-
-We provided a statistical assessment of different implementation choices in a simple and widely used framework for the computation of influenza intensity thresholds.
-We found that log-transforming intensity measures prior to the computation of thresholds empirically led to better behaved thresholds with closer to nominal exceedance rates.
-Concerning the question of how many observations per historical season should be included in the computation of thresholds, we provided analytical arguments and empirical evidence that the common choice $n = 30/m$ leads to too low thresholds when few historical seasons are available.
-We therefore recommend adopting $n = 1$ irrespective of the number $m$ of available historical seasons.
-This is possible using the available implementation in the \texttt{R} package \texttt{mem} and only requires setting \texttt{i.n.max = 1} in the call of the \texttt{memmodel} to override the default value 30/m.
-
-We focused on the computation of influenza intensity thresholds, but the same general arguments apply to the computation of thresholds to determine the onset of a season \citep{Vega2012}.
-As variability among observations around the season onset may be lower than around the peak the problem may be less severe; we did not assess this empirically as not all our data sets fully covered the off-season.
-
-More sophisticated methods to determine influenza intensity thresholds could be imagined, e.g.\ based on the broad literature on outbreak and anomaly detection \citep{Unkel2012}.
-However, we think that in practice a simple and interpretable tool with a robust software implementation like the MEM is a good choice.
-The emergence of a standard approach at the European level and beyond will contribute to improved comparability of results and facilitate the evaluation and refinement of the applied tool.
-With this work we hope to contribute a statistical perspective on this topic that can complement public health practitioners' experience from applied analyses.
-
-\section*{Data and code}
-
-Materials to reproduce the presented results are available at \url{https://github.com/jbracher/mem}.
+\newpage
 
@@
 \bibliography{bibliography_mem}
+
 \end{document}

### 2021-02-11 Simulations/plot_timeseries.R
Updating time series plot.

@@ -12,11 +12,11 @@ plot(dat_us[, 5])
 matr <- dat_ch
 xlim <- c(1990, 2020)
 
-plot_mem_ts <- function(matr, xlim, ylab = "inc / 100,000", ...){
+plot_mem_ts <- function(matr, xlim, ylab = "inc / 100,000", fact_ylim = 1.2, ...){
   years <- as.numeric(substr(colnames(matr), start = 13, stop = 16)) - 1
   labels_years <- gsub(".", "/", substr(colnames(matr), start = 8, stop = 16), fixed = TRUE)
   
-  plot(NULL, xlim = xlim, ylim = c(0, 1.2*max(matr)), xlab = "", ylab = ylab, axes = FALSE,...)
+  plot(NULL, xlim = xlim, ylim = c(0, fact_ylim*max(matr)), xlab = "", ylab = ylab, axes = FALSE,...)
   axis(1, at = years + 0.5, labels = labels_years)
   axis(2)
   box()
@@ -38,18 +38,23 @@ bplot_ranks <- function(matr, ylab = ""){
 }
 
 pdf("../Draft/figure/plot_data.pdf", width = 9, height = 5.5)
+
 par(las = 1, mar = c(3, 4, 0.5, 1))
 layout(matrix(1:8, ncol = 2, byrow = TRUE), widths = c(8, 2))
-plot_mem_ts(dat_fr, xlim = c(1985.5, 2019.5)); legend("topleft", legend = " France", bty = "n", cex = 1.2)
+plot_mem_ts(dat_fr, xlim = c(1985.5, 2019.5), fact_ylim = 1.1)
+legend("top", legend = " France  ", bty = "o", cex = 1.2, bg = "white")
 bplot_ranks(dat_fr)
 
-plot_mem_ts(dat_es, xlim = c(1998.5, 2019.5)); legend("topleft", legend = " Spain", bty = "n", cex = 1.2)
+plot_mem_ts(dat_es, xlim = c(1998.5, 2019.5))
+legend("top", legend = " Spain  ", bty = "o", cex = 1.2, bg = "white")
 bplot_ranks(dat_es)
 
-plot_mem_ts(dat_ch, xlim = c(2000.5, 2015.5)); legend("topleft", legend = " Switzerland", bty = "n", cex = 1.2)
+plot_mem_ts(dat_ch, xlim = c(2000.5, 2015.5), fact_ylim = 1.55);
[…]

### 2021-02-11 Draft/Draft_MEM_short.tex
Updating manuscript.

@@
 \title{An empirical assessment of influenza intensity thresholds obtained from the moving epidemic and WHO methods}
-\author{Johannes Bracher$^{1, 2}$, Jonas Littek$^1$}
-\date{ \small
-$^1$Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics\\
-$^2$Heidelberg Institute for Theoretical Studies, Computational Statistics Group}
+% \author{Johannes Bracher$^{1, 2}$, Jonas Littek$^1$}
+% \date{ \small
+% $^1$Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics\\
+% $^2$Heidelberg Institute for Theoretical Studies (HITS), Computational Statistics Group}
 
@@
 The WHO method uses only the highest value from each season.
-We perform simulation studies on how these choices impact thresholds and exceedance proportions.
-These are based on re-sampling of ILI data from France, Spain, Switzerland and the US.
+To assess the impact of these choices on thresholds and exceedance proportions we perform simulation studies which are based on re-sampling of ILI data from France, Spain, Switzerland and the US.
 When no transformation is applied, a large proportion of season peaks are classified as high or very high intensity.
@@
 
-\medskip
+% \medskip
 
-\noindent \textbf{Keywords:} influenza-like illness, intensity thresholds, moving epidmeic method, seasonal influenza, WHO method
+% \noindent \textbf{Keywords:} influenza-like illness, intensity thresholds, moving epidemic method, seasonal influenza, WHO method
 
@@
 
-Following the 2009 influenza H1N1 pandemic, the need for a rapid assessment tool for influenza intensity was recognized as a part of improved pandemic preparedness.
-The \textit{Review Committee on the Functioning of the
+Following the 2009 influenza H1N1 pandemic, the need for a rapid assessment tool for influenza intensity was recognized. The \textit{Review Committee on the Functioning of the
 International Health Regulations and on Pandemic Influenza (H1N1)} recommended that member states perform yearly updates and evaluations of intensity thresholds \citep[p.118]{WHO2011}.
 In the subsequent \textit{WHO Pandemic Infuenza Severity Assessment (PISA)} guideline \citep{WHO2017} the so-called WHO method \citep{WHO2014} and the moving epidemic method (MEM; \citealt{Vega2013, Vega2015}) were recommended to this end.
-The latter, which has also been recommended by the European Centre for Disease Prevention and Control (e.g., \citealt{ECDC2017}), has been adopted by numerous national public health agencies; see Supplement \textbf{XXX} for an overview of recent applications.
+The latter, which is also employed by the European Centre for Disease Prevention and Control (e.g., \citealt{ECDC2017}), has been adopted by numerous national public health agencies (e.g., \citealt{Dahlgren2019}, \citealt{Rakocevic2019}, \citealt{RedondoBravo2020}, \citealt{Vos2019}); see Supplement C for an overview of recent applications.
 The statistical properties of the MEM and WHO methods, however, have not yet been studied in detail.
 We here address this aspect via simulation experiments based on re-sampling of historical influenza data.
-Our results indicate that it may be beneficial to use a hybrid of the default settings of these different methods.% These are complement with some analytical statistical results.
+Our results indicate that it may be beneficial to use a hybrid of the default settings of these two methods.% These are complement with some analytical statistical results.
 
@@
 \item Select the $n$ largest observations from each of the $m$ past seasons to construct a set of reference observations.
-\item Apply an (invertible) transformation $f$ to all selected observations.
-\item Assume that the $m \times n$ transformed reference observarions come from a normal distribution and compute estimates $\mean, \sd$ of its mean and standard deviation.
+\item Apply an (invertible) transformation to all selected observations.
+\item Assume that the $m \times n$ transformed reference observations come from a normal distribution and compute estimates $\mean, \sd$ of its mean and standard deviation.
 \item Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\mean, \sd^2)$. A common choice for both methods is
@@
 \end{itemize}
-\item Transform thresholds back to the original scale by applying the inverse transformation $f^{-1}$ (e.g., the exponential function if $f$ is the natural logarithm).
+\item Transform thresholds back to the original scale (e.g. using the exponential function if a log trnsformation was used).
 \end{enumerate}
@@
 If peak incidences vary strongly across seasons, a log transformation is recommended \citep{WHO2017}.
-\item In the MEM method, the default for $f$ is the natural logarithm, while the number of included observations per season is set to $n = 30/m$, rounded to the nearest integer.
+\item In the MEM method, the default transformation is the natural logarithm, while the number of included observations per season is set to $n = 30/m$, rounded to the nearest integer.
 The total number of historical observations is thus kept approximately fixed.
@@
 However, thresholds for all levels will then converge to the arithmetic or geometric mean if sufficient historical data are available.
-As illustrated in Supplement \textbf{XXX}, this is not a desirable behaviour.
+As illustrated in Supplement B, this is not a desirable behaviour.
 
@@
 \end{itemize}
-The latter aspect is intuitive, but also supported by formal statistical considerations detailed in Supplement \textbf{XXX}.
-These imply that thresholds are unbiased for $n = 1$ in the sense that in the long run they will be exceeded by around 40\%, 90\% and 97.5\% of the season peaks, depending on the level.
+The latter aspect is intuitive, but also supported by formal statistical considerations detailed in Supplement A.
+These imply that thresholds are unbiased for $n = 1$ in the sense that, as intended, they will be exceeded by around 40\%/90\%/97.5\% of the season peaks in the long run.
 When $n > 1$, thresholds will tend to be lower and exceeded more frequently.
@@
 \end{itemize}
-The range $m =5, \dots, 15$ is motivated by the range of values found in real-world applications, see our overview in Supplement \textbf{XXX}.
+The range $m =5, \dots, 15$ is motivated by the range of values found in real-world applications, see our overview in Supplement C.
 All analyses were performed using the R language for statistical computing \citep{RCT2020} and the package \texttt{mem} \citep{Lozano2020}.
@@
 Results from our simulation study are shown in Figures \ref{fig:results1} (France, Spain) and \ref{fig:results2} (Switzerland and US).
-In the first and third colum of each figure we also show an analytical approximation of the expected thresholds, see Supplement \textbf{XXX} for details.
-In all four countries, using a log transformation leads to increased thresholds, in particular for the very high level.
+In the first and third colum of each figure we also show an analytical approximation of the expected thresholds, see Supplement A for details.
+In all four countries, using a log transformation leads to increased thresholds, in particular for the high and very high levels.
 Especially in France and Spain, thresholds obtained without the log transformation are too low, as can be seen from the large proportion of season peaks classified as very high (around 10\% when using $n = 1$).
@@
 As a particularly striking example, the  average threshold for high intensity in France (when using a log transformation) increases from 834 for $m = 5, n = 6$ to 1011 for $m = 10, n = 3$ and 1080 for $m = 15, n = 2$.
-For $n = 1$, in which case thresholds can be interpreted as unbiased (Supplement \textbf{XXX}), the average is 1147.
+For $n = 1$, in which case thresholds can be interpreted as unbiased (Supplement A), the average is 1147.
 Including historical observations which are not from actual peak weeks thus leads to a considerable lowering of alarm thresholds and will increase the number of alerts for high and very high influenza activity.
@@
 fifth largest observation per season, see right column of Figure \ref{fig:data}.
-In the US and France, on average one in three season peaks is classified as high and one in six as very high intensity when applying the MEM with log transformation and $m = 5, n = 6$.
+In the US and France, on average one in three season peaks is classified as high and one in seven as very high intensity when applying the MEM with log transformation and $m = 5, n = 6$.
 This is rather far from the intended excedance probabilities of 10\% and 2.5\%.
@@
 When always using $n = 1$, the average thresholds and shares of the different categories are more well-behaved also for small $m$.
-This holds especially when applying a log transformation, even though for the French data a certain mismatch with the nominal exceedance probabilities remains.
-Also, there is considerable variability in the estimated thresholds.
+This holds especially when applying a log transformation, even though certain mismatches with the nominal exceedance probabilities remain.
+Also, there is considerable variability in the estimated thresholds (shaded areas in Figures \ref{fig:results1} and \ref{fig:results2}).
 These difficulties, however, are inherent in the problem of estimating a 90\% or 97.5\% quantile based on 5--10 observations.
@@
 We provided a statistical assessment of implementation choices in a widely used framework for the computation of influenza intensity thresholds.
-We found applying a log transformation led to better behaved thresholds and closer to nominal exceedance rates.
-Concerning the question of how many observations per historical season should be included in the computation of thresholds, we found that the common choice $n = 30/m$ results in too low thresholds when few historical seasons are available.
+We found that applying a log transformation led to better behaved thresholds and closer to nominal exceedance rates.
+Concerning the question of how many observations per historical season should be included to computate thresholds, we found that the common choice $n = 30/m$ results in too low thresholds when few historical seasons are available.
 We therefore recommend adopting $n = 1$ irrespective of the number of available historical seasons.
-This is possible in the available \texttt{R} package \texttt{mem} by setting \texttt{i.n.max = 1} when calling the function \texttt{memmodel}.
+This is possible in the available \texttt{R} package \texttt{mem} by setting \texttt{i.n.max = 1} when the function \texttt{memmodel} is called.
 
@@
 
-We think that a simple and interpretable tool with a robust software implementation like the MEM is a valuable tool in practice.
+We think that a simple and interpretable tool with a well-documented and open source software implementation like \texttt{mem} is a valuable tool in practice.
 The use of a standard approach at the European level will improve comparability of results, facilitating the evaluation and refinement of the tool.

### 2021-02-11 Draft/Supplement_MEM_short.tex
Updating manuscript.

@@
 \title{Supplementary Materials for ``An empirical assessment of influenza intensity thresholds obtained from the moving epidemic and WHO methods''}
-\author{Johannes Bracher, Jonas Littek}
-\date{ \small
-Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics}
+% \date{ \small
+% $^1$Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics\\
+% $^2$Heidelberg Institute for Theoretical Studies, Computational Statistics Group}
 
 \maketitle
+
+\appendix
 
@@
 
-We start by re-stating the definition of the MEM and WHO method form Section \textbf{XXX} of the main manuscript in a slighty more formal way and introducing relevant notation.
-We assume again that thresholds are based on influenza intensity data from $m$ past seasons and applied to the ($m + 1$)-th season.
-Thresholds for a given intensity measure $X$ are then obtained via the following steps.
+We start by re-stating the definitions of the MEM and WHO method from Section 2 of the main manuscript in a slighty more formal way and introducing relevant notation.
+We assume again that thresholds are based on data from $m$ past seasons and applied to the ($m + 1$)-th season.
+Thresholds for an measure $X$ are then obtained as follows.
 
@@
 \begin{itemize}
-\item the 40th percentile $q_{Y, 0.4} = \mean - 0.25 \sd$ as the threshold between low and medium intensity;
-\item the 90th percentile $q_{Y, 0.9} = \mean + 1.28 \sd$ as the threshold between medium and high intensity;
-\item the 97.5th percentile $q_{Y, 0.975} = \mean + 1.96\sd$ as the threshold between high and very high intensity.
+\item the 40th percentile $q_{Y, 0.4} = \bar{y} - 0.25 \sd$ as the threshold between low and medium intensity;
+\item the 90th percentile $q_{Y, 0.9} = \bar{y} + 1.28 \sd$ as the threshold between medium and high intensity;
+\item the 97.5th percentile $q_{Y, 0.975} = \bar{y} + 1.96\sd$ as the threshold between high and very high intensity.
 \end{itemize}
@@
 \item In the WHO method, $n = 1$ observation per season is used irrespective of the number $m$ of historical seasons.
-The standard procedure is to apply no transformation (i.e.\ $f$ is set to the identity function).
+The standard procedure is to apply no transformation, i.e.\ set $f$ to the identity function.
 \item In the MEM method, the default choice for $f$ is the natural logarithm, while the number of observations per season is set to $n = 30/m$.
-(The exact specification used in the \texttt{R} package \texttt{mem} is $n = \max(1, \lfloor 30/m \rceil)$, where $\lfloor \cdot \rceil$ denotes rounding to the nearest integer.) The total number of historical observations is thus kept approximately fixed, ensuring a reasonably large number of reference observations.
+(The exact specification used in the R package \texttt{mem} is $n = \max(1, \lfloor 30/m \rceil)$, where $\lfloor \cdot \rceil$ denotes rounding to the nearest integer.) The total number of historical observations is thus kept approximately fixed.
 \end{itemize}
@@
 
-We start by considering how the expectations of the empirical mean $\bar{Y}$ and standard deviation $S$ entering into equation \eqref{eq:def_q} depend on $n$ (note that we now capitalize these as we consider the random variables rather than concrete observed values).
-Denote by $\mathbf{Y}^{(j)}$ the vector of the $n$ largest transformed incidences from season $j$ in decreasing order, i.e.
+We assume that the different historical seasons are independent realizations of the same random process, which is certainly not strictly true, but should hold in good approximation.
+Denote by $\mathbf{Y}^{(j)}$ the random vector of the $n$ largest transformed incidences from season $j$ in decreasing order, i.e.
 $\mathbf{Y}^{(j)} = (Y^{(j)}_1, \dots, Y^{(j)}_n)^\top$ with $Y^{(j)}_1 \geq Y^{(j)}_2 \geq \dots \geq Y^{(j)}_n$.
-Now assume that $\mathbf{Y}^{(1)}, \dots, \mathbf{Y}^{(m)}$ are independent and identically distributed with mean and covariance matrix
+The mean and covariance matrix of $\mathbf{Y}^{(j)}$ are denoted by
 \begin{align}
@@
 respectively, where to make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$.
-It can then be shown that the expectations of $\mean$ and $\sd^2$ when using $m$ seasons and $n$ observations from each season are given by
+It can then be shown that the expectations of $\mean$ and $S^2$ when using $m$ seasons and $n$ observations from each season are given by
 \begin{align}
@@
 \label{eq:expectation_mu}\\
-\mathbb{E}(\sd^2) & = \frac{m}{mn - 1} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{n(mn - 1)} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ - \ \frac{m}{n(mn - 1)}\left(\sum_{i = 1}^n \mu_i\right)^2,
+\mathbb{E}(S^2) & = \frac{m}{mn - 1} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{n(mn - 1)} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ - \ \frac{m}{n(mn - 1)}\left(\sum_{i = 1}^n \mu_i\right)^2,
 \label{eq:expectation_sigma2}
@@
 \begin{equation}
-\mathbb{E}(q_\alpha) \approx f^{-1}\left\{\mathbb{E}(\mean) + z_\alpha \sqrt{\mathbb{E}(\sd^2)}\right\}
+\mathbb{E}(q_\alpha) \approx f^{-1}\left\{\mathbb{E}(\mean) + z_\alpha \sqrt{\mathbb{E}(S^2)}\right\}
 \label{eq:expectation_q}
 \end{equation}
-usually holds in good approximation in our applied setting (with $f^{-1}$ the idenity or exponential function). This implies that for $n = 1$, we have
+usually holds in good approximation in our applied setting. It can be shown that for $n = 1$, we have
 $$
-\mathbb{E}(q_\alpha) \approx f^{-1}(\mu_1 + z_\alpha \sigma_1).
+\mathbb{E}(q_\alpha) \approx f^{-1}(\mu_1 + z_\alpha \sigma_1),
 $$
+usually in good approximation.
 This is a desirable property.
-Loosely speaking, if in addition the $Y_{1}^{(j)}$ come from a normal distribution, the nominal exceedance probabilities for the different thresholds would be achieved on average.
+Loosely speaking, if in addition the $Y_{1}^{(j)}$ come from a normal distribution, the nominal exceedance probabilities (40\%/90\%/97.5\%) for the different thresholds would in the long run be achieved.
 If a different value $n > 1$ is chosen, this will generally no longer be the case.
@@
 As a consequence, when choosing $n > 1$, one must expect to classify a larger number of season peaks as high or very high intensity.
-When choosing $n = 30/m$ as suggested for the MEM, threholds will then tend to increase the more years of data are used to compute thresholds.
+When choosing $n = 30/m$ as suggested for the MEM, thresholds will then tend to increase the more years of data are used to compute thresholds.
 
@@
 
-We assume that the different historical seasons are independent realizations of the same random process, which is certainly not strictly true, but should hold in good approximation.
-Then $\mathbf{Y}^{(1)}, \dots, \mathbf{Y}^{(m)}$ are independent and identically distributed with mean and covariance matrix
-\begin{align}
-\mathbb{E}(\mathbf{Y}^{(j)}) = \boldsymbol{\mu} = \left(\begin{array}{c}
-\mu_1\\
-\vdots\\
-\mu_n
-\end{array}\right) \ \ \ \text{and} \ \ \ \text{Cov}(\mathbf{Y}^{(j)}) = \boldsymbol{\Sigma} =
-\left(\begin{array}{ccc}
-\sigma_{1, 1} & \cdots & \sigma_{1, n}\\
-\vdots & \ddots &\vdots\\
-\sigma_{n, 1} & \cdots & \sigma_{n, n}
-\end{array}\right),
-\end{align}
-respectively, where to make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$. We are interested in the expectations of the random variables $\mean$ and $\sd$, where
+We start by considering the addressing the expectations of the random variables $\mean$ and $S$, where
 \begin{align*}
 \mean & = \frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n Y_i^{(j)}\\
-\sd & = \sqrt{\frac{1}{mn - 1} \sum_{j = 1}^m \sum_{i = 1}^n \left(Y_i^{(j)} - \mean\right)^2}.
+S & = \sqrt{\frac{1}{mn - 1} \sum_{j = 1}^m \sum_{i = 1}^n \left(Y_i^{(j)} - \mean\right)^2}.
 \end{align*}
@@
 \mathbb{E}(\mean) = \frac{1}{n} \sum_{i = 1}^n \mu_i.
-\label{eq:expectation_mu}
 \end{equation}
-Instead of the empirical standard deviation $\sd$ we first consider the variance $\sd^2$, which can also be written as
+Instead of the empirical standard deviation $S$ we first consider the variance $S^2$, which can also be written as
 \begin{align}
-\sd^2 & = \frac{mn}{mn - 1} \left\{\frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n \left(Y_i^{(j)} - \mean\right)^2 \right\}\\
+S^2 & = \frac{mn}{mn - 1} \left\{\frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n \left(Y_i^{(j)} - \mean\right)^2 \right\}\\
 & = \frac{mn}{mn - 1} \left\{ \underbrace{\frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n Y_i^{(j)2}}_{= a} \ \ \ - \ \ \ \underbrace{\left(\frac{1}{mn} \sum_{j = 1}^m \sum_{i = 1}^n Y_i^{(j)} \right)^2}_{= b} \right\} \label{eq:sigma2hat}
@@
 \end{align*}
-Plugging this back into equation \eqref{eq:sigma2hat} we obtain
+Plugging these results back into equation \eqref{eq:sigma2hat} we obtain
 \begin{equation}
-\mathbb{E}(\sd^2) = \frac{m}{mn - 1} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2) \ \ \ - \ \ \ \frac{1}{n(mn - 1)} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ \ \ - \ \ \ \frac{m}{n(mn - 1)}\left(\sum_{i = 1}^K \mu_i\right)^2.
-\label{eq:expectation_sigma2}
+\mathbb{E}(S^2) = \frac{m}{mn - 1} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2) \ \ \ - \ \ \ \frac{1}{n(mn - 1)} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ \ \ - \ \ \ \frac{m}{n(mn - 1)}\left(\sum_{i = 1}^n \mu_i\right)^2.
+\label{eq:expectation_sigma2_2}
 \end{equation}
-It is straightforward to see that for $n = 1$ the expressions \eqref{eq:expectation_mu} and \eqref{eq:expectation_sigma2} simplify to
+It is straightforward to see that for $n = 1$ the expressions \eqref{eq:expectation_mu} and \eqref{eq:expectation_sigma2_2} simplify to
 $$
-\mathbb{E}(\mean^2) = \mu_1 \ \ \text{ and } \ \ \mathbb{E}(\sd^2) = \sigma^2_1.
+\mathbb{E}(\mean^2) = \mu_1 \ \ \text{ and } \ \ \mathbb{E}(S^2) = \sigma^2_1,
 $$
+as given in equation \eqref{eq:expectation_sigma2}.
 
-There is no general way of computing the expectation $\mathbb{E}(\sd)$, but unless the true distribution of  $\sd$ has strong excess curtosis,
+There is no general way of computing the expectation $\mathbb{E}(S)$, but unless the true distribution of  $S^2$ has strong excess curtosis,
 \begin{equation}
-\mathbb{E}(\sd) \approx \sqrt{\mathbb{E}(\sd^2)}
+\mathbb{E}(S) \approx \sqrt{\mathbb{E}(S^2)}
 \label{eq:expectation_sigma}
@@
 is a reasonable approximation.
-We can then plug equations \eqref{eq:expectation_mu} and \eqref{eq:expectation_sigma} into the formulae for the thresholds $q_{Y, \alpha}$ (with $\alpha \in \{0.4, 0.9, 0.975\}$) on the transformed scale and obtain
+We can then plug equations \eqref{eq:expectation_mu} and \eqref{eq:expectation_sigma} into the formulae for the thresholds $q_{Y, \alpha}$ on the transformed scale and obtain
 $$
-\mathbb{E}(q_{Y, \alpha}) \approx \mathbb{E}(\mean) + z_\alpha \sqrt{\mathbb{E}(\sd^2)},
+\mathbb{E}(q_{Y, \alpha}) \approx \mathbb{E}(\mean) + z_\alpha \sqrt{\mathbb{E}(S^2)},
 $$
-where $z_\alpha$ is the $\alpha$ quantile of the standard normal distribution.
+where $z_\alpha$ is the $\alpha$ quantile of the standard normal distribution (with $\alpha \in \{0.4, 0.9, 0.975\}$).
 
@@
 \end{equation}
+As can be seen from Figures 2 and 3 from the main manuscript, this approximation works very well in praxis.
+To compute the values indicated by the small black dots, we just plugged the empirical means and standard deviations of the $\mathbf{y}^{(j)}$ in the respective data sets into equations \eqref{eq:expectation_mu}--\eqref{eq:expectation_q}.
 
@@
 These thresholds show somewhat peculiar behaviour.
-We illustrate this by repeating the simulation study from Section \textbf{XXX} of the main manuscript for France with the respective settings.
+We illustrate this by repeating the simulation study from Section 3 of the main manuscript for France with the respective settings.
 The result is shown in Figure \ref{fig:france_ci}.
@@
 \includegraphics[width=0.9\textwidth]{figure/plot_results_ci.pdf}
-\caption{Simulation results for France as in Figure \textbf{XX} of the main manuscript, but using confidence intervals based on standard errors rather than prediction intervals based on standard deviations.}
+\caption{Simulation results for France as in Figure 2 of the main manuscript, but using confidence intervals based on standard errors rather than prediction intervals based on standard deviations.}
 \label{fig:france_ci}
@@
 The results are summarized in Table \ref{tab:literature}.
-As can be seen from the numerous entries from the years 2019 and 2020, the MEM has quickly become a standard approach in the determination of intensity thresholds for influenza and other respiratory diseases.
-Indeed, the contributions come from numerous countries and in many cases have been co-authored by representatives of national or regional public health agencies.
-In most analyses, the suggested threshold levels at the 40th, 90th and 97.5th percentile as described in Section \ref{sec:definitions} are used.
-Variability with respect to the number $m$ of historical seasons included is considerable, with a range from three to 16 seasons.
-Consequently, the number $n$ of observations included per season ranges from two to ten (none of the above papers indicated a modification of the default setting $n = 30/m$ of the moving epidemic method).
-We only found three published applications of the WHO method, one of them providing a comparison to the thresholds from the MEM method \citep{Rguig2020}.
 
@@
 
-\newpage
+As can be seen from the numerous entries from the years 2019 and 2020, the MEM has quickly become a standard approach in the determination of intensity thresholds for influenza and other respiratory diseases.
+Indeed, the contributions come from numerous countries and in many cases have been co-authored by representatives of national or regional public health agencies.
+In most analyses, the suggested threshold levels at the 40th, 90th and 97.5th percentile as described in Section \ref{sec:definitions} are used.
+Variability with respect to the number $m$ of historical seasons included is considerable, with a range from three to 16 seasons.
+Consequently, the number $n$ of observations included per season ranges from two to ten (none of the above papers indicated a modification of the default setting $n = 30/m$ of the moving epidemic method).
+We only found three published applications of the WHO method, one of them providing a comparison to the thresholds from the MEM method \citep{Rguig2020}.
 

### 2021-02-16 Draft/Draft_MEM_short.tex
Updating Draft.

@@
 To assess the impact of these choices on thresholds and exceedance proportions we perform simulation studies which are based on re-sampling of ILI data from France, Spain, Switzerland and the US.
-When no transformation is applied, a large proportion of season peaks are classified as high or very high intensity.
+When no transformation is applied, a rather large proportion of season peaks are classified as high or very high intensity.
 This can be mitigated by a logarithmic transformation.
@@
 International Health Regulations and on Pandemic Influenza (H1N1)} recommended that member states perform yearly updates and evaluations of intensity thresholds \citep[p.118]{WHO2011}.
-In the subsequent \textit{WHO Pandemic Infuenza Severity Assessment (PISA)} guideline \citep{WHO2017} the so-called WHO method \citep{WHO2014} and the moving epidemic method (MEM; \citealt{Vega2013, Vega2015}) were recommended to this end.
+In the subsequent \textit{WHO Pandemic Influenza Severity Assessment} (PISA) guideline \citep{WHO2017} the so-called WHO method \citep{WHO2014} and the moving epidemic method (MEM; \citealt{Vega2013, Vega2015}) were recommended to this end.
 The latter, which is also employed by the European Centre for Disease Prevention and Control (e.g., \citealt{ECDC2017}), has been adopted by numerous national public health agencies (e.g., \citealt{Dahlgren2019}, \citealt{Rakocevic2019}, \citealt{RedondoBravo2020}, \citealt{Vos2019}); see Supplement C for an overview of recent applications.
@@
 \end{itemize}
-\item Transform thresholds back to the original scale (e.g. using the exponential function if a log trnsformation was used).
+\item Transform thresholds back to the original scale (e.g. using the exponential function if a log transformation was used).
 \end{enumerate}
@@
 \end{itemize}
-The rationale behind the percentile-based approach is that ``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold'' \citep{WHO2017}.
+The rationale behind the percentile-based approach is that ``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold'' \citep[p.10]{WHO2017}.
 
@@
 In previous works (\citealt{WHO2014, Vega2015}), the above thresholds have been described as upper ends of one-sided confidence intervals for the the arithmetic (WHO method) or geometric mean (MEM) of the reference observations.
-This, however, is imprecise terminology as in the computations the standard deviation $\sd$ rather than the standard error $\sd/\sqrt{nm}$ is used (see documentation of the \texttt{mem} package and \citealt{WHO2014}, p.
-69).
+This, however, is imprecise terminology as in the computations the standard deviation $\sd$ rather than the standard error $\sd/\sqrt{nm}$ is used (see documentation of the \texttt{mem} package and \citealt{WHO2014}, p.69).
 We note that the use of actual confidence intervals is also possible in the \texttt{mem} package (\texttt{i.type.intensity = 1} for the geometric, \texttt{i.type.intensity = 2} for the arithmetic mean).
@@
 
-\section{Simulaton study}
+\section{Simulation study}
 \label{sec:simulation}
@@
 \item The application of a logarithmic transformation is expected to lead to higher thresholds for the extreme categories and thus a lower proportion of seasons peaks classified as high or very high.
-\item In the MEM, the reference set also contains past obervations which are not actually peak observations.
+\item In the MEM, the reference set also contains past observations which are not actually peak observations.
 This may introduce a downward bias in thresholds, especially if due to a small number $m$ of available seasons $n$ is large.
@@
 In order to closely mimic the seasonal patterns of influenza, we re-sample historical surveillance data rather than generating fully synthetic data.
-Assume we have $M$ years of historical data on a measure of influenza activity.
+Assume the historical data contain $M$ seasons of historical data on a measure of influenza activity.
 We then repeat the following steps 500 times:
@@
 \end{itemize}
-The range $m =5, \dots, 15$ is motivated by the range of values found in real-world applications, see our overview in Supplement C.
+The range $m =5, \dots, 15$ is motivated by the range of values found in real-world applications, see overview in Supplement C.
 All analyses were performed using the R language for statistical computing \citep{RCT2020} and the package \texttt{mem} \citep{Lozano2020}.
@@
 Results from our simulation study are shown in Figures \ref{fig:results1} (France, Spain) and \ref{fig:results2} (Switzerland and US).
-In the first and third colum of each figure we also show an analytical approximation of the expected thresholds, see Supplement A for details.
+In the first and third column of each figure we also show an analytical approximation of the expected thresholds, which agrees very well with the simulation results; see Supplement A for details.
 In all four countries, using a log transformation leads to increased thresholds, in particular for the high and very high levels.
@@
 In the US and France, on average one in three season peaks is classified as high and one in seven as very high intensity when applying the MEM with log transformation and $m = 5, n = 6$.
-This is rather far from the intended excedance probabilities of 10\% and 2.5\%.
+This is rather far from the intended exceedance probabilities of 10\% and 2.5\%.
 
@@
 \includegraphics[page=1, width=0.95\textwidth]{figure/plot_results.pdf}
-\caption{Results of re-sampling simulation study, France and Spain.
+\caption{Results of simulation study, France and Spain.
 First and third column: mean intensity thresholds along with bands delimited by the empirical 5\% and 95\% quantiles.
-Analytical approximations of the mean threshold values are marked as small black dots.
+Analytical approximations of the mean threshold values are marked as small black crosses.
 Second and fourth columns: resulting average shares of seasons classified as low, medium, high and very high intensity.}
@@
 \includegraphics[page=2, width=0.95\textwidth]{figure/plot_results.pdf}
-\caption{Results of re-sampling simulation study, Switzerland and United States.
+\caption{Results of simulation study, Switzerland and United States.
 First and third column: mean intensity thresholds along with bands delimited by the empirical 5\% and 95\% quantiles.
-Analytical approximations of the mean threshold values are marked as small black dots.
+Analytical approximations of the mean threshold values are marked as small black crosses.
 Second and fourth columns: resulting average shares of seasons classified as low, medium, high and very high intensity.}
@@
 We found that applying a log transformation led to better behaved thresholds and closer to nominal exceedance rates.
-Concerning the question of how many observations per historical season should be included to computate thresholds, we found that the common choice $n = 30/m$ results in too low thresholds when few historical seasons are available.
+Concerning the question of how many observations per historical season should be included to compute thresholds, we found that the common choice $n = 30/m$ results in too low thresholds when few historical seasons are available.
 We therefore recommend adopting $n = 1$ irrespective of the number of available historical seasons.
-This is possible in the available \texttt{R} package \texttt{mem} by setting \texttt{i.n.max = 1} when the function \texttt{memmodel} is called.
+This is possible in the \texttt{R} package \texttt{mem} by setting \texttt{i.n.max = 1} when the function \texttt{memmodel} is called.
 
@@
 The use of a standard approach at the European level will improve comparability of results, facilitating the evaluation and refinement of the tool.
-With this work we hope to contribute a statistical perspective on this topic which complements public health practitioners' experience from applied analyses.
+With this work we hope to contribute a statistical perspective on this topic, complementing public health practitioners' experience from applied analyses.
 
@@
 
-Materials to reproduce the presented results are available at \url{https://github.com/jbracher/mem}.
+Materials to reproduce the presented results are available at \textit{(URL removed for double-blind peer review)}. % \url{https://github.com/jbracher/mem}.
 
-\section*{Author contributions}
+% \section*{Author contributions}
 
-Study concept: JB; Data processing: JB and JL; Literature review: JB and JL; Statistical analysis / implementation: JB; Writing and editing: JB and JL.
+% Study concept: JB; Data processing: JB and JL; Literature review: JB and JL; Statistical analysis / implementation: JB; Writing and editing: JB and JL.
 
-\section*{Acknowledgements}
+% \section*{Acknowledgements}
 
-We would like to thank Laura Werlen for helpful feedback. Johannes Bracher was supported by the Helmholtz Foundation via the SIMCARD Information and Data Science Pilot Project.
+% We would like to thank Laura Werlen for helpful feedback. Johannes Bracher was supported by the Helmholtz Foundation via the SIMCARD Information and Data Science Pilot Project.
 

### 2021-02-16 Simulations/functions_simulations_mem.R
Updating codes.

@@ -122,6 +122,7 @@ plot_exceedance_summary <- function(summary_exceedance, ...){
   box()
 }
 
+# custon legend for plots
 legend_summary <- function(){
   plot(NULL, xlim = 0:1, ylim = 0:1, axes = FALSE, xlab = "", ylab = "")
   legend("top", legend = c("Intensity levels:", "very high", "high", "medium", "low", 
@@ -131,10 +132,21 @@ legend_summary <- function(){
                  NA, "black", "grey", "black", NA,
                  NA, "black", "red"),
          pch = c(NA, 15, 15, 15, 15, 
-                 NA, 15, 15, 1, NA,
+                 NA, 15, 15, 4, NA,
                  NA, NA, 15, NA), 
          lty = c(NA, NA, NA, NA, NA,
                  NA, 2, NA, NA, NA,
                  NA, 3, NA, NA),
+         pt.cex = c(1, 1, 1, 1, 1,
+                 1, 1, 1, 0.6, 1,
+                 1, 1, 2, 1),
          ncol = 3, bty = "n")
+}
+
+
+# adding analytical approximations
+lines_approx_expectations <- function(approx_expectations, lty = 0, pch = 4, cex = 0.6){
+  lines(approx_expectations[, "n"], approx_expectations[, "medium_0.4"], type = "o", lty = lty, pch = pch, cex = cex)
+  lines(approx_expectations[, "n"], approx_expectations[, "high_0.9"], type = "o", lty = lty, pch = pch, cex = cex)
+  lines(approx_expectations[, "n"], approx_expectations[, "very_high_0.975"], type = "o", lty = lty, pch = pch, cex = cex)
 }

### 2021-02-16 Simulations/summarize_simulations.R

@@ -120,11 +120,6 @@ save(summary_thresholds_us, summary_thresholds1_us, summary_thresholds_nolog_us,
 
 load("Results/approx_expectations.rda")
 
-lines_approx_expectations <- function(approx_expectations, lty = 0, pch = 1, cex = 0.6){
-  lines(approx_expectations[, "n"], approx_expectations[, "medium_0.4"], type = "o", lty = lty, pch = pch, cex = cex)
-  lines(approx_expectations[, "n"], approx_expectations[, "high_0.9"], type = "o", lty = lty, pch = pch, cex = cex)
-  lines(approx_expectations[, "n"], approx_expectations[, "very_high_0.975"], type = "o", lty = lty, pch = pch, cex = cex)
-}
 
 pdf("../Draft/figure/plot_results.pdf", width = 9, height = 11)
 par(mfrow = c(4, 4), mar = c(4, 4, 3, 1), las = 1)
@@ -144,7 +139,7 @@ mtext("log-transformed, n = 30 / m", 3, at = 0, line = lin, cex = ce)
 
 plot_sim_summary(summary_thresholds_nolog_fr, xlab = "# included seasons m", ylab = "threshold", ylim = c(0, 2500))
 lines_approx_expectations(approx_expectations_nolog_fr)
-mtext("France", 3, at = -1, line = 1.2, cex = 1)
+mtext("France", 3, at = -1, line = 1.2, cex = 1, font = 2)
 plot_exceedance_summary(summary_exceedance_nolog_fr, xlab = "# included seasons m", ylab = "average share")
 mtext("natural scale, n = 30 / m", 3, at = 0, line = lin, cex = ce)
 
@@ -170,7 +165,7 @@ mtext("log-transformed, n = 30 / m", 3, at = 0, line = lin, cex = ce)
 
 plot_sim_summary(summary_thresholds_nolog_es, xlab = "# included seasons m", ylab = "threshold", ylim = c(0, 1200))
 lines_approx_expectations(approx_expectations_nolog_es)
-mtext("Spain", 3, at = -1, line = 1.2, cex = 1)
+mtext("Spain", 3, at = -1, line = 1.2, cex = 1, font = 2)
 plot_exceedance_summary(summary_exceedance_nolog_es, xlab = "# included seasons m", ylab = "average share")
 mtext("natural scale, n = 30 / m", 3, at = 0, line = lin, cex = ce)
 
[…]
